### 2016-02-22 Paper/sigplanconf-template.tex
Add paper draft

%-----------------------------------------------------------------------------
%
%               Template for sigplanconf LaTeX Class
%
% Name:         sigplanconf-template.tex
%
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
%               sigplanconf-guide.pdf
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%
%-----------------------------------------------------------------------------

\documentclass[preprint]{sigplanconf}

% The following \documentclass options may be useful:

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% numbers       To obtain numeric citation style instead of author/year.

\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{amsthm}

\newcommand{\cL}{{\cal L}}

\begin{document}

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country}
\copyrightyear{20yy}
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
\copyrightdoi{nnnnnnn.nnnnnnn}

% Uncomment the publication rights you want to use.
%\publicationrights{transferred}
%\publicationrights{licensed}     % this is the default
%\publicationrights{author-pays}

% \titlebanner{banner above paper title}        % These are ignored unless
% \preprintfooter{short description of paper}   % 'preprint' option specified.

\title{A New Approach for Parallel Functional Arrays}

\authorinfo{Ananya Kumar}
           {Carnegie Mellon University}
           {ananyak@andrew.cmu.edu}
\authorinfo{Guy Blelloch}
           {Carnegie Mellon University}
           {blelloch@cs.cmu.edu}

\maketitle

\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}

\begin{abstract}
In this paper we introduce a $O(1)$ wait-free, parallel, functional array, that allows $O(1)$ reads and writes to the most recent version of the array. We describe the cost dynamics and sketch out a provable implementation. We show favorable benchmarks comparing our functional arrays with regular arrays in Java.
\end{abstract}

\category{CR-number}{subcategory}{third-level}

\keywords
array, parallel, cost semantics

\section{Introduction}

Arrays are very important in functional programming languages because they allow work-efficient implementations of algorithms like depth-first search. Accessing old versions of arrays can be useful for efficient checkpointing, logging, and event handling. In this paper we introduce an efficient, parallel, functional array that allows $O(1)$ reads and writes to the most recent version of the array.

\section{Previous Approaches}

Many functional programming languages use monads to build arrays. However, monads do not allow accessing old versions of an array.

Another approach is to use compiler based reference counting. If the number of references to an array is provably one, the compiler allows the array to mutate, otherwise the compiler copies the array before writing to the array. This approach makes it difficult for programmers to reason about the time complexity of array operations because it depends on the compiler. Furthermore, if multiple variables reference the same array, the array is copied even if all writes are to the most recent version. This could make the time complexity of programs significantly higher than they need to be.

A third approach is to use linear types to ensure that programmers can only access and mutate arrays in valid ways. This approach is very difficult to implement in existing programming languages.

O' Neill's functional arrays guarantee $O(1)$ access and insertions when used like an imperative array. However, the arrays use version trees which are inefficient in practice. Additionally, the reads and writes to the most recent version of the array might take $O(\log{n})$ work if the array is not used like an imperative array.

Chuang's approach supports $O(1)$ accesses and insertions to the most recent version of arrays, however accessing old versions of the array takes $O(n)$ work which is often impractical.

None of the existing approaches support concurrent operations on arrays. O' Neill suggests having a lock for each element of the array. However, when many threads contend for the same element, this would serialize accesses and make accesses not $O(1)$. Additionally, per-element locks add significant memory overhead. Separation logic could be used to parallelize divide and conquer algorithms on arrays however they would severely restrict the way functional arrays can be used.

\section{Dynamics}

We use a standard applicative-order language defined as follows:
$$L = x \; | \; c  \; | \; \lambda x.e \; | \; f \; x \; | \; e_1 || e_2 \; | \; \text{if } e_1 \text{ then } e_2 \text{ else } e_3$$

$L$ contains the usual arithmetic types, such as the natural numbers, and numerical operations such as sums and products. Dynamics for function application and fork-join are given as examples.
$$\frac{e_1 \Downarrow v_1 \;\;\; e_2 \Downarrow v_2}{e_1 || e_2 \Downarrow (v_1, v_2)} \text{ (fork-join)}$$
$$\frac{f \Downarrow \lambda x . e \;\;\; y \Downarrow v}{f \; y \Downarrow [v/x]e} \text{ (function-app)}$$

The language also includes 3 functions for working with arrays: tabulate, get, and set. The dynamics for tabulate is given below:
$$\frac{f(1) \Downarrow v_1, ..., f(n) \Downarrow v_n}{\text{tabulate} \; f \; n \Downarrow [1 \mapsto v_1, ..., n \mapsto v_n]}$$

The dynamics for the other two array functions are standard.

\section{Approach}

An array $A$ is NEW if the set function has not been called with $A$ as an argument, otherwise $A$ is OLD. While the dynamics of our arrays are functional, the costs are different for OLD and NEW arrays.

\begin{figure}[!ht]
\centering
\includegraphics[scale=0.3]{new_array_A}
\nocaptionrule \caption{New functional array filled with 0s}
\label{fig:new_array_A}
\end{figure}

Suppose that $A$ is a NEW functional array  (see figure \ref{fig:new_array_A}). $A$ has a version number $V$, and a pointer to an ArrayData object $AD$. $AD$ keeps a regular (mutable) array of values, which corresponds to the values in $A$. $AD$ has a version number, which is the same as $A'$s ($V$) to indicate that $A$ is NEW. For each element of the array in $AD$, keep a log of values that used to be at that index. The logs are initially empty.

Get on NEW arrays: to get the $i^{\text{th}}$ element in $A$, simply access array[i] in $AD$.

\begin{figure}[!ht]
\centering
\includegraphics[scale=0.3]{set_A_return_B}
\nocaptionrule \caption{$B = set(A, 3, 5)$ changes the array in $AD$ and adds a log entry.}
\label{fig:set_A_return_B}
\end{figure}

Set on NEW arrays: Suppose that $A[i]$ is $v_{old}$ and we want to do $B = set(A, i, v_{new})$ (see figure \ref{fig:set_A_return_B}). Add an entry $(V, v_{old})$ to the log at index $i$, expressing that $A[i]$ was $v_{old}$ at version $V$. Increment the version of $AD$ to $V+1$. Set $array[i]$ in $AD$ to $v_{new}$. Create a new functional array $B$, with version $V+1$, which points to $AD$. Notice that both $A$ and $B$ point to $AD$. However, the version of $A$ is $V$ and the version of $AD$ is $V+1$, which indicates that $A$ is OLD. The version of $B$ is $V+1$ which indicates that $B$ is NEW.

Get on OLD array: Suppose we do $get(A, i)$ where $A$ is OLD and has version $V$. We binary search the log at index $i$ to find what value was stored in the array at version $V$.

\begin{figure}[!ht]
\centering
\includegraphics[scale=0.3]{set_old_array}
\nocaptionrule \caption{$D = set(A, 1, 9)$ copies $AD$ because $A$ is OLD}
\label{fig:set_old_array}
\end{figure}

Suppose we do $D = set(A, i,v)$ where $A$ is old (see figure \ref{fig:set_old_array}). Create a new ArrayData object $AD'$ (with a new array). Copy the values from the array in $AD$ to the array in $AD'$, and then set $array[i]$ to $v$ in $AD'$.

The logs are stored in a doubling array which doubles in size when full. This guarantees amortized $O(1)$ insertion of a log entry, and allows us to binary search with $O(\log{m})$ work, where $m$ is the number of log entries.

Suppose that the size of the array is $n$. Every $n$ times an ArrayData object is modified, we create a new ArrayData object, and copy the values over to the new ArrayData object. This ensures that $m \leq n$ i.e. we don't have too many log entries in an ArrayData object. Copying is amortized $O(1)$.

It follows that in the RAM model the work of set and get is $O(1)$ for a NEW array. Get in an OLD array involves a binary search and has work $O(\log{n})$. Set in an OLD array involves copying the array, and has work $O(n)$.

\section{Cost Dynamics}

Arrays are represented by $(l, D)$ where $l$ is a label that uniquely identifies the array, and $D$ contains the values in the array. $D[i]$ is the $i^{\text{th}}$ value in the array. If the set method has been applied on an array, the array is OLD (represented by $-$), otherwise the array is NEW (represented by $+$).

Array methods have different costs depending on whether the array is NEW or OLD. Let $C(l)$ be the number of get operations applied on array $A = (l, D)$ that had $O(1)$ work. We thread a store $\delta$ through each operation. $\delta$ maps array labels $l$ to $(+/-, c)$, where $c = C(l)$.

Our cost dynamics are defined by the following judgement, where $\delta$ is the store, $f$ is the array function being applied, $\delta'$ is the new store, $v$ is the value returned by applying $f$ on $args$, $w$ is the work, and $s$ is the span.
$$\delta; f \; args \Downarrow \delta'; v; w; s$$
The cost dynamics for the methods on a functional array of size $n$ are:
$$\frac{l = \text{new label} \;\;\; f(1) \Downarrow v_1; w_1; s_1; ... f(n) \Downarrow v_{n}; w_{n}; s_{n}}{\delta; \text{tabulate } f \; n \Downarrow \delta[l \mapsto (+, 0)]; (l, \bigcup_{i=1}^n [i \mapsto v_i]); 1+\sum w_i; 1+\max s_i}$$
$$\frac{\delta[l \mapsto (+, c)] \;\;\; A = (l,D) \;\;\; l' = \text{new label}}{\delta; \text{set} \; A \; i \; v \Downarrow \delta[l \mapsto (-, c), l' \mapsto (+, 0)]; (l', D[i \mapsto v]); 1; 1} \text{  (set-new)}$$
$$\frac{\delta[l \mapsto (-, c)] \;\;\; A = (l,D) \;\;\;  l' = \text{new label}}{\delta; \text{set} \; A \; i \; v \Downarrow \delta[l \mapsto (-, c), l' \mapsto (+, 0)]; (l', D[i \mapsto v]); n; 1} \text{  (set-old)}$$
$$\frac{\delta[l \mapsto (+, c)] \;\;\; A = (l,D)}{\delta; \text{get} \; A \; i \Downarrow \delta[l \mapsto (+, c+1)]; D[i]; 1; 1} \text{  (get-new)}$$
$$\frac{\delta[l \mapsto (-, c)] \;\;\; A = (l,D)}{\delta; \text{get} \; A \; i \Downarrow \delta; D[i]; \log{n}; 1} \text{  (get-old)}$$

The fork-join cost semantics are the most interesting. We give the cost dynamic using 2 helper functions. Let $L(\delta)$ denote the set of labels in the store $\delta$. Let $f(n)$ be the cost of calling get on an OLD array and $g(n)$ be the cost of calling set on an OLD array.
\begin{align*}
\text{def NEW}&\text{-MAP: }L(\delta_1) \cup L(\delta_2) \to (+/-, \text{int}) = \\
\text{Case } &\delta[l \mapsto (-,c)]: (-, c)  \\
\text{Case } &\delta[l \mapsto (+,c)]:  \\
&\text{Case } \delta_1[l \mapsto (+,c+x)], \delta_2[l \mapsto (+,c+y)]: (+, c+x+y) \\
&\text{Case } \delta_1[l \mapsto (+,c+x)], \delta_2[l \mapsto (-,c+y)]: (-, c+y) \\
&\text{Case } \delta_1[l \mapsto (-,c+x)], \delta_2[l \mapsto (+,c+y)]: (-, c+x) \\
&\text{Case } \delta_1[l \mapsto (-,c+x)], \delta_2[l \mapsto (-,c+y)]: (-, c) \\
\text{Else: } & \\
&\text{Case } \delta_1[l \mapsto (s,c)]: (s,c) \\
&\text{Case } \delta_2[l \mapsto (s,c)]: (s,c) \\
\text{def EXT}&\text{RA-WORK: }L(\delta_1) \cup L(\delta_2) \to int = \\
\text{Case } &\delta[l \mapsto (+,c)]:  \\
&\text{Case } \delta_1[l \mapsto (+,c+x)], \delta_2[l \mapsto (+,c+y)]: 0 \\
&\text{Case } \delta_1[l \mapsto (+,c+x)], \delta_2[l \mapsto (-,c+y)]: x f(n) \\
&\text{Case } \delta_1[l \mapsto (-,c+x)], \delta_2[l \mapsto (+,c+y)]: y f(n) \\
&\text{Case } \delta_1[l \mapsto (-,c+x)], \delta_2[l \mapsto (-,c+y)]: (x+y)f(n) + g(n) \\
\text{Else: } & 0
\end{align*}											
$$\delta' = \bigcup_{l \in L(\delta_1) \cup L(\delta_2)} \text{NEW-MAP} (l)$$
$$w' = \sum_{l \in L(\delta_1) \cup L(\delta_2)} \text{EXTRA-WORK} (l)$$
Then, the fork-join cost semantics are:
$$\delta; e_1 || e_2 \Downarrow \delta'; (v_1, v_2); 1 + w_1 + w_2 + w'; 1 + \max(s_1, s_2)$$

\section{Machine Model}

Our target language is a Java-like language augmented with tabulate, fork-join, lambda functions, and link load store conditional (LLSC). Fork-join spawns two threads, one thread executing each sub-expression. Link load store conditional (LLSC) takes an (address, old value, new value) tuple as argument. LLSC first checks that the value at address is the same as old value, if it is different then LLSC returns false. Otherwise LLSC atomically stores new value at address and returns true, however if the value at address was modified between the load and store, LLSC returns false.

Our target language is run on a P processor machine. In a single time step the machine takes at most P runnable instructions and executes them.  The effect of executing the ($\leq P$) instructions in a time step is the same as some (arbitrary) sequential ordering of the instructions. We assume that the machine does not reorder instructions in the target language. In reality, languages like Java have very relaxed consistency models, and we deal with this in our real implementation using memory fences.

The link load store conditional architecture deserves special mention. Call an LLSC operation pending if the load operation has completed but the store operation has not. Multiple threads can load the value at a particular address and compare it with old value in a single time step. Only a single store operation to a particular address can be executed at a time step. However, after the store operation, in each time step the bus arbiter notifies all pending LLSC operations to the same memory address that they have failed. This assumption is reasonable because all LLSC instructions are typically processed by the bus arbiter, so the bus arbiter can keep track of pending LLSC instructions to each memory address.

\section{Proof of Correctness}

We begin by proving that functional arrays work correctly when executed sequentially. Specifically, in any fork join, we assume that the left fork is evaluated before the right fork.

\begin{lemma}
The implementation in section 4 correctly implements the dynamics of functional arrays when executed sequentially.
\end{lemma}

\begin{proof} 
We define a data structure invariant, prove the invariant by structural induction on the array functions, and finally use the invariant to show that get always returns required value.

Data structure invariant: consider arbitrary array $A$ in the environment. Let $A[i]$ denote the actual value of $A$ at index $i$. $A$ points to a structure $AD$ and has version $v$. If $v$ is the same as the version of $AD$, then $AD.array[i]$ and $A[i]$ are the same. Otherwise, consider arbitrary index $i$ and let $L$ be the log array of $AD$ at index $i$. If the version of all log entries in $L$ is less than $v$, then $A[i]$ is the same as $AD.array[i]$. Otherwise, let $(v,e)$ be the log entry with the lowest version $v$ s.t. $v \geq V$. Then $A[i] = e$.

We prove that this invariant holds by structural induction. (Add proof outline)

This then trivially gives us that get returns the correct value.

\end{proof}

\begin{lemma}
Functional arrays work correctly when evaluated sequentially.
\end{lemma}

\begin{proof}

\end{proof}

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.

\bibliography{references}

\end{document}

### 2016-02-22 ArraySpeedTest.java

@@ -205,29 +205,29 @@ public class ArraySpeedTest {
     // Tests comparing persistent arrays with regular arrays.
     System.out.println("\nseq_read_test");
     for (int i = 0; i < num_trials; i++) {
-      long st_time = st_seq_read_test(8000000, 50000000);
-      long persistent_time = persistent_seq_read_test(8000000, 50000000);
+      long st_time = st_seq_read_test(3000000, 15000000);
+      long persistent_time = persistent_seq_read_test(3000000, 15000000);
       System.out.println(st_time + "," + persistent_time);
     }
 
     System.out.println("\nrnd_read_test");
     for (int i = 0; i < num_trials; i++) {
-      long st_time = st_rnd_read_test(100000, 10000000);
-      long persistent_time = persistent_rnd_read_test(100, 10000000);
+      long st_time = st_rnd_read_test(3000000, 15000000);
+      long persistent_time = persistent_rnd_read_test(3000000, 15000000);
       System.out.println(st_time + "," + persistent_time);
     }
 
     System.out.println("\nseq_write_test");
     for (int i = 0; i < num_trials; i++) {
-      long st_time = st_seq_write_test(100000, 5000000);
-      long persistent_time = persistent_seq_write_test(100, 10000000);
+      long st_time = st_seq_write_test(3000000, 5000000);
+      long persistent_time = persistent_seq_write_test(3000000, 5000000);
       System.out.println(st_time + "," + persistent_time);
     }
 
     System.out.println("\nrnd_write_test");
     for (int i = 0; i < num_trials; i++) {
-      long st_time = st_rnd_write_test(100000, 5000000);
-      long persistent_time = persistent_rnd_write_test(100, 1000000);
+      long st_time = st_rnd_write_test(3000000, 5000000);
+      long persistent_time = persistent_rnd_write_test(3000000, 5000000);
       System.out.println(st_time + "," + persistent_time);
     }
 
@@ -235,28 +235,28 @@ public class ArraySpeedTest {
     // persistent arrays.
     System.out.println("\nrnd_read_test (old vs new)");
     for (int i = 0; i < num_trials; i++) {
-      long old_time = rnd_old_read_test(100000, 900000, 900000);
-      long new_time = rnd_new_read_test(100000, 900000, 900000);
+      long old_time = rnd_old_read_test(2100000, 20000000, 5000000);
+      long new_time = rnd_new_read_test(2100000, 20000000, 5000000);
       System.out.println(old_time + "," + new_time);
     }
 
[…]

### 2016-02-26 Paper/sigplanconf-template.tex
Add correctness outline

@@
 \usepackage{amsthm}
+\usepackage{listings}
 
@@
 \newtheorem{lemma}[theorem]{Lemma}
+\theoremstyle{definition}
+\newtheorem{definition}{Definition}[section]
 
@@
 
+\section{Concurrent Implementation}
+
+We implement the functional array functions using an imperative target language.
+
+\subsection{Concurrency Model}
+
+We assume a sequential consistency model of concurrency.
+That is, when analyzing the effects of a program, we assume that at each time step exactly one pending instruction is executed.
+Consider the execution of a program $P$.
+We label the time steps in the execution from 1 to $t$.
+
+\begin{definition}
+Consider an object $A$.
+Suppose that the program invokes methods $M_1$, $M_2$, ..., $M_n$ on $A$.
+Consider arbitrary $i$, and suppose that the instructions in $M_i$ were executed at times $s_1$, $s_2$, ..., $s_m$.
+We say that $A$ is \emph{linearizable} if the effect of the program is as if $M_i$ executed atomically at some time $s_j$, and this holds for all $M_i$.
+\end{definition}
+
+\begin{definition}
+Suppose that first instruction in function call $f$ was executed at time $s_1$ and the last instruction at time $e_1$.
+Suppose that first instruction in function call $g$ was executed at time $s_2$ and the last instruction at time $e_2$.
+We say the function calls do not \emph{overlap} if the intervals $[s_1, e_1]$ and $[s_2, e_2]$ do not overlap.
+\end{definition}
+
+Our programs have access to the link load store conditional (LLSC) function.
+LLSC takes an (address, old value, new value) tuple as argument.
+LLSC first checks that the value at address is the same as old value, if it is different then LLSC returns false.
+Otherwise LLSC atomically stores new value at address and returns true, however if the value at address was modified between the load and store, LLSC returns false.
+
+\subsection{Push Arrays}
+
+We use an auxiliary data structure, \emph{PushArrays}, to store log entries.
+PushArrays are initially empty and support 3 methods.
+Suppose we have a PushArray $A$.
+$A.push(e)$ inserts entry $e$, $A.size()$ returns the number of entries inserted, $A.get(i)$ returns the $i^{\text{th}}$ entry inserted.
+Note that $get$ is only defined between indices 0 and $A.size()-1$.
+All operations are amortized $O(1)$.
+
+PushArrays can be used semi-concurrently.
+At most one thread can execute $push$ at any time, but multiple threads can call $size$ and $get$.
+In a PushArray $A$, $A.size$ stores the number of entries and $A.data$ references an array that stores the actual entries.
+Pseudocode for $A.push(e)$ is shown below.
+
+\begin{lstlisting}[language=Java]
+if (isFull(A.data)) {
+  newData = new Array(A.data.capacity * 2);
+  copyValues(from = A.data, to = newData);
+  A.data = newData;
+}
+A.data[A.size] = e;
+A.size += 1;
+\end{lstlisting}
+
+\begin{theorem}
+Given that different calls to push (on the same PushArray) do not overlap, PushArrays are linearizable.
+\end{theorem}
+
+\begin{proof}
+The linearization point of $push$ is $A.size$ $+=$ $1$, and the linearization point of the other methods is their single instruction.
+In push, expanding the capacity of $A.data$ when it is full does not interfere with accessing $A.data$ because we copy values to the new array before pointing $A.data$ to the new array.
+The effects of adding $e$ are only observable after we increase $size$, since, by the specification of PushArrays, the result of calling $A.get(i)$ where $i \geq A.size()$ is not defined.
+\end{proof}
+
+\subsection{Functional Array Implementation}
+
+Suppose we have a FunctionalArray $A$.
+$A$ points to an ArrayData object which is referenced by $A.data$.
+The (imperative) array in $A.data$ is referenced by $A.data.values$, and the log entries corresponding to the $i^{\text{th}}$ index are stored in a PushArray $A.data.undo\_lists[i]$.
+We first explain the implementation of $A.set(pos, val)$.
+
+We use a link load store conditional to ensure that only 1 thread can modify an ArrayData object at any time.
+If $m$ threads try to modify $A.data$, then $m-1$ threads will fail the LLSC (returning false).
+Instead of modifying $A.data$, these threads will create a new ArrayData object, and will copy over values from $A.data$, as shown below.
+
+\begin{lstlisting}[language=Java]
+FunctionalArray newA;
+newA.version = A.version + 1;
+if (!LLSC(&A.data.version, A.version, 
+	newA.version)) {
+  // Create a new ArrayData object AD
+  // Copy A.data.values to AD.values
+  AD.values[pos] = val;
+  newA.data = AD;
+  return newA;
+}
+\end{lstlisting}
+
+One of the threads trying to modify the ArrayData object will get past the LLSC, will insert a log entry and then modify $A.data.values$.
+The ordering of inserting a log entry and modifying $A.data.values$ matters in the proof of get The pseudocode for this step is shown below.
+
+\begin{lstlisting}[language=Java]
+newA.data = A.data;
+old_value = newA.data.values[pos];
+undo_list = newA.data.undo_lists[pos];
+undo_list.push((A.version, old_value));
+newA.data.values[pos] = val;
+return newA;
+\end{lstlisting}
+
+Next we explain the implementation of $A.get(pos)$.
+
+\begin{lstlisting}[language=Java]
+guess_val = A.data.values[pos];
+if (A.version == A.data.version) {
+  return guess_val;
+}
+
+undo_list = this.data.undo_lists[pos];
+if (undo_list.size() == 0) {
+  return guess_val;
+}
+
+upper = undo_list.size() - 1;
+if (undo_list[upper].version < A.version) {
+  return guess_val;
+}
+\end{lstlisting}
+
+The last step is to binary search $undo\_list$, between the indices 0 and $upper$ (inclusive of 0 and upper).
+We find the log entry $L$ with the smallest version $X$ such that $A.version \leq X$.
+We return $L.value$.
+
+The implementation of tabulate is straightforward - just allocate a new array and fill the array with the required values.
+
+\subsection{Proof of Correctness}
+
+\begin{theorem}
+Functional Arrays are linearizable.
+\end{theorem}
+
+\begin{proof}
+We can choose any instruction in the execution of get as the linearization point - we arbitrarily choose the first instruction.
+The linearization point of set is the LLSC.
+Any instruction in tabulate can be its linearization point - we omit discussion of tabulate because the details are uninteresting.
+
+First, we prove the linearization point of get. (fill this out)
+
+Next, we prove the linearization point of set. (fill this out)
+
+\end{proof}
+
+In a previous section, we gave the evaluational (big step) dynamics of functional arrays.
+Alternatively, we could give the non-deterministic structural (little step) dynamics which describes the steps in the execution of a program in the source language.
+At each fork-join, if both sides of the fork can take a step, we could take a step on either side.
+
+\begin{theorem}
+Suppose that the execution of a program in the target language has a certain linear order of function calls $M_1$, $M_2$, ...
+$M_n$.
+Then there exists a (little step) transition sequence in the source language where the functions are called in the same order.
+\end{theorem}
+
+\begin{proof}
+Fill this out. Use induction.
+\end{proof}
+
+\begin{theorem}
+Suppose that the execution of a program in the target language has a certain linear order of function calls $M_1$, $M_2$, ...
+$M_n$ and consider a little step transition sequence in the source language where the functions are called in the same order $F_1$, $F_2$, ..., $F_n$.
+For any $F_i$, if $F_i$ is a get, then the value returns by $M_i$ and $F_i$ are the same.
+\end{theorem}
+
+\begin{proof}
+This is the sequential proof of correctness.  
+Let the environment $E$ contain all the arrays. 
+We prove this by induction on the number of functional calls.
+The base case, when there are no function calls, is trivial.
+For the 
+\end{proof}
+
+\begin{theorem}
+The non-deterministic structural dynamics and the evaluational dynamics produce the same result in the source language.
+\end{theorem}
+
+\begin{proof}
+This is a standard programming languages proof so we omit the proof.
+\end{proof}
+
+We have shown that there exists a transition sequence that our implementation correctly simulates, and that all transition sequences are equivalent to the evaluational dynamics, so our implementation correctly simulates the evaluational dynamics.
+
+\section{Correctness}
+
 \section{Machine Model}
@@
 
-\section{Proof of Correctness}
-
-We begin by proving that functional arrays work correctly when executed sequentially. Specifically, in any fork join, we assume that the left fork is evaluated before the right fork.
-
-\begin{lemma}
-The implementation in section 4 correctly implements the dynamics of functional arrays when executed sequentially.
-\end{lemma}
-
-\begin{proof} 
-We define a data structure invariant, prove the invariant by structural induction on the array functions, and finally use the invariant to show that get always returns required value.
-
-Data structure invariant: consider arbitrary array $A$ in the environment.
-Let $A[i]$ denote the actual value of $A$ at index $i$.
-$A$ points to a structure $AD$ and has version $v$.
-If $v$ is the same as the version of $AD$, then $AD.array[i]$ and $A[i]$ are the same.
-Otherwise, consider arbitrary index $i$ and let $L$ be the log array of $AD$ at index $i$.
-If the version of all log entries in $L$ is less than $v$, then $A[i]$ is the same as $AD.array[i]$.
-Otherwise, let $(v,e)$ be the log entry with the lowest version $v$ s.t.
-$v \geq V$.
-Then $A[i] = e$.
-
-We prove that this invariant holds by structural induction. (Add proof outline)
-
-This then trivially gives us that get returns the correct value.
-
-\end{proof}
-
-\begin{lemma}
-Functional arrays work correctly when evaluated sequentially.
-\end{lemma}
-
-\begin{proof}
-
-\end{proof}
-
 % We recommend abbrvnat bibliography style.

### 2016-02-26 Paper/sigplanconf-template.tex
Sequential proof

@@
 $M_n$ and consider a little step transition sequence in the source language where the functions are called in the same order $F_1$, $F_2$, ..., $F_n$.
-For any $F_i$, if $F_i$ is a get, then the value returns by $M_i$ and $F_i$ are the same.
+For any $F_i$, if $F_i$ is a get, then the value returned by $M_i$ and $F_i$ are the same.
 \end{theorem}
@@
 \begin{proof}
-This is the sequential proof of correctness.  
-Let the environment $E$ contain all the arrays. 
-We prove this by induction on the number of functional calls.
-The base case, when there are no function calls, is trivial.
-For the 
+Note that this is the sequential proof of correctness.
+Consider arbitrary $i$ and suppose that functional calls $F_j$ for $j \leq i$ have executed in the source language and function calls $M_j$ for $j \leq i$ have executed in the target language.
+Consider an arbitrary array $A'$ in the source language and corresponding array $A$ in the target language.
+We claim that the following data structure invariant holds:
+
+Consider arbitrary index $i$ and suppose that $A'[i]$ has value $val$.
+Suppose that $A$ has version $V$ and is pointing to an ArrayData object $AD$.
+Either there exists some version in $AD.logs[i]$ which is greater than $V$ in which case there exists a log entry $(v', val)$ where $v'$ is the smallest version in $AD.logs[i]$ that is $\geq V$.
+Or $AD.values[i]$ is val.
+
+The data structure invariant can be proved by induction, we present a sketch.
+The base case is trivial.
+In the inductive case, get does not mutate any array.
+When we do a set operation, the invariant holds for the new array because we modify the value in its ArrayData object.
+It holds for other arrays in the target language that might have been mutated because we insert log entries.
+
+Given that the data structure invariant holds, it is easy to see that any call to get returns the same value in the source and target language (maybe we need a little more detail here?).
 \end{proof}

### 2016-03-09 Paper/sigplanconf-template.tex
Add sequential proof

--- 2016-02-26 Paper/sigplanconf-template.tex
@@
 
-Many functional programming languages use monads to build arrays. However, monads do not allow accessing old versions of an array.
+There is a lot of prior work on functional arrays, but previous approaches have significant limitations.
+Many functional programming languages use monads to build arrays.
+However, monads do not allow accessing old versions of an array.
 
@@
 
-The dynamics for the other two array functions are standard.
+$get(A,i)$ returns the value at the $i$th index of $A$.
+$set(A,i,v)$ returns a new array where the value at index $i$ is $v$ and the value at all other indices is the same as in $A$.
+The evaluation of a simple program in our language is shown below:\\
+
+$(\lambda A. \; get(A, 2) + get(A,3)) set(tabulate(\lambda i . \; i, 5), 2, 10)$
+
+$(\lambda A. \; get(A, 2) + get(A,3)) set([1,2,3,4,5], 2, 10)$
+
+$(\lambda A. \; get(A, 2) + get(A,3)) [1,10,3,4,5]$
+
+$get([1,10,3,4,5], 2) + get([1,10,3,4,5],3) $
+
+$10+3$
+
+$13$
 
@@
 if (!LLSC(&A.data.version, A.version, 
-	newA.version)) {
+	newA.version) OR
+    A.version % A.size == 0) {
   // Create a new ArrayData object AD
-  // Copy A.data.values to AD.values
+  // Forall i = 1:n, AD.values[i] = get(A, i)
   AD.values[pos] = val;
@@
 
-One of the threads trying to modify the ArrayData object will get past the LLSC, will insert a log entry and then modify $A.data.values$.
-The ordering of inserting a log entry and modifying $A.data.values$ matters in the proof of get The pseudocode for this step is shown below.
+At most one of the threads trying to modify the ArrayData object will get past the LLSC, will insert a log entry and then modify $A.data.values$.
+The ordering of inserting a log entry and modifying $A.data.values$ matters in the proof of get.
+The pseudocode for this step is shown below.
 
@@
 \begin{proof}
-We can choose any instruction in the execution of get as the linearization point - we arbitrarily choose the first instruction.
+
+The linearization point of get is when we compare if A.version and AD.version are the same.
 The linearization point of set is the LLSC.
@@
 
+Linearizability allows us to assume that array operations happen atomically.
+Note that all non array operations in our source language are executed in a single time step, so any well formed program is linearizable.
+In fact, since linearizability composes, we can show that more complicated languages that use our arrays are also linearizable.
+
 In a previous section, we gave the evaluational (big step) dynamics of functional arrays.
@@
 At each fork-join, if both sides of the fork can take a step, we could take a step on either side.
-
-\begin{theorem}
-Suppose that the execution of a program in the target language has a certain linear order of function calls $M_1$, $M_2$, ...
-$M_n$.
-Then there exists a (little step) transition sequence in the source language where the functions are called in the same order.
-\end{theorem}
-
-\begin{proof}
-Fill this out. Use induction.
-\end{proof}
-
-\begin{theorem}
-Suppose that the execution of a program in the target language has a certain linear order of function calls $M_1$, $M_2$, ...
-$M_n$ and consider a little step transition sequence in the source language where the functions are called in the same order $F_1$, $F_2$, ..., $F_n$.
-For any $F_i$, if $F_i$ is a get, then the value returns by $M_i$ and $F_i$ are the same.
-\end{theorem}
-
-\begin{proof}
-This is the sequential proof of correctness.  
-Let the environment $E$ contain all the arrays. 
-We prove this by induction on the number of functional calls.
-The base case, when there are no function calls, is trivial.
-For the 
-\end{proof}
 
@@
 \begin{proof}
-This is a standard programming languages proof so we omit the proof.
+This is a standard implicit parallelism type theorem (ref: PFPL) so we omit the details.
 \end{proof}
 
-We have shown that there exists a transition sequence that our implementation correctly simulates, and that all transition sequences are equivalent to the evaluational dynamics, so our implementation correctly simulates the evaluational dynamics.
+Now, to show that our implementation correctly implements the source language, it suffices to show that any valid execution of our implementation produces the same result as a little step transition sequence in the source language (since all transition sequences produce the same value as the evaluational dynamics).
 
-\section{Correctness}
+\begin{theorem}
+For any program $P$ in the source language, executing our implementation $I$ of $P$ produces the same result as some transition sequence of $P$. Note that $I$ is a program in the target language.
+\end{theorem}
+
+\begin{proof}
+We prove the claim by induction. We focus on the part of the proof dealing with arrays and give a sketch of the rest.
+
+Our inductive hypothesis is that after $t$ time steps in an arbitrary execution of $I$, there exists a transition sequence $S$ s.t.:
+\begin{enumerate}
+\item $I$ becomes $I'$ after this particular execution, $S$ takes $P$ to $P'$
+\item $I'$ and $P'$ are equivalent in the following sense.
+Any non array value is identical in $I'$ and $P'$.
+For any abstract array $A$ in $P'$ there exists a corresponding representation $A_I$ s.t.
+the following holds.
+Consider arbitrary index $i$ and suppose that $A[i]$ has value $val$.
+Suppose that $A_I$ has version $V$ and is pointing to an ArrayData object $AD$.
+If there exists some version in $AD.logs[i]$ which is greater than $V$ then there exists a log entry $(v', val)$ where $v'$ is the smallest version in $AD.logs[i]$ that is $\geq V$.
+If no version in $AD.logs[i]$ is greater than $V$, then $AD.values[i]$ is val.
+\end{enumerate}
+
+The base case holds since $P$ and $I$ are initially identical.
+For the inductive step, when $I'$ takes a step, we take the corresponding step in $P$ to get a new transition sequence $S'$.
+We know that such a step exists because $I'$ and $P'$ are equivalent.
+If the step is not an array function then the IH holds since the step will be executed in the same way in the source and target languages.
+
+Suppose that the next step is tabulate.
+By the IH, tabulate is called with the same arguments $n$ and $f$ in the source language and the implementation.
+In the source language, tabulate returns an array $A$ with $A[i] = f(i)$ for all $1 \leq i \leq n$.
+In the implementation, tabulate returns an array that points to a new ArrayData object $AD$ with empty logs and $AD.values[i] = f(i)$ for all $1 \leq i \leq n$.
+So the inductive step holds.
+
+Suppose that the next step is get.
+By the IH, get is called on equivalent arrays and the same index $i$ in the source language and implementation.
+It is easy to trace through the implementation of get and see that get returns the same value in the source language and implementation.
+
+Suppose that the next step is set.
+Suppose that in the target language we execute $set(A_I, i, v)$.
+Then by IH, in the source language the corresponding step is $set(A, i, v)$ where $A$ and $A_I$ are equivalent.
+Denote $set(A, i, v)$ by $B$ and $set(A_I, i, v)$ by $B_I$.
+We need to show that $B$ and $B_I$ are equivalent, and all arrays that were previously equivalent are still equivalent.
+
+Case 1: we create a new ArrayData object $AD$ in set.
+In this case, for all $j \neq i$ we set $AD.values[j]$ to $get(A_I, j)$ which by the IH is the same as $A[j]$.
+Since we only modified index $i$, $j \neq i \Rightarrow A[j] = B[j]$.
+Also, we set $AD.values[i]$ to $v$, so $AD.values[i] = B[i]$.
+Since all logs are empty, this means that $B$ and $B_I$ are equivalent.
+Since we create a new ArrayData object, we do not modify any other arrays, so arrays that were previously equivalent are still equivalent.
+
+Case 2: we modify the ArrayData object $AD$ that $A_I$ points to.
+This means that $A.version = AD.version$.
+We can show that all log entries in $AD$ have version $< AD.version$, and are in increasing order.
+Then the IH implies that before modifying $AD$, $AD.values[j] = A[j]$ for all indices $j$.
+So after setting $AD.values[i]$ to $v$, $AD.values[j] = B[j]$ for all $j$, which means $B_I$ and $B$ are equivalent.
+
+Consider arbitrary abstract array $C$ in the source language and corresponding representation $C_I$ in the implementation with $C_I \neq B_I$.
+If $C_I$ did not point to $AD$ then $C_I$ remains equivalent to $C$.
+Regardless, we note that for all indices $j \neq i$, $C_I$ remains the same and is equivalent to $C$.
+At index $i$, if the value at $C[i]$ was previously stored in a log entry, then $C_I$ and $C$ remain equivalent since log entries are added in increasing version numbers.
+If the value at $C[i]$ was stored in $AD[i]$ then we add a log entry at index $i$ so $C_I$ and $C$ remain equivalent.
+
+\end{proof}
 

### 2016-03-10 Paper/paper.tex
Add explanations, rename files

--- 2016-03-09 Paper/sigplanconf-template.tex
@@
 Accessing old versions of arrays can be useful for efficient checkpointing, logging, and event handling.
-In this paper we introduce an efficient, parallel, functional array that allows $O(1)$ reads and writes to the most recent version of the array.
 
-\section{Previous Approaches}
+\subsection{Previous Approaches}
 
-There is a lot of prior work on functional arrays, but previous approaches have significant limitations.
+There has been a lot of great work on functional arrays, but previous approaches have some limitations.
 Many functional programming languages use monads to build arrays.
@@
 
-Another approach is to use compiler based reference counting.
+Another creative approach is to use compiler based reference counting.
 If the number of references to an array is provably one, the compiler allows the array to mutate, otherwise the compiler copies the array before writing to the array.
-This approach makes it difficult for programmers to reason about the time complexity of array operations because it depends on the compiler.
+However, this approach makes it difficult for programmers to reason about the time complexity of array operations because it depends on the compiler.
 Furthermore, if multiple variables reference the same array, the array is copied even if all writes are to the most recent version.
@@
 O' Neill's functional arrays guarantee $O(1)$ access and insertions when used like an imperative array.
-However, the arrays use version trees which are inefficient in practice.
+However, the arrays explicitly implement version trees which are inefficient in practice.
 Additionally, the reads and writes to the most recent version of the array might take $O(\log{n})$ work if the array is not used like an imperative array.
@@
 
+\subsection{Our Approach}
+
+Our arrays support 3 methods: tabulate, get, and set. The dynamics are functional, in particular set returns a new array and leaves the original array intact.
+
+\begin{figure}[!ht]
+\centering
+\includegraphics[scale=0.5]{leaf_interior_intro}
+\nocaptionrule \caption{Example usage of functional arrays}
+\label{fig:leaf_interior_intro}
+\end{figure}
+
+In figure \ref{fig:leaf_interior_intro}, A and C are \emph{interior} arrays and B and D are \emph{leaf} arrays.
+Calling get and set on leaf arrays costs $O(1)$ but calling get and set on interior arrays is more expensive.
+In the cost dynamics, we thread a store that keeps track of whether an array is an interior or leaf array.
+
+We also give and prove fork join cost dynamics for our arrays.
+The dynamics for arrays used concurrently are non trivial because one thread might call get on a leaf array while another thread calls set.
+To deal with this, we compute the work of each fork separately, and then charge additional work at the join point if an array is used in both forks.
+
+In our implementation of a functional array, we keep an array of values (multiple functional arrays might point to the same array).
+For each index of the array, we also keep a history of values that were previously at that index.
+To access an element in an interior array, we need to search through the history of values.
+
+We give a wait-free concurrent implementation that uses careful synchronization.
+We prove linearizability of the implementation and use that to prove that arrays work correctly when used concurrently.
+Finally, we give a provable implementation for our cost dynamics.
+
 \section{Dynamics}
@@
 We use a standard applicative-order language defined as follows:
-$$L = x \; | \; c  \; | \; \lambda x.e \; | \; f \; x \; | \; e_1 || e_2 \; | \; \text{if } e_1 \text{ then } e_2 \text{ else } e_3$$
+$$e = x \; | \; c  \; | \; \lambda x.e \; | \; e_1 e_2 \; | \; e_1 || e_2 \; | \; \text{if } e_1 \text{ then } e_2 \text{ else } e_3$$
 
-$L$ contains the usual arithmetic types, such as the natural numbers, and numerical operations such as sums and products. Dynamics for function application and fork-join are given as examples.
+$L$ contains the usual arithmetic types, such as the natural numbers, and numerical operations such as sums and products.
+Our dynamics are given by the judgement $e \Downarrow v$ which means that expression $e$ evaluates to $v$.
+Dynamics for function application and fork-join are given as examples.
+
 $$\frac{e_1 \Downarrow v_1 \;\;\; e_2 \Downarrow v_2}{e_1 || e_2 \Downarrow (v_1, v_2)} \text{ (fork-join)}$$
@@
 $13$
+
+\section{Cost Dynamics}
+
+The dynamics given in the previous section are functional.
+However, the costs of get and set depend on whether the array is a leaf array or an internal array.
+To deal with this, we thread a store through the cost dynamics to keep track of whether each array is a leaf or internal array.
+
+The cost dynamics of fork-join pose an additional challenge.
+An array can be used concurrently by multiple threads and the costs could depend on the order in which the instructions are interleaved.
+Figure \ref{fig:fork_join_intuition} shows 3 ways a leaf array $A$ might be used.
+In particular, one side of the fork might call get on $A$ while the other side calls set (see the center panel).
+The calls to get on the left branch would be $O(1)$ if and only if they execute before the call to set on the right branch.
+
+\begin{figure}[!ht]
+\centering
+\includegraphics[scale=0.3]{fork_join_intuition}
+\nocaptionrule \caption{Different ways an array might be used in a fork join}
+\label{fig:fork_join_intuition}
+\end{figure}
+
+We cannot make any assumptions about how the instructions in different branches of a fork are interleaved, so we give the worst case time complexity over all possible interleavings.
+We first compute the work of each fork separately.
+Then for arrays that are used in both forks, we add additional work at the join point.
+To compute the additional work at the join point, we keep track of the number of gets that cost $O(1)$ on each side of the fork.
+If the array was modified on the other side of the fork, then the modification might have happened before the get, so we need to charge additional work for the get.
+
+Our cost dynamics are defined by the following judgement, where $\delta$ is the store, $e$ is the expression, $\delta'$ is the new store, $v$ is the value returned by evaluating $e$, $w$ is the work, and $s$ is the span.
+$$\delta; f \; args \boldsymbol\Downarrow \delta'; v; w; s$$
+
+We give the cost dynamics for expressions in the language: (insert cost dynamics for standard functions later)
+
+Arrays are represented by $(l, V)$ where $l$ is a label that uniquely identifies the array, and $V$ contains the values in the array.
+$V[i]$ is the $i^{\text{th}}$ value in the array.
+If the set method has been applied on an array, the array is a leaf array (represented by $-$), otherwise the array is an interior array (represented by $+$).
+
+Array methods have different costs depending on whether the array is a leaf or interior array.
+Let $C(l)$ be the number of get operations applied on array $A = (l, V)$ that had $O(1)$ work.
+We thread a store $\delta$ through each operation.
+$\delta$ maps array labels $l$ to $(+/-, c)$, where $c = C(l)$.
+Let $f(n)$ be the work of calling get on an interior array and $g(n)$ be the work of calling set on an interior array.
+
+The cost dynamics for the methods on a functional array of size $n$ are:
+$$\frac{l = \text{new label} \;\;\; f(1) \Downarrow v_1; w_1; s_1; ...
+f(n) \Downarrow v_{n}; w_{n}; s_{n}}{\delta; \text{tabulate } f \; n \Downarrow \delta[l \mapsto (+, 0)]; (l, \bigcup_{i=1}^n [i \mapsto v_i]); 1+\sum w_i; 1+\max s_i}$$
+$$\frac{A = (l,V) \;\;\; \delta[l \mapsto (+, c)] \;\;\; l' = \text{new label}}{\delta; \text{set} \; A \; i \; v \Downarrow \delta[l \mapsto (-, c), l' \mapsto (+, 0)]; (l', V[i \mapsto v]); 1; 1} \text{  (set-leaf)}$$
+$$\frac{A = (l,V) \;\;\; \delta[l \mapsto (-, c)] \;\;\;  l' = \text{new label}}{\delta; \text{set} \; A \; i \; v \Downarrow \delta[l \mapsto (-, c), l' \mapsto (+, 0)]; (l', V[i \mapsto v]); g(n); 1} \text{  (set-interior)}$$
+$$\frac{A = (l,V) \;\;\; \delta[l \mapsto (+, c)]}{\delta; \text{get} \; A \; i \Downarrow \delta[l \mapsto (+, c+1)]; V[i]; 1; 1} \text{  (get-leaf)}$$
+$$\frac{A = (l,V) \;\;\; \delta[l \mapsto (-, c)]}{\delta; \text{get} \; A \; i \Downarrow \delta; V[i]; f(n); 1} \text{  (get-interior)}$$
+
+Set creates a new label and array and modifies the store to indicate that the newly created array is a leaf array, and the array set was called on has become an interior array.
+The work for set is higher for interior arrays.
+Get returns the value at the desired index, if the array is a leaf then the operation costs $O(1)$ but we increment the counter of $O(1)$ gets in the store,
+
+The fork-join cost semantics are the most interesting. We give the cost dynamic using 2 helper functions. Let $L(\delta)$ denote the set of labels in the store $\delta$.
+Simplify the fork join cost dynamics
+\begin{align*}
+\text{def NEW}&\text{-MAP: }L(\delta_1) \cup L(\delta_2) \to (+/-, \text{int}) = \\
+\text{Case } &\delta[l \mapsto (-,c)]: (-, c)  \\
+\text{Case } &\delta[l \mapsto (+,c)]:  \\
+&\text{Case } \delta_1[l \mapsto (+,c+x)], \delta_2[l \mapsto (+,c+y)]: (+, c+x+y) \\
+&\text{Case } \delta_1[l \mapsto (+,c+x)], \delta_2[l \mapsto (-,c+y)]: (-, c+y) \\
+&\text{Case } \delta_1[l \mapsto (-,c+x)], \delta_2[l \mapsto (+,c+y)]: (-, c+x) \\
+&\text{Case } \delta_1[l \mapsto (-,c+x)], \delta_2[l \mapsto (-,c+y)]: (-, c) \\
+\text{Else: } & \\
+&\text{Case } \delta_1[l \mapsto (s,c)]: (s,c) \\
+&\text{Case } \delta_2[l \mapsto (s,c)]: (s,c) \\
+\text{def EXT}&\text{RA-WORK: }L(\delta_1) \cup L(\delta_2) \to int = \\
+\text{Case } &\delta[l \mapsto (+,c)]:  \\
+&\text{Case } \delta_1[l \mapsto (+,c+x)], \delta_2[l \mapsto (+,c+y)]: 0 \\
+&\text{Case } \delta_1[l \mapsto (+,c+x)], \delta_2[l \mapsto (-,c+y)]: x f(n) \\
+&\text{Case } \delta_1[l \mapsto (-,c+x)], \delta_2[l \mapsto (+,c+y)]: y f(n) \\
+&\text{Case } \delta_1[l \mapsto (-,c+x)], \delta_2[l \mapsto (-,c+y)]: (x+y)f(n) + g(n) \\
+\text{Else: } & 0
+\end{align*}											
+$$\delta' = \bigcup_{l \in L(\delta_1) \cup L(\delta_2)} \text{NEW-MAP} (l)$$
+$$w' = \sum_{l \in L(\delta_1) \cup L(\delta_2)} \text{EXTRA-WORK} (l)$$
+Then, the fork-join cost semantics are:
+$$\delta; e_1 || e_2 \Downarrow \delta'; (v_1, v_2); 1 + w_1 + w_2 + w'; 1 + \max(s_1, s_2)$$
 
@@
 Set in an OLD array involves copying the array, and has work $O(n)$.
-
-\section{Cost Dynamics}
-
-Arrays are represented by $(l, D)$ where $l$ is a label that uniquely identifies the array, and $D$ contains the values in the array.
-$D[i]$ is the $i^{\text{th}}$ value in the array.
-If the set method has been applied on an array, the array is OLD (represented by $-$), otherwise the array is NEW (represented by $+$).
-
-Array methods have different costs depending on whether the array is NEW or OLD.
-Let $C(l)$ be the number of get operations applied on array $A = (l, D)$ that had $O(1)$ work.
-We thread a store $\delta$ through each operation.
-$\delta$ maps array labels $l$ to $(+/-, c)$, where $c = C(l)$.
-
-Our cost dynamics are defined by the following judgement, where $\delta$ is the store, $f$ is the array function being applied, $\delta'$ is the new store, $v$ is the value returned by applying $f$ on $args$, $w$ is the work, and $s$ is the span.
-$$\delta; f \; args \Downarrow \delta'; v; w; s$$
-The cost dynamics for the methods on a functional array of size $n$ are:
-$$\frac{l = \text{new label} \;\;\; f(1) \Downarrow v_1; w_1; s_1; ...
-f(n) \Downarrow v_{n}; w_{n}; s_{n}}{\delta; \text{tabulate } f \; n \Downarrow \delta[l \mapsto (+, 0)]; (l, \bigcup_{i=1}^n [i \mapsto v_i]); 1+\sum w_i; 1+\max s_i}$$
-$$\frac{\delta[l \mapsto (+, c)] \;\;\; A = (l,D) \;\;\; l' = \text{new label}}{\delta; \text{set} \; A \; i \; v \Downarrow \delta[l \mapsto (-, c), l' \mapsto (+, 0)]; (l', D[i \mapsto v]); 1; 1} \text{  (set-new)}$$
-$$\frac{\delta[l \mapsto (-, c)] \;\;\; A = (l,D) \;\;\;  l' = \text{new label}}{\delta; \text{set} \; A \; i \; v \Downarrow \delta[l \mapsto (-, c), l' \mapsto (+, 0)]; (l', D[i \mapsto v]); n; 1} \text{  (set-old)}$$
-$$\frac{\delta[l \mapsto (+, c)] \;\;\; A = (l,D)}{\delta; \text{get} \; A \; i \Downarrow \delta[l \mapsto (+, c+1)]; D[i]; 1; 1} \text{  (get-new)}$$
-$$\frac{\delta[l \mapsto (-, c)] \;\;\; A = (l,D)}{\delta; \text{get} \; A \; i \Downarrow \delta; D[i]; \log{n}; 1} \text{  (get-old)}$$
-
-The fork-join cost semantics are the most interesting.
-We give the cost dynamic using 2 helper functions.
-Let $L(\delta)$ denote the set of labels in the store $\delta$.
-Let $f(n)$ be the cost of calling get on an OLD array and $g(n)$ be the cost of calling set on an OLD array.
-\begin{align*}
-\text{def NEW}&\text{-MAP: }L(\delta_1) \cup L(\delta_2) \to (+/-, \text{int}) = \\
-\text{Case } &\delta[l \mapsto (-,c)]: (-, c)  \\
-\text{Case } &\delta[l \mapsto (+,c)]:  \\
-&\text{Case } \delta_1[l \mapsto (+,c+x)], \delta_2[l \mapsto (+,c+y)]: (+, c+x+y) \\
-&\text{Case } \delta_1[l \mapsto (+,c+x)], \delta_2[l \mapsto (-,c+y)]: (-, c+y) \\
-&\text{Case } \delta_1[l \mapsto (-,c+x)], \delta_2[l \mapsto (+,c+y)]: (-, c+x) \\
-&\text{Case } \delta_1[l \mapsto (-,c+x)], \delta_2[l \mapsto (-,c+y)]: (-, c) \\
-\text{Else: } & \\
-&\text{Case } \delta_1[l \mapsto (s,c)]: (s,c) \\
-&\text{Case } \delta_2[l \mapsto (s,c)]: (s,c) \\
-\text{def EXT}&\text{RA-WORK: }L(\delta_1) \cup L(\delta_2) \to int = \\
-\text{Case } &\delta[l \mapsto (+,c)]:  \\
-&\text{Case } \delta_1[l \mapsto (+,c+x)], \delta_2[l \mapsto (+,c+y)]: 0 \\
-&\text{Case } \delta_1[l \mapsto (+,c+x)], \delta_2[l \mapsto (-,c+y)]: x f(n) \\
-&\text{Case } \delta_1[l \mapsto (-,c+x)], \delta_2[l \mapsto (+,c+y)]: y f(n) \\
-&\text{Case } \delta_1[l \mapsto (-,c+x)], \delta_2[l \mapsto (-,c+y)]: (x+y)f(n) + g(n) \\
-\text{Else: } & 0
-\end{align*}											
-$$\delta' = \bigcup_{l \in L(\delta_1) \cup L(\delta_2)} \text{NEW-MAP} (l)$$
-$$w' = \sum_{l \in L(\delta_1) \cup L(\delta_2)} \text{EXTRA-WORK} (l)$$
-Then, the fork-join cost semantics are:
-$$\delta; e_1 || e_2 \Downarrow \delta'; (v_1, v_2); 1 + w_1 + w_2 + w'; 1 + \max(s_1, s_2)$$
 

### 2016-03-11 Paper/paper.tex
Improve intro, add references, improve fork join dynamics

@@
 
-There has been a lot of great work on functional arrays, but previous approaches have some limitations.
-Many functional programming languages use monads to build arrays.
-However, monads do not allow accessing old versions of an array.
+There has been a lot of great work on functional arrays, but previous approaches have limitations.
 
-Another creative approach is to use compiler based reference counting.
-If the number of references to an array is provably one, the compiler allows the array to mutate, otherwise the compiler copies the array before writing to the array.
-However, this approach makes it difficult for programmers to reason about the time complexity of array operations because it depends on the compiler.
-Furthermore, if multiple variables reference the same array, the array is copied even if all writes are to the most recent version.
-This could make the time complexity of programs significantly higher than they need to be.
+In many functional programming languages, like Haskell, monads ~\cite{wadler-monads} can be used to enforce single threaded use of arrays.
+However, enforcing single threaded use obstructs persistence - it prevents us from accessing or modifying old versions of an array.
 
-A third approach is to use linear types to ensure that programmers can only access and mutate arrays in valid ways. This approach is very difficult to implement in existing programming languages.
+Another approach is to use reference counting ~\cite{hudak}.
+If the number of references to an array is one, the array can be updated in place, otherwise the array is copied and then updated.
+This approach makes it difficult for programmers to statically analyze the time complexity of array operations.
+Furthermore, even if there are multiple reference to an array the array might not need to be copied, so the unnecessary copying could make the implementation inefficient.
 
-O' Neill's functional arrays guarantee $O(1)$ access and insertions when used like an imperative array.
+A third approach is to use linear types ~\cite{wadler-lintypes} to ensure that programmers can only access and mutate arrays in valid ways.
+Linear types have not been used practically for a few decades since its inception, and would require a complete re-think of a language's type system to be useful.
+
+O' Neill's ~\cite{oneill} functional arrays guarantee $O(1)$ access and insertions when used like an imperative array.
 However, the arrays explicitly implement version trees which are inefficient in practice.
-Additionally, the reads and writes to the most recent version of the array might take $O(\log{n})$ work if the array is not used like an imperative array.
+Additionally, the reads and writes to the most recent version of the array might take $O(\log{n})$ work if the array is not used in a single threaded way.
 
-Chuang's approach supports $O(1)$ accesses and insertions to the most recent version of arrays, however accessing old versions of the array takes $O(n)$ work which is often impractical.
+Chuang's approach ~\cite{chuang} supports $O(1)$ accesses and insertions to the most recent version of arrays, however accessing old versions of the array takes $O(n)$ work which is often impractical.
 
@@
 Additionally, per-element locks add significant memory overhead.
-Separation logic could be used to parallelize divide and conquer algorithms on arrays however they would severely restrict the way functional arrays can be used.
+Separation logic ~\cite{reynolds} could be used to parallelize divide and conquer algorithms on arrays however they would severely restrict the way functional arrays can be used.
 
@@
 
-Our arrays support 3 methods: tabulate, get, and set. The dynamics are functional, in particular set returns a new array and leaves the original array intact.
+Our arrays support 3 functions: tabulate, get, and set. The dynamics are functional, in particular set returns a new array and leaves the original array intact.
 
@@
 \centering
-\includegraphics[scale=0.5]{leaf_interior_intro}
+\includegraphics[scale=0.45]{leaf_interior_intro}
 \nocaptionrule \caption{Example usage of functional arrays}
@@
 
-In figure \ref{fig:leaf_interior_intro}, A and C are \emph{interior} arrays and B and D are \emph{leaf} arrays.
+After all the operations performed in figure \ref{fig:leaf_interior_intro}, A and C are \emph{interior} arrays and B and D are \emph{leaf} arrays.
 Calling get and set on leaf arrays costs $O(1)$ but calling get and set on interior arrays is more expensive.
@@
 We give a wait-free concurrent implementation that uses careful synchronization.
-We prove linearizability of the implementation and use that to prove that arrays work correctly when used concurrently.
+We prove linearizability of the implementation and use that to prove that our arrays work correctly when used concurrently.
 Finally, we give a provable implementation for our cost dynamics.
@@
 In particular, one side of the fork might call get on $A$ while the other side calls set (see the center panel).
-The calls to get on the left branch would be $O(1)$ if and only if they execute before the call to set on the right branch.
+Calls to get on the left branch would be $O(1)$ if and only if they execute before the call to set on the right branch.
 
@@
 Our cost dynamics are defined by the following judgement, where $\delta$ is the store, $e$ is the expression, $\delta'$ is the new store, $v$ is the value returned by evaluating $e$, $w$ is the work, and $s$ is the span.
-$$\delta; f \; args \boldsymbol\Downarrow \delta'; v; w; s$$
+$$\delta; e \boldsymbol\Downarrow \delta'; v; w; s$$
 
@@
 $V[i]$ is the $i^{\text{th}}$ value in the array.
-If the set method has been applied on an array, the array is a leaf array (represented by $-$), otherwise the array is an interior array (represented by $+$).
+If the set function has been called on an array, the array is an interior array (represented by $-$), otherwise the array is a leaf array (represented by $+$).
 
-Array methods have different costs depending on whether the array is a leaf or interior array.
+Array functions have different costs depending on whether the array is a leaf or interior array.
 Let $C(l)$ be the number of get operations applied on array $A = (l, V)$ that had $O(1)$ work.
@@
 
-The cost dynamics for the methods on a functional array of size $n$ are:
+The cost dynamics for the functions on a functional array of size $n$ are:
 $$\frac{l = \text{new label} \;\;\; f(1) \Downarrow v_1; w_1; s_1; ...
@@
 $$\frac{A = (l,V) \;\;\; \delta[l \mapsto (+, c)] \;\;\; l' = \text{new label}}{\delta; \text{set} \; A \; i \; v \Downarrow \delta[l \mapsto (-, c), l' \mapsto (+, 0)]; (l', V[i \mapsto v]); 1; 1} \text{  (set-leaf)}$$
-$$\frac{A = (l,V) \;\;\; \delta[l \mapsto (-, c)] \;\;\;  l' = \text{new label}}{\delta; \text{set} \; A \; i \; v \Downarrow \delta[l \mapsto (-, c), l' \mapsto (+, 0)]; (l', V[i \mapsto v]); g(n); 1} \text{  (set-interior)}$$
+$$\frac{A = (l,V) \;\;\; \delta[l \mapsto (-, c)] \;\;\;  l' = \text{new label}}{\delta; \text{set} \; A \; i \; v \Downarrow \delta[l' \mapsto (+, 0)]; (l', V[i \mapsto v]); g(n); 1} \text{  (set-interior)}$$
 $$\frac{A = (l,V) \;\;\; \delta[l \mapsto (+, c)]}{\delta; \text{get} \; A \; i \Downarrow \delta[l \mapsto (+, c+1)]; V[i]; 1; 1} \text{  (get-leaf)}$$
@@
 The work for set is higher for interior arrays.
-Get returns the value at the desired index, if the array is a leaf then the operation costs $O(1)$ but we increment the counter of $O(1)$ gets in the store,
+Get returns the value at the desired index, if the array is a leaf then the operation costs $O(1)$ but we increment the counter of $O(1)$ gets in the store.
 
-The fork-join cost semantics are the most interesting. We give the cost dynamic using 2 helper functions. Let $L(\delta)$ denote the set of labels in the store $\delta$.
-Simplify the fork join cost dynamics
-\begin{align*}
-\text{def NEW}&\text{-MAP: }L(\delta_1) \cup L(\delta_2) \to (+/-, \text{int}) = \\
-\text{Case } &\delta[l \mapsto (-,c)]: (-, c)  \\
-\text{Case } &\delta[l \mapsto (+,c)]:  \\
-&\text{Case } \delta_1[l \mapsto (+,c+x)], \delta_2[l \mapsto (+,c+y)]: (+, c+x+y) \\
-&\text{Case } \delta_1[l \mapsto (+,c+x)], \delta_2[l \mapsto (-,c+y)]: (-, c+y) \\
-&\text{Case } \delta_1[l \mapsto (-,c+x)], \delta_2[l \mapsto (+,c+y)]: (-, c+x) \\
-&\text{Case } \delta_1[l \mapsto (-,c+x)], \delta_2[l \mapsto (-,c+y)]: (-, c) \\
-\text{Else: } & \\
-&\text{Case } \delta_1[l \mapsto (s,c)]: (s,c) \\
-&\text{Case } \delta_2[l \mapsto (s,c)]: (s,c) \\
-\text{def EXT}&\text{RA-WORK: }L(\delta_1) \cup L(\delta_2) \to int = \\
-\text{Case } &\delta[l \mapsto (+,c)]:  \\
-&\text{Case } \delta_1[l \mapsto (+,c+x)], \delta_2[l \mapsto (+,c+y)]: 0 \\
-&\text{Case } \delta_1[l \mapsto (+,c+x)], \delta_2[l \mapsto (-,c+y)]: x f(n) \\
-&\text{Case } \delta_1[l \mapsto (-,c+x)], \delta_2[l \mapsto (+,c+y)]: y f(n) \\
-&\text{Case } \delta_1[l \mapsto (-,c+x)], \delta_2[l \mapsto (-,c+y)]: (x+y)f(n) + g(n) \\
-\text{Else: } & 0
-\end{align*}											
-$$\delta' = \bigcup_{l \in L(\delta_1) \cup L(\delta_2)} \text{NEW-MAP} (l)$$
-$$w' = \sum_{l \in L(\delta_1) \cup L(\delta_2)} \text{EXTRA-WORK} (l)$$
+The fork-join cost semantics are the most interesting. Suppose we have an expression $e_L || e_R$ and
+\[ \delta, e_L \Downarrow \delta_L; v_L; w_L, s_L \]
+\[ \delta, e_R \Downarrow \delta_R; v_R; w_R, s_R \]
+
+We assume that newly created labels in $\delta_L$ and $\delta_R$ do not conflict.
+Consider an array $A = (l, V)$ with $\delta[l] = (s,c)$, $\delta_L[l] = (s_L, c + c_L)$, $\delta_R[l] = (s_R, c + c_R)$.
+When multiplying two signs or multiplying a sign with an integer, consider + to be 1 and - to be 0.
+
+Combine describes how to combine the values in the store on both sides of the fork-join.
+The array is a leaf array iff it is a leaf array on both sides of the fork-join.
+$O(1)$ gets on one side of the fork remain $O(1)$ iff there were no calls to set on the other side of the fork.
+\begin{equation*}
+  \begin{aligned}
+   combine&((s, c), (s_L, c + c_L), (s_R, c + c_R))= \\
+   &(s_L s_R, c + s_R c_L + s_L c_R)
+  \end{aligned}
+\end{equation*}
+Extrawork gives the additional cost incurred if an array was modified on either side of a fork-join.
+If the array was an interior array before the fork-join, then all functions incurred their maximal cost and there is no additional cost.
+Otherwise, $O(1)$ gets on one side of the fork become $O(f(n))$ iff the other side of the fork called set.
+If both sides of the fork called set, then one of the sets came first and is $O(1)$, the subsequent set is $O(g(n))$.
+\begin{equation*}
+  \begin{aligned}
+  extrawork&((s, c), (s_L, c + c_L), (s_R, c + c_R)) = \\
+  &s \cdot ( (\neg s_R) c_L f(n)  + (\neg s_L) c_R f(n) + (\neg s_R) (\neg s_L) g(n) )
+ \end{aligned}
+\end{equation*}
+
+\begin{equation*}
+  \begin{aligned}
+  \delta' = \; &(\delta_L / \delta) \; \cup \; (\delta_R / \delta) \; \cup \\
+               &\bigcup_{l \in L(\delta_L) \cap L(\delta_R)} [l \mapsto combine(\delta[l], \delta_L[l], \delta_R[l])]
+  \end{aligned}
+\end{equation*}
+$$w' = \sum_{l \in L(\delta_1) \cap L(\delta_2)} extrawork(\delta[l], \delta_L[l], \delta_R[l])$$
 Then, the fork-join cost semantics are:
@@
 Consider an object $A$.
-Suppose that the program invokes methods $M_1$, $M_2$, ..., $M_n$ on $A$.
+Suppose that the program invokes functions $M_1$, $M_2$, ..., $M_n$ on $A$.
 Consider arbitrary $i$, and suppose that the instructions in $M_i$ were executed at times $s_1$, $s_2$, ..., $s_m$.
@@
 We use an auxiliary data structure, \emph{PushArrays}, to store log entries.
-PushArrays are initially empty and support 3 methods.
+PushArrays are initially empty and support 3 functions.
 Suppose we have a PushArray $A$.
@@
 \begin{proof}
-The linearization point of $push$ is $A.size$ $+=$ $1$, and the linearization point of the other methods is their single instruction.
+The linearization point of $push$ is $A.size$ $+=$ $1$, and the linearization point of the other functions is their single instruction.
 In push, expanding the capacity of $A.data$ when it is full does not interfere with accessing $A.data$ because we copy values to the new array before pointing $A.data$ to the new array.
@@
 \begin{proof}
-This is a standard implicit parallelism type theorem (ref: PFPL) so we omit the details.
+This is a standard implicit parallelism type theorem, for example see Chapter 39 in~\cite{harper}, so we omit the details. 
 \end{proof}

### 2016-03-11 Paper/paper.tex
Add cost dynamics for regular functions

@@
 $L$ contains the usual arithmetic types, such as the natural numbers, and numerical operations such as sums and products.
+Lambda expressions and constants (including arrays) are terminal values in our language.
 Our dynamics are given by the judgement $e \Downarrow v$ which means that expression $e$ evaluates to $v$.
@@
 $$\frac{e_1 \Downarrow v_1 \;\;\; e_2 \Downarrow v_2}{e_1 || e_2 \Downarrow (v_1, v_2)} \text{ (fork-join)}$$
-$$\frac{f \Downarrow \lambda x . e \;\;\; y \Downarrow v}{f \; y \Downarrow [v/x]e} \text{ (function-app)}$$
+$$\frac{f \Downarrow \lambda x . e \;\;\; y \Downarrow v \;\;\; [v/x]e \Downarrow v'}{f \; y \Downarrow v'} \text{ (function-app)}$$
 
@@
 
-We give the cost dynamics for expressions in the language: (insert cost dynamics for standard functions later)
+We give the cost dynamics for arbitrary expressions in our language:
+
+$$\frac{}{\delta; c \Downarrow \delta; c; 1; 1} \text{ (constants)}$$
+
+$$\frac{\delta; e_1 \Downarrow \delta'; \lambda x .
+e; w_1 ; s_1 \;\;\; \delta'; e_2 \Downarrow \delta'';v; w_2 ; s_2 \;\;\; \delta''; [v/x]e \Downarrow \delta'''; v'; w_3 ; s_3}{\delta; e_1 \; e_2 \Downarrow \delta'''; v'; 1+w_1+w_2+w_3; 1+s_1+s_2+s_3}$$
+
+$$\frac{\delta; e_1 \Downarrow \delta'; true; w_1; s_1 \;\;\; \delta'; e_2 \Downarrow \delta''; v; w_2; s_2}{\delta; \text{if } e_1 \text{ then } e_2 \text{ else } e_3 \Downarrow \delta''; v; 1+w_1+w_2; 1+s_1+s_2} \text{ (if-true)}$$
+
+$$\frac{\delta; e_1 \Downarrow \delta'; false; w_1; s_1 \;\;\; \delta'; e_3 \Downarrow \delta''; v; w_2; s_2}{\delta; \text{if } e_1 \text{ then } e_2 \text{ else } e_3 \Downarrow \delta''; v; 1+w_1+w_2; 1+s_1+s_2} \text{ (if-false)}$$
 

### 2016-05-05 Paper/intro.tex
Syncing github with other repo

\section{Introduction}
\label{sec:intro}

Supporting sequences (arrays) with efficient (constant time) random
access updates has been a persistent problem in functional languages
(pun intended).  It is easy to implement such sequences with constant
time random access reads (using arrays stored as contiguous memory),
or to implement with logarithmic time reads and updates using balanced
trees, but getting both in constant time cannot be done without some
form of extension.
%~\cite{Pippenger97}.  
This means that algorithms
for many fundamental problems are a logarithmic factor slower in
functional languages than in imperative languages.  This includes
algorithms for basic problems such as generating a random permutation,
and for many important graph problems (e.g.,
shortest-unweighted-paths, connected components, biconnected
components, topological sort, and cycle detection).  Simple algorithms
for these problems take linear time in the imperative setting, but an
additional logarithmic factor in time in the functional setting, at
least without extensions.

A variety of approaches have been suggested to alleviate this problem.
Many of the approaches are based on the observation that if there is
only a single reference to an array, it can be safely updated in
place.  The most common such approach is to use
monads~\cite{Moggi89,Wadler95}.  Monads can force code to be single
threaded and can thread mutable arrays (or other mutable
state) implicitly so the only reference to them is the current one.
Haskell supplies the ST monad for this purpose.  King and Launchbery,
for example, use STMonads with an array to implement depth first (DFS)
search in linear work~\cite{KL95}.  Guzman and Hudak's single-threaded
lambda calculus also uses the idea of single threading the
computation~\cite{GH90}, and motivate the approach based on allowing
for fast array updates.  The problem with using these approaches is
that it is basically implementing a second imperative language within
a functional language.  Using monads means using a new syntax not just
for the operations themselves but for everything the operations are
embedded in.  Indeed King and Launchbery have to write completely
different code for DFS, even though it is only for efficiency and not
correctness (it is trivial to implement DFS in $O(m \log n)$ time
purely functionally).  Monads also do not allow keeping persistent
versions of structures, exactly because the state can be changed.
More importantly, however, it forces the program itself to be single
threaded, inhibiting parallelism.

A second approach is to use a type system that enforces the
single-threadedness of individual arrays rather than the program as a
whole.  This can be done with linear types~\cite{Girard87,Wadler90},
which can be used to ensure that references are not duplicated.  This
is more flexible than monads, since it allows the threading of the
data to be distinct from the threading of the program.  However, such
type systems are not available in most languages, and they can be hard
to reason about.

A third approach is to support fully functional arrays in a general
functional language, and to check in the program if they happen to be
used in a single threaded manner.  This can be done, for example, by
using reference counting~\cite{HB85,Hudak86}.  The idea is to keep
track of how many references there are to an array and update inplace
if there is only one.  Hudak describes techniques to statically
analyze code to determine that at certain points in the code a count
must be one, and therefore it is safe to replace an update with an
inplace update.  The problem with this approach is that the efficiency
of the code depends heavily on the effectiveness of the compiler and
the specific way the code is written, and it can therefore be hard for
the user to reason about efficiency.

The last approach is to fully support functional arrays, even with
sharing, but more efficiently than as a simple array or tree.  This is
sometimes referred to as version tree arrays~\cite{AHN88} or fully
persistent arrays~\cite{DSST89}.  The idea is to maintain a version
tree for an array (or more generally other data types), such that any
node of the tree can be updated to create a new leaf below it with the
modified value, without effecting the value of other nodes.  Dietz
showed that arrays can be maintained fully persistently using
$O(\log\log n)$ time per operation (read or write).  Holmstrom, Hughes
and others (see \cite{AHN88}) suggested storing a version tree by
keeping the full array at the root, and a each node represents an
update.  Chuang's approach ~\cite{chuang} supports $O(1)$ accesses and
insertions to the most recent version of arrays, however accessing old
versions of the array takes $O(n)$ work which is often impractical.
O'Neill and Warren~\cite{oneill} describe various improvements.
The problem with these approaches is that they can be very complicated,
and only efficient in certain cases.   Furthermore it can be hard to
reason about the performance.

None of the existing approaches support concurrent operations on
arrays.  O' Neill suggests (in passing) having a lock for each element
of the array. However, when many threads contend for the same element,
this would serialize accesses and make accesses not
$O(1)$. Additionally, per-element locks add significant memory
overhead.  
% Separation logic ~\cite{reynolds} could be used to
% parallelize divide and conquer algorithms on arrays however they would
% severely restrict the way functional arrays can be used.

\subsection*{Our Approach}

In this paper we present an approach for efficiently supporting
functional arrays.  It uses some ideas from the previous approaches
but unlike the previous approaches it supplies a well
defined cost semantics, and supports parallelism, and without language
extensions.  More specifically the approach has the
following important features.
\begin{enumerate}
\item It has fully functional value semantics (dynamic
  semantics)---i.e., when not considering costs, arrays act no
  differently than purely functional arrays.
\item It requires no changes in existing languages and no special
  types, syntactic extensions, etc.  Programs can be efficient---i.e.,
  constant time reads and writes---when using a completely standard
  functional programming style.
\item The approach supports nested parallelism---arrays can be passed
  to parallel calls and safely updated and read, again with a purely
  functional semantics.
\item We supply a well defined cost semantics, which can be used to
  formally analyze the cost of any program.  The semantics captures
  both sequential and parallel costs.
\item We describe a cost-bounded implementation which maps costs in the
  model to costs on either a sequential or parallel RAM (PRAM).
\item
Although accessing old versions of an array is more expensive than
accessing new versions (as defined in the cost semantics), reading old
versions never costs more than $O(\log n)$ work.
\end{enumerate}
We have implemented the approach and in this paper present some
performance numbers.

Although our value semantics are purely functional (and compositional)
our cost semantics are not---they require passing a store and modeling
costs based on the order the store is threaded.  The reader might
object to this non-functional nature of the cost semantics, but we
note that non-functional costs are also requires in call-by-need
(lazy) languages, which supply a functional value semantics but an
imperative cost semantics.  In fact from the value point of view,
call-by-need is no different from call-by-name, and it is only when
considering costs that they differ.  The cost semantics of
call-by-need~\cite{AMOFW95} is completely non-functional
(non-compositional).  Furthermore it is likely significantly more
complicated to reason about than the cost semantics we present in this
paper.

\begin{figure}[!ht]
\centering 
\includegraphics[scale=0.45]{leaf_interior_intro}
\nocaptionrule \caption{Example usage of functional arrays}
\label{fig:leaf_interior_intro}
\end{figure}

In this paper we consider arrays with three functions: \tabulate{},
\get{} (read), and \set{} (write).  As usual, \tabulate{}$(f,n)$
creates a new array of length $n$ by applying the function $f$ to
$[0, \ldots, n-1]$, \get$(A,i)$ returns the $i^{th}$ element of $A$, and
\set$(A,i,v)$ sets the $i^{th}$ element of $A$ to $v$ returning a
``new'' array (in the following discussion $n$ refers to the length of
an array).  Figure~\ref{fig:leaf_interior_intro} gives an example.  As
in previous work~\cite{AHN88} the history of an array after it is
created with \tabulate{} can be viewed as a version tree.  A version
tree has interior nodes and leaves.  In the figure, after all
functions are applied, arrays $A$ and $C$ are \emph{interior nodes},
and arrays $B$ and $D$ are \emph{leaves} of the version tree.  In our
cost semantics applying \get{} and \set{} to arrays at the leafs takes
constant work, but applying them to interior nodes is more expensive.
We define a cost semantics that identifies whether an array is a leaf
or interior node and charges for the cost of \get{} and \set{}
appropriately.  This requires threading a store that maps labels
(corresponding to each array) to indicate whether the array is a leaf
or interior.

Since we are also interested in parallelism, we include fork-join
parallelism in our language and semantics.    Our cost semantics then
models the cost of arrays when used in parallel.  This is non-trivial
since parallel tasks can read and update an array in parallel.
To deal with this, we compute the work of each forked task separately,
and then charge additional work at the join point if an array is used
in both forks.

To implement our functional arrays, we maintain at each leaf of the
version tree an array of the most recent values.  We also keep with
each location a change-log that keeps track of values at interior
nodes of the version tree.  Applying \get{} to a leaf only requires a
lookup in the array.  Applying \set{} to a leaf only requires updating
the array, and adding an element to the change log for the old
version.  Applying \get{} to an interior nodes requires a binary
search on the change log, which can be bounded by $O(\log n)$ time.
To ensure that change logs do not get too large, whenever the total
size across all change logs reaches $n$, the array is copied.   This
requires $O(n)$ time, but can be amortized against the updates.
Applying \set{} to an interior node requires creating a new array,
and copying values to it.   This requires $O(n)$ time.

We give a wait-free concurrent implementation that uses careful
synchronization. We prove linearizability of the implementation and
use that to prove that our arrays work correctly when used
concurrently. Finally, we give a provable implementation for our cost
dynamics.

### 2016-05-05 Paper/paper.tex
Add author
Syncing github with other repo

@@
 
+\newcommand{\func}[1]{{\sc #1}}
+\newcommand{\tabulate}{\func{tabulate}}
+\newcommand{\get}{\func{get}}
+\newcommand{\set}{\func{set}}
+\newcommand{\tget}{\text{\get}}
+\newcommand{\tset}{\text{\set}}
+\newcommand{\ttab}{\text{\tabulate}}
+
 \begin{document}
@@
            {blelloch@cs.cmu.edu}
-
+\authorinfo{Robert Harper}
+           {Carnegie Mellon University}
+           {rwh@cs.cmu.edu}
 \maketitle
@@
 
-\section{Introduction}
-
-Arrays are very important in functional programming languages because they allow work-efficient implementations of algorithms like depth-first search.
-Accessing old versions of arrays can be useful for efficient checkpointing, logging, and event handling.
-
-\subsection{Previous Approaches}
-
-There has been a lot of great work on functional arrays, but previous approaches have limitations.
-
-In many functional programming languages, like Haskell, monads ~\cite{wadler-monads} can be used to enforce single threaded use of arrays.
-However, enforcing single threaded use obstructs persistence - it prevents us from accessing or modifying old versions of an array.
-
-Another approach is to use reference counting ~\cite{hudak}.
-If the number of references to an array is one, the array can be updated in place, otherwise the array is copied and then updated.
-This approach makes it difficult for programmers to statically analyze the time complexity of array operations.
-Furthermore, even if there are multiple reference to an array the array might not need to be copied, so the unnecessary copying could make the implementation inefficient.
-
-A third approach is to use linear types ~\cite{wadler-lintypes} to ensure that programmers can only access and mutate arrays in valid ways.
-Linear types have not been used practically for a few decades since its inception, and would require a complete re-think of a language's type system to be useful.
-
-O' Neill's ~\cite{oneill} functional arrays guarantee $O(1)$ access and insertions when used like an imperative array.
-However, the arrays explicitly implement version trees which are inefficient in practice.
-Additionally, the reads and writes to the most recent version of the array might take $O(\log{n})$ work if the array is not used in a single threaded way.
-
-Chuang's approach ~\cite{chuang} supports $O(1)$ accesses and insertions to the most recent version of arrays, however accessing old versions of the array takes $O(n)$ work which is often impractical.
-
-None of the existing approaches support concurrent operations on arrays.
-O' Neill suggests having a lock for each element of the array.
-However, when many threads contend for the same element, this would serialize accesses and make accesses not $O(1)$.
-Additionally, per-element locks add significant memory overhead.
-Separation logic ~\cite{reynolds} could be used to parallelize divide and conquer algorithms on arrays however they would severely restrict the way functional arrays can be used.
-
-\subsection{Our Approach}
-
-Our arrays support 3 functions: tabulate, get, and set. The dynamics are functional, in particular set returns a new array and leaves the original array intact.
-
-\begin{figure}[!ht]
-\centering
-\includegraphics[scale=0.45]{leaf_interior_intro}
-\nocaptionrule \caption{Example usage of functional arrays}
-\label{fig:leaf_interior_intro}
-\end{figure}
-
-After all the operations performed in figure \ref{fig:leaf_interior_intro}, A and C are \emph{interior} arrays and B and D are \emph{leaf} arrays.
-Calling get and set on leaf arrays costs $O(1)$ but calling get and set on interior arrays is more expensive.
-In the cost dynamics, we thread a store that keeps track of whether an array is an interior or leaf array.
-
-We also give and prove fork join cost dynamics for our arrays.
-The dynamics for arrays used concurrently are non trivial because one thread might call get on a leaf array while another thread calls set.
-To deal with this, we compute the work of each fork separately, and then charge additional work at the join point if an array is used in both forks.
-
-In our implementation of a functional array, we keep an array of values (multiple functional arrays might point to the same array).
-For each index of the array, we also keep a history of values that were previously at that index.
-To access an element in an interior array, we need to search through the history of values.
-
-We give a wait-free concurrent implementation that uses careful synchronization.
-We prove linearizability of the implementation and use that to prove that our arrays work correctly when used concurrently.
-Finally, we give a provable implementation for our cost dynamics.
+\input{intro}
 
@@
 
-We use a standard applicative-order language defined as follows:
-$$e = x \; | \; c  \; | \; \lambda x.e \; | \; e_1 e_2 \; | \; e_1 || e_2 \; | \; \text{if } e_1 \text{ then } e_2 \text{ else } e_3$$
+We use a standard applicative-order source language defined as follows:
+$$e = x \; | \; c  \; | \; \lambda x.e \; 
+| \; e_1 e_2 \; 
+| \; (e_1, e_2) \; 
+| \; (e_1 || e_2) \; 
+| \; \text{if } e_1 \text{ then } e_2 \text{ else } e_3$$
 
-$L$ contains the usual arithmetic types, such as the natural numbers, and numerical operations such as sums and products.
-Lambda expressions and constants (including arrays) are terminal values in our language.
-Our dynamics are given by the judgement $e \Downarrow v$ which means that expression $e$ evaluates to $v$.
-Dynamics for function application and fork-join are given as examples.
-
+The constants $c$ contains the usual arithmetic types, such as the
+natural numbers and numerical functions.  The expression $(e_1||e_2)$
+indicates that the two expressions can run in parallel, returning a
+pair when both expressions are fully evaluated, and $(e_1,e_2)$
+generates a pair sequentially.  The terminal (fully evaluated) values
+are lambda expressions, as well as pairs and arrays of terminal
+values, and are denoted by the judgement val.   Our value semantics are given by the judgement:
+\[e \Downarrow v\]
+indicating that the expression $e$ evaluates to $v$.  The rules for
+function application and parallel pair are given as examples, and
+other rules are standard.
 $$\frac{e_1 \Downarrow v_1 \;\;\; e_2 \Downarrow v_2}{e_1 || e_2 \Downarrow (v_1, v_2)} \text{ (fork-join)}$$
@@
 
-The language also includes 3 functions for working with arrays: tabulate, get, and set. The dynamics for tabulate is given below:
-$$\frac{f(1) \Downarrow v_1, ..., f(n) \Downarrow v_n}{\text{tabulate} \; f \; n \Downarrow [1 \mapsto v_1, ..., n \mapsto v_n]}$$
+The language also includes three functions for working with arrays: \tabulate{}, \get{}, and \set{}. The dynamics for \tabulate{} is given below:
+$$\frac{f(1) \Downarrow v_1, ..., f(n) \Downarrow v_n}{\text{\tabulate{}} \; f \; n \Downarrow [1 \mapsto v_1, ..., n \mapsto v_n]}$$
 
-$get(A,i)$ returns the value at the $i$th index of $A$.
-$set(A,i,v)$ returns a new array where the value at index $i$ is $v$ and the value at all other indices is the same as in $A$.
+\get$(A,i)$ returns the value at the $i$th index of $A$.
+\set$(A,i,v)$ returns a new array where the value at index $i$ is $v$ and the value at all other indices is the same as in $A$.
 The evaluation of a simple program in our language is shown below:\\
 
-$(\lambda A. \; get(A, 2) + get(A,3)) set(tabulate(\lambda i . \; i, 5), 2, 10)$
+$(\lambda A. \; \tget(A, 2) + \tget(A,3)) \tset(\ttab(\lambda i . \; i, 5), 2, 10)$
 
-$(\lambda A. \; get(A, 2) + get(A,3)) set([1,2,3,4,5], 2, 10)$
+$(\lambda A. \; \tget(A, 2) + \tget(A,3)) \tset([1,2,3,4,5], 2, 10)$
 
-$(\lambda A. \; get(A, 2) + get(A,3)) [1,10,3,4,5]$
+$(\lambda A. \; \tget(A, 2) + \tget(A,3)) [1,10,3,4,5]$
 
-$get([1,10,3,4,5], 2) + get([1,10,3,4,5],3) $
+$\tget([1,10,3,4,5], 2) + \tget([1,10,3,4,5],3) $
 
@@
 The dynamics given in the previous section are functional.
-However, the costs of get and set depend on whether the array is a leaf array or an internal array.
+However, the costs of get and set depend on whether the array is a leaf or internal array.
 To deal with this, we thread a store through the cost dynamics to keep track of whether each array is a leaf or internal array.
@@
 
-$$\frac{\delta; e_1 \Downarrow \delta'; \lambda x .
-e; w_1 ; s_1 \;\;\; \delta'; e_2 \Downarrow \delta'';v; w_2 ; s_2 \;\;\; \delta''; [v/x]e \Downarrow \delta'''; v'; w_3 ; s_3}{\delta; e_1 \; e_2 \Downarrow \delta'''; v'; 1+w_1+w_2+w_3; 1+s_1+s_2+s_3}$$
+$$\frac{\begin{gathered}\delta; e_1 \Downarrow \delta'; \lambda x .
+e; w_1 ; s_1 \;\;\; \delta'; e_2 \Downarrow \delta'';v; w_2 ; s_2 \;\;\; \\ \delta''; [v/x]e \Downarrow \delta'''; v'; w_3 ; s_3\end{gathered}}{\delta; e_1 \; e_2 \Downarrow \delta'''; v'; 1+w_1+w_2+w_3; 1+s_1+s_2+s_3} \text{ (func-app)}$$
 
@@
 $V[i]$ is the $i^{\text{th}}$ value in the array.
-If the set function has been called on an array, the array is an interior array (represented by $-$), otherwise the array is a leaf array (represented by $+$).
+If \set{} has been called on an array, the array is an interior array (represented by $-$), otherwise the array is a leaf array (represented by $+$).
 
 Array functions have different costs depending on whether the array is a leaf or interior array.
-Let $C(l)$ be the number of get operations applied on array $A = (l, V)$ that had $O(1)$ work.
 We thread a store $\delta$ through each operation.
-$\delta$ maps array labels $l$ to $(+/-, c)$, where $c = C(l)$.
-Let $f(n)$ be the work of calling get on an interior array and $g(n)$ be the work of calling set on an interior array.
+$\delta$ maps array labels $l$ to $(+/-, c)$, where $c$ represents the number of cheap \get{}s on the array.
+$L(\delta)$ denotes the set of labels in the store $\delta$.
+Let $f(n)$ be the work of calling \get{} on an interior array of size $n$ and $g(n)$ be the work of calling \set{} on an interior array of size $n$.
 
 The cost dynamics for the functions on a functional array of size $n$ are:
-$$\frac{l = \text{new label} \;\;\; f(1) \Downarrow v_1; w_1; s_1; ...
-f(n) \Downarrow v_{n}; w_{n}; s_{n}}{\delta; \text{tabulate } f \; n \Downarrow \delta[l \mapsto (+, 0)]; (l, \bigcup_{i=1}^n [i \mapsto v_i]); 1+\sum w_i; 1+\max s_i}$$
-$$\frac{A = (l,V) \;\;\; \delta[l \mapsto (+, c)] \;\;\; l' = \text{new label}}{\delta; \text{set} \; A \; i \; v \Downarrow \delta[l \mapsto (-, c), l' \mapsto (+, 0)]; (l', V[i \mapsto v]); 1; 1} \text{  (set-leaf)}$$
-$$\frac{A = (l,V) \;\;\; \delta[l \mapsto (-, c)] \;\;\;  l' = \text{new label}}{\delta; \text{set} \; A \; i \; v \Downarrow \delta[l' \mapsto (+, 0)]; (l', V[i \mapsto v]); g(n); 1} \text{  (set-interior)}$$
-$$\frac{A = (l,V) \;\;\; \delta[l \mapsto (+, c)]}{\delta; \text{get} \; A \; i \Downarrow \delta[l \mapsto (+, c+1)]; V[i]; 1; 1} \text{  (get-leaf)}$$
-$$\frac{A = (l,V) \;\;\; \delta[l \mapsto (-, c)]}{\delta; \text{get} \; A \; i \Downarrow \delta; V[i]; f(n); 1} \text{  (get-interior)}$$
+$$\frac{A = (l,V) \;\;\; i \text{ val} \;\;\; v \text{ val} \;\;\; \delta[l \mapsto (+, c)] \;\;\; l' \not\in L(\delta)}{\delta; \tset \; A \; i \; v \Downarrow \delta[l \mapsto (-, c), l' \mapsto (+, 0)]; (l', V[i \mapsto v]); 1; 1} \text{  (set-leaf)}$$
 
-Set creates a new label and array and modifies the store to indicate that the newly created array is a leaf array, and the array set was called on has become an interior array.
-The work for set is higher for interior arrays.
-Get returns the value at the desired index, if the array is a leaf then the operation costs $O(1)$ but we increment the counter of $O(1)$ gets in the store.
+$$\frac{A = (l,V) \;\;\; i \text{ val} \;\;\; v \text{ val} \;\;\; \delta[l \mapsto (-, c)] \;\;\;  l' \not\in L(\delta)}{\delta; \tset \; A \; i \; v \Downarrow \delta[l' \mapsto (+, 0)]; (l', V[i \mapsto v]); g(n); 1} \text{  (set-interior)}$$
+
+$$\frac{\begin{gathered}\delta; e_1 \Downarrow \delta_1; A; w_1; s_1 \;\;\; \delta_1; e_2 \Downarrow \delta_2; i; w_2; s_2 \;\;\; \\ \delta_2; e_3 \Downarrow \delta_3; v; w_3; s_3 \;\;\; \delta_3; \tset \; A \; i \; v \Downarrow \delta'; A'; w'; s' \end{gathered}}{\delta; \tset \; e_1 \; e_2 \; e_3 \Downarrow \delta'; A'; w_1+w_2+w_3+w'; s_1+s_2+s_3+s'} \text{ (set-eval)}$$
+
+$$\frac{A = (l,V) \;\;\; i \text{ val} \;\;\; \delta[l \mapsto (+, c)]}{\delta; \text{get} \; A \; i \Downarrow \delta[l \mapsto (+, c+1)]; V[i]; 1; 1} \text{  (get-leaf)}$$
+
+$$\frac{A = (l,V) \;\;\; i \text{ val} \;\;\; \delta[l \mapsto (-, c)]}{\delta; \text{get} \; A \; i \Downarrow \delta; V[i]; f(n); 1} \text{  (get-interior)}$$
+
+$$\frac{\begin{gathered}\delta; e_1 \Downarrow \delta_1; A; w_1; s_1 \;\;\; \delta_1; e_2 \Downarrow \delta_2; i; w_2; s_2 \;\;\; \\ \delta_2; \tget \; A \; i \Downarrow \delta'; v'; w'; s' \end{gathered}}{\delta; \tget \; A \; i \Downarrow \delta'; v'; w_1 + w_2 + w'; s_1 + s_2 + s'}\text{ (get-eval)}$$
+
+$$\frac{\begin{gathered}\delta, e_L \Downarrow \delta_L; v_L; w_L, s_L \;\;\; \delta, e_R \Downarrow \delta_R; v_R; w_R, s_R \;\;\; \\ (L(\delta_L) \setminus L(\delta)) \cap (L(\delta_R) \setminus L(\delta)) = \emptyset\end{gathered}}{\delta; (e_L || e_R) \Downarrow \delta'; (v_L || v_R); 1 + w_L + w_R + w'; \max(s_L, s_R)} \text{  (fork-join)}$$
+
+\set{} creates a new label and array and modifies the store to indicate that the newly created array is a leaf array, and the array \set{} was called on has become an interior array.
+The work for \set{} is higher for interior arrays.
+\get{} returns the value at the desired index, if the array is a leaf then the operation costs 1 unit of work but we increment the counter of cheap \get{}s in the store.
 
@@
 
-We assume that newly created labels in $\delta_L$ and $\delta_R$ do not conflict.
+Further, suppose that $(\delta_L \setminus \delta) \cap (\delta_R \setminus \delta) = \emptyset$ (the new labels produced on both sides of the fork-join do not conflict).
 Consider an array $A = (l, V)$ with $\delta[l] = (s,c)$, $\delta_L[l] = (s_L, c + c_L)$, $\delta_R[l] = (s_R, c + c_R)$.
@@
 
-Combine describes how to combine the values in the store on both sides of the fork-join.
+\func{Combine} describes how to combine the values in the store on both sides of the fork-join.
 The array is a leaf array iff it is a leaf array on both sides of the fork-join.
-$O(1)$ gets on one side of the fork remain $O(1)$ iff there were no calls to set on the other side of the fork.
+Cheap \get{}s on one side of the fork remain cheap iff there were no calls to \set{} on the other side of the fork.
 \begin{equation*}
   \begin{aligned}
-   combine&((s, c), (s_L, c + c_L), (s_R, c + c_R))= \\
+   \text{\func{combine}}&((s, c), (s_L, c + c_L), (s_R, c + c_R))= \\
    &(s_L s_R, c + s_R c_L + s_L c_R)
@@
 \end{equation*}
-Extrawork gives the additional cost incurred if an array was modified on either side of a fork-join.
+$$\delta' = \; (\delta_L / \delta) \; \cup \; (\delta_R / \delta) \; \cup \bigcup_{l \in L(\delta)} [l \mapsto \text{\func{combine}}(\delta[l], \delta_L[l], \delta_R[l])]$$
+\func{Extrawork} gives the additional cost incurred if an array of size $n$ was modified on either side of a fork-join.
 If the array was an interior array before the fork-join, then all functions incurred their maximal cost and there is no additional cost.
-Otherwise, $O(1)$ gets on one side of the fork become $O(f(n))$ iff the other side of the fork called set.
-If both sides of the fork called set, then one of the sets came first and is $O(1)$, the subsequent set is $O(g(n))$.
+Otherwise, cheap \get{}s on one side of the fork are charged $f(n)$ work iff the other side of the fork called \set{}.
+If both sides of the fork called \set{}, then one of the \set{}s came first and has work $1$, the subsequent \set{} has work $g(n)$.
 \begin{equation*}
   \begin{aligned}
-  extrawork&((s, c), (s_L, c + c_L), (s_R, c + c_R)) = \\
-  &s \cdot ( (\neg s_R) c_L f(n)  + (\neg s_L) c_R f(n) + (\neg s_R) (\neg s_L) g(n) )
+  \text{\func{extra}}&\text{\func{work}}(l, (s, c), (s_L, c + c_L), (s_R, c + c_R)) = \\
+  & ((\neg s_R) c_L  + (\neg s_L) c_R)(f(size(l))-1) + \\
+  &s(\neg s_R) (\neg s_L) (g(size(l))-1)
  \end{aligned}
 \end{equation*}
+$$w' = \sum_{l \in L(\delta), \delta[l \mapsto (+, c)]} \text{\func{extrawork}}(l, \delta[l], \delta_L[l], \delta_R[l])$$
+Then, the fork-join cost semantics are:
+$$\delta; (e_L || e_R) \Downarrow \delta'; (v_L || v_R); 1 + w_L + w_R + w'; \max(s_L, s_R)$$
 
-\begin{equation*}
-  \begin{aligned}
-  \delta' = \; &(\delta_L / \delta) \; \cup \; (\delta_R / \delta) \; \cup \\
-               &\bigcup_{l \in L(\delta_L) \cap L(\delta_R)} [l \mapsto combine(\delta[l], \delta_L[l], \delta_R[l])]
-  \end{aligned}
-\end{equation*}
-$$w' = \sum_{l \in L(\delta_1) \cap L(\delta_2)} extrawork(\delta[l], \delta_L[l], \delta_R[l])$$
-Then, the fork-join cost semantics are:
-$$\delta; e_1 || e_2 \Downarrow \delta'; (v_1, v_2); 1 + w_1 + w_2 + w'; 1 + \max(s_1, s_2)$$
+\tabulate{} involves multiple function calls in parallel, so its cost semantics use \func{combine} and \func{extrawork} multiple times.
+We omit discussion of \tabulate{} here and in the rest of the paper because it involves messy details that are not relevant to the goals of this paper.
+
+\section{Analysis}
+
+\subsection{Interleaved Structural Dynamics}
+
+To show that our implementation is correct and to validate our cost dynamics, we introduce interleaved structural dynamics for the source language.
+The structural dynamics describe the possible steps a program in our source language can take and are given by the following judgement, where $S$ is the store, $e$ is the expression, $S'$ is the resulting store, $e'$ is the resulting expression after a step has been taken, and $w$ is the work.
+
+$$S, e \to S', e', w$$
+
+For now we let the span of each step be 1 (we account for the spans later on).
+
+As before, arrays are represented by $(l, V)$ where $l$ is a unique label and $V$ contains the values in the array.
+However, the store $S$ maps labels only to $+$ (leaf) or $-$ (interior).
+$L(S)$ denotes the set of labels in the store $S$.
+\func{size}$(l)$ denotes the size of the array referenced by label $l$.
+The judgement \func{val} describes terminal values.
+The structural dynamics are given as follows:
+$$\frac{e_2 \; \text{val}}{S, (\lambda x.e)e_2 \to S, [e_2/x]e, 1} \text{ (app-sub)}$$
+$$\frac{S, e_2 \to S', e_2', w}{S, (\lambda x.e) e_2 \to S', (\lambda x.e) e_2', w} \text{ (app-arg)}$$
+$$\frac{S, e_1 \to S', e_1', w}{S, e_1 e_2 \to S', e_1' e_2, w} \text{ (app-func)}$$
+$$\frac{}{S, \text{if } true \text{ then } e_2 \text{ else } e_3 \to S, e_2, 1} \text{ (if-true)}$$
+$$\frac{}{S, \text{if } false \text{ then } e_2 \text{ else } e_3 \to S, e_3, 1}  \text{ (if-false)}$$
+$$\frac{S, e_1 \to S', e_1', w}{S, \text{if } e_1 \text{ then } e_2 \text{ else } e_3 \to S', \text{if } e_1' \text{ then } e_2 \text{ else } e_3, w} \text{ (if-cond)}$$
+$$\frac{A = (l, V) \;\;\; S[l] = + \;\;\; i \text{ val}}{S, get(A, i) \to S, V[i], 1} \text{ (get-leaf)}$$
+$$\frac{A = (l, V) \;\;\; S[l] = -  \;\;\; i \text{ val}}{S, get(A, i) \to S, V[i], f(size(l))} \text{ (get-interior)}$$
+$$\frac{A = (l, V) \;\;\; S, e_2 \to S', e_2', w}{S, get(A, e_2) \to S', get(A, e_2'), w} \text{ (get-index)}$$
+$$\frac{S, e_1 \to S', e_1', w}{S, get(e_1, e_2) \to S', get(e_1', e_2), w} \text{ (get-array)}$$
+$$\frac{A = (l, V) \;\;\; S[l] = + \;\;\;  l' \not\in L(S) \;\;\; i \text{ val} \;\;\; v \text{ val}}{S, set(A, i, v) \to S[l \mapsto -], (l', V[i \mapsto v]), 1} \text{ (set-leaf)}$$
+$$\frac{A = (l, V) \;\;\; S[l] = - \;\;\;  l' \not\in L(S) \;\;\; i \text{ val} \;\;\; v \text{ val}}{S, set(A, i, v) \to S, (l', V[i \mapsto v]), g(size(l))} \text{ (set-interior)}$$
+$$\frac{A = (l, V) \;\;\; i \text{ val} \;\;\; S, e_3 \to S', e_3', w}{S, set(A, i, e_3) \to S', set(A,i,e_3'), w} \text{ (set-val)}$$
+$$\frac{A = (l, V) \;\;\; S, e_2 \to S', e_2', w}{S, set(A, e_2, e_3) \to S', set(A,e_2',e_3), w} \text{ (set-index)}$$
+$$\frac{e_1 \to S', e_1', w}{S, set(e_1, e_2, e_3) \to S', set(e_1',e_2,e_3), w} \text{ (set-array)}$$
+
+The structural dynamics for fork-join is non-deterministic since either side of the fork can take a step.
+$$\frac{S, e_1 \to S', e_1', w}{S, e_1 || e_2 \to S', e_1' || e_2, w} \text{ (fork-left)}$$
+$$\frac{S, e_2 \to S', e_2', w}{S, e_1 || e_2 \to S', e_1 || e_2', w} \text{ (fork-right)}$$
+
+The structural dynamics for tabulate is also non-deterministic.
+
+\begin{definition}
+A \emph{transition sequence} $T$ is a sequence of states $[(S_0, e_0), ..., (S_n, e_n)]$ and costs $[w_1, ..., w_n]$ s.t.
+for all $0 \leq i < n$, $S_i, e_i \to S_{i+1}, e_{i+1}, w_{i+1}$ .
+We say that $T$ takes $S_0, e_0$ to $S_n, e_n$ and has length $n$ and denote this by $S_0, e_0 \to^n S_n, e_n$.
+\end{definition}
+
+\begin{definition}
+We say that $T$ is \emph{maximal} if there does not exist $S', e', w'$ s.t. $S_n, e_n \to S', e', w'$.
+\end{definition}
+
+\begin{definition}
+A \emph{transition subsequence} $T_{i, j}$ of $T$ is given by the sequence of states $[(S_i, e_i), ..., (S_j, e_j)]$ and costs $[w_{i+1}, ..., w_j]$. Note that $T = T_{0,n}$.
+\end{definition}
+
+\begin{definition}
+The \emph{work of a transition sequence} $T$ is given by:
+$$W(T) = \sum_{i=1}^n w_i$$
+Different transition sequences starting and ending at the same states may have different work.
+\end{definition}
+
+\subsection{Sequential Cost Analysis}
+
+\begin{definition}
+Consider a transition sequence $T$.
+We say that $S_i, e_i \to S_{i+1}, e_{i+1}, w_{i+1}$ is a get on $l$ if the step involves evaluating the get function on array $(l, V)$.
+We say it is a \emph{cheap get} on $l$ if $w_{i+1} = 1$.
+The number of cheap gets on $l$ in $T$ is denoted by $SC_l(T)$.
+\end{definition}
+
+\begin{definition}
+Unlike the store in the evaluational dynamics, the store in the structural dynamics only stores whether each array is $+$ (leaf) or $-$ (interior).
+$sign(\delta)$ takes a store that maps $l \mapsto (s,c)$ and returns a store which maps $l \mapsto s$.
+\end{definition}
+
+\begin{definition}
+Suppose that $\delta, e \Downarrow \delta', e', w', s'$.
+We define the number of cheap gets in going from $\delta$ to $\delta'$ in the evaluational dynamics as follows.
+Suppose $l \mapsto (s', c') \in \delta'$.
+If $l \mapsto (s,c) \in \delta$ then $EC_l(\delta, \delta') = c' - c$ and if $l \not\in \delta$ then $EC_l(\delta, \delta') = c'$.
+If $l \not\in \delta'$ then $EC_l(\delta, \delta') = 0$.
+\end{definition}
+
+\begin{definition}
+A \emph{relabeling} of $L_1$ is a bijective function $R$ from label sets $L_1 \to L_2$.
+$R$ can be used to relabel stores and expressions.
+$R(\delta) = \{ R(l) \mapsto e \; | \; l \mapsto e \in \delta \wedge l \in L_1  \} \cup \{ l \mapsto e \; | \; l \mapsto e \in \delta \wedge l \not\in L_1  \}$.
+In other words, $R$ relabels some of the labels in $\delta$.
+Similarly, $R(e)$ returns an expression $e'$ where each occurrence of a label $l$ in $e$ is substituted by $R(l)$.
+\end{definition}
+
+\begin{lemma}
+\label{relabeling_lemma}
+Suppose that there exists a derivation of length $m$ that $\delta, e \Downarrow \delta', e', w, s$.
+Let $R$ be an arbitrary labeling.
+Then there exists a derivation of length $m$ that $R(\delta), R(e) \Downarrow R(\delta'), R(e'), w, s$.
+\end{lemma}
+
+\begin{proof}
+By applying the relabeling to each step of the derivation, and noting that all rules in the evaluational dynamics hold under relabelings.
+\end{proof}
+
+\begin{lemma}
+\label{stateless_lemma}
+If $S, e \to S', e', w$ and $U$ is a store s.t. for all $l \in e$, $l \in U$, then there exists $U', w'$ s.t. $U, e \to U', e', w'$
+\end{lemma}
+
+\begin{proof}
+By casing on each rule in the structural dynamics, and noting that transitions of expressions never depend on the values in the store.
+\end{proof}
+
+\begin{theorem}
+\label{main_work_theorem}
+Suppose that 
+$$\delta, e_0 \Downarrow \delta_E, e_E, w_E, s_E$$
+
+and consider arbitrary maximal transition sequence T taking 
+$$S_0 = sign(\delta), e_0 \to^n S_n, e_n$$
+å
+The following hold:
+\begin{enumerate}
+\item For some relabeling $R$ of $L(\delta_E) \setminus L(\delta)$, $S_n = sign(R(\delta_E))$ and $e_n = R(e_E)$.
+\item For all $l \in \delta_E$, $EC_l(\delta, \delta_E) \leq SC_{R(l)}(T)$.
+\item The sum of work and cheap gets is conserved: \begin{align*}
+w_E + &\sum_{l \in L(\delta_E)} EC_l(\delta, \delta_E) (f(size(l))-1) =\\
+ &W(T) + \sum_{l \in L(\delta_E)} SC_{R(l)}(T) (f(size(R(l))) - 1)
+ \end{align*}
+\end{enumerate}
+\end{theorem}
+
+\begin{proof}
+By induction on the length of the shortest derivation in the evaluation dynamics.
+We prove the theorem for 2 of the rules in the evaluation dynamics: get-leaf and fork-join.
+The other cases follow from a similar line of reasoning.
+\\
+
+\noindent \textbf{Case get-leaf}: Suppose $e_0 = get(e_L, e_R)$ and
+$$\delta, e_L \Downarrow \delta_L, e_L', w_L$$
+$$\delta_L, e_R \Downarrow \delta_R, e_R', w_R$$
+$$\delta_R, get(e_L', e_R') \Downarrow \delta_E, e_E, 1$$
+which gives us
+$$\delta, get(e_L, e_R) \Downarrow \delta_E, e_E, w_E$$
+with $w_E = w_L + w_R + 1$.
+
+Note that the structural dynamics first step the left argument to get, then the right argument, and finally evaluate the get.
+So there exists $m$ such that $T_{0,m}$ involves stepping the left argument, $T_{m,n-1}$ involves stepping the right argument, and $T_{n-1,n}$ involves evaluating the get.
+
+\textbf{Part 1}: Suppose that in $T_{0,m}$, for $0 \leq i \leq m$, $e_i = get(e_{L,i}, e_R)$.
+Consider a new projected transition sequence $T_{0,m}'$ with states $[(S_0, e_{L,0}), ..., (S_m, e_{L,m})]$ and costs the same as $T_{0,m}$: $[w_1, ..., w_m]$.
+It is easy to verify that $T_{0,m}'$ is a valid transition sequence starting at $(S_0, e_{L,0})$ and is maximal.
+
+$S_0 = sign(\delta)$ and $e_{L,0} = e_L$ so we can apply the IH to $T_{0,m}'$.
+For some relabeling $R$ of labels in $L(\delta_L) / \delta$, $s_m = sign(R(\delta_L))$ and $e_{L,m} = R(e_L')$ $\Rightarrow$ $e_m' = (e_{L,m}, e_R) = (R(e_L'), e_R)$.
+WLOG suppose that $\delta_L$ was labeled such that the above hold (we can do this because of lemma \ref{relabeling_lemma}).
+Note that in subsequent parts of the proof, we omit creating the projected transition sequence and apply the IH directly to $T_{0,m}$ when needed.
+
+Using similar logic on $T_{m,n-1}$ we get that for some relabeling of $\delta_R / \delta_L$, applying that relabeling to $\delta_R$ and $(e_L', e_R')$ gives us $e_{n-1} = (e_L', e_R')$ and $S_{n-1} = sign(\delta_R)$.
+WLOG suppose that $\delta_R$ was labeled so that the above holds.
+
+By combining the relabelings of $\delta_L / \delta$ and $\delta_R / \delta_L$ and applying that to $(e_L', e_R')$ and $\delta_R$, we get that $e_{n-1} = get(e_L', e_R')$ and $S_{n-1} = sign(\delta_R)$.
+Note that get does not change the sign of any label in the store in either the structural or the evaluation dynamics, so $sign(\delta_E) = sign(\delta_R) = S_{n-1} = S_n$.
+Further, the structural and evaluational dynamics apply get in the same way, so $e_n = e_E$.
+Therefore the first part of this theorem holds.
+WLOG suppose that the stores and expressions were relabeled s.t.
+the first part holds.
+
+\textbf{Part 2}: From lemma \ref{relabeling_lemma}, the length of the shortest derivation for the relabeled evaluational dynamics does not change, so we can apply the inductive hypothesis.
+Applying the IH to $T_{0,m}$, for all $l \in \delta_L$, $EC_l(\delta, \delta_L) \leq SC_l(T_{0,m}') = SC_l(T_{0,m})$.
+If $l \in \delta_R$ but $l \not\in \delta_L$ then $EC_l(\delta, \delta_L) = SC_l(T_{0,m}) = 0$ (intuitively the label did not exist and so did not have any cheap gets), so in particular, for all $l \in \delta_R$, $EC_l(\delta, \delta_L) \leq SC_l(T_{0,m})$.
+Applying the IH to $T_{m,n-1}$ we get that for all $l \in \delta_R$, $EC_l(\delta_L, \delta_R) \leq SC(T_{m,n-1})$.
+
+We note that $EC$ and $SC$ are additive, that is,
+$$EC_l(\delta, \delta_R) = EC_l(\delta, \delta_L) + EC_l(\delta_L, \delta_R)$$
+$$SC_l(T_{0,n-1}) = SC_l(T_{0,m}) + SC_l(T_{m,n-1})$$
+This implies that for all $l \in \delta_R$, $EC_l(\delta, \delta_R) \leq SC_l(T_{0,n-1})$.
+
+Now, let $e_L' = (l',V)$ with $l' \in \delta_R$.
+Then, $EC_{l'}(\delta, \delta_E) = EC_{l'}(\delta, \delta_R) + 1$ and $SC_{l'}(T) = SC_{l'}(T_{0,n-1}) + 1$, so $EC_{l'}(\delta, \delta_E) \leq SC_{l'}(T)$.
+If $l \neq l'$ then $EC_{l}(\delta, \delta_E) = EC_{l}(\delta, \delta_R)$ and $SC_{l}(T) = SC_{l}(T_{0,n-1})$, so $EC_{l}(\delta, \delta_E) \leq SC_{l}(T)$.
+So the second part of the theorem holds.
+
+\textbf{Part 3}: The third part of the theorem follows by similarly applying the IH on $T_{0,m}$ and $T_{m,n-1}$, and noting that $W(T) = W(T_{0,m}) + W(T_{m,n-1}) + 1$.
+\\
+
+\noindent \textbf{Case fork-join}: Suppose $e_0 = (e_L || e_R)$ and
+$$\delta, e_L \Downarrow \delta_L, v_L, w_L$$
+$$\delta, e_R \Downarrow \delta_R, v_R, w_R$$
+which gives us
+$$\delta, (e_L || e_R) \Downarrow \delta', (v_1 || v_2), 1 + w_L + w_R + w'$$
+with $w'$ and $\delta'$ defined in terms of \func{extrawork} and \func{combine} as given in the evaluational dynamics.
+
+In T, for all $0 \leq i \leq n$ let $e_i = (e^L_i || e^R_i)$.
+Let $I_L$ be the sequence of all $0 \leq i \leq n-1$ where transitioning from $e_i$ to $e_{i+1}$ involved stepping the left argument.
+Let $m$ be the length of $I_L$.
+Similarly define $I_R$ for the right argument, and let $k$ be the length of $I_R$.
+
+Let $A$ be the sequence of $e^L$s corresponding the indices in $I_L$, and add $e^L_n$ at the end of $A$.
+Similarly, let $B$ be the sequence of $e^R$s corresponding the indices in $I_R$, adding $e^R_n$ at the end.
+
+Let $S^L_0 = S_0 = sign(\delta)$.
+By inductively applying lemma \ref{stateless_lemma}, there exists $S^L_1, ...
+S^L_m$ s.t.
+$T_L = [(S^L_0, A_0), ..., (S^L_m, A_m)]$ is a transition sequence.
+$T_L$ is maximal, because if $A_m$ in $T_L$ can take a step then $e^L_n$ in $T$ can take the corresponding step.
+Similarly, let  $S^R_0 = S_0 = sign(\delta)$.
+Then, there exists $S^R_1, ...
+S^R_k$ s.t.
+$T_R = [(S^R_0, B_0), ..., (S^R_k, B_k)]$ is a maximal transition sequence.
+
+Applying the IH on $T_L$, we get a relabeling $R_L$ on $L(\delta_L) \setminus L(\delta)$ s.t.
+$R_L(v_L) = A_m = e^L_n$ and $sign(R_L(\delta_L)) = S^L_m$.
+Similarly, we get a relabeling $R_R$ on $L(\delta_R) \setminus L(\delta)$ s.t.
+$R_R(v_R) = B_k = e^R_n$ and $sign(R_R(\delta_R)) = S^R_k$.
+
+By inducting on the product $mk$, we can show that the new labels produced in $T_L$ and $T_R$ do not overlap, that is: $(L(S^L_m) \setminus L(S_0)) \cap (L(S^R_k) \setminus L(S_0)) = \emptyset$.
+(Note to self: Should I sketch out this induction?).
+Compose the relabelings to get relabeling R.
+
+Applying lemma $\ref{relabeling_lemma}$ and the fork join rule, we get,
+$$\delta, R(e_L) \Downarrow R(\delta_L), R(v_L), w_L$$
+$$\delta, R(e_R) \Downarrow R(\delta_R), R(v_R), w_R$$
+which gives us
+$$\delta, R((e_L || e_R)) \Downarrow R(\delta'), R((v_1 || v_2)), 1 + w_L + w_R + w'$$
+
+with $R((v_1||v_2)) = (R(v_1) || R(v_2)) = (e_n^L || e_n^R)$.
+To reduce clutter, WLOG suppose that the stores and expressions in the evaluational dynamics were labeled as such.
+By lemma \ref{relabeling_lemma} we can apply the IH on the relabeled evaluational dynamics since the length of the shortest derivation is invariant under relabelings.
+
+\begin{lemma}
+\label{label_projection_lemma}
+$l \in S_n$ if and only if either $l \in S^L_m$ or $l \in S^R_k$
+\end{lemma}
+
+\begin{proof}
+If $l \in S_0$ the claim holds trivially. Otherwise,
+
+$(\Rightarrow)$ By considering the step in $T$ that first introduced $l$, and examining the corresponding step in either $T_L$ or $T_R$.
+
+$(\Leftarrow)$ By considering the step in $T_L$ (or $T_R$) that first introduced $l$ and examining the corresponding step in $T$.
+\end{proof}
+
+\begin{lemma}
+\label{sign_projection_lemma}
+$S_n[l] = -$ if and only if either $l \in S^L_m$ and $S^L_m[l] = -$ or $l \in S^R_k$ and $S^R_k[l] = -$
+\end{lemma}
+
+\begin{proof}
+Similar to lemma \ref{label_projection_lemma}.
+\end{proof}
+
+\textbf{Part 1}: We claim that the set of labels in $\delta'$ and $S_n$ are the same.
+From the evaluational dynamics, $L(\delta') = L(\delta_L) \cup L(\delta_R)$.
+Since $sign(\delta_L) = S^L_m$ and $sign(\delta_R) = S^R_k$, $L(\delta_L) = L(S^L_m)$ and $L(\delta_R) = L(S^R_k)$.
+From lemma $\ref{label_projection_lemma}$, $L(S_n) = L(S^L_m) \cup L(S^R_k) = L(\delta')$ which proves the claim.
+
+To show that $sign(\delta') = S_n$ we case on arbitrary $l \in L(\delta')$ and show that $sign(\delta')[l] = S_n[l]$:
+
+\textbf{Case $l \in L(\delta)$}:
+
+If $\delta[l] = (-,\_)$ then $S_0[l] = -$. The evaluational and structural dynamics do not change $-$ to $+$ so $\delta'[l] = (-,\_)$ and $S_n[l] = -$ so $sign(\delta')[l] = S_n[l]$.
+
+Suppose $\delta[l] = (+, \textunderscore)$, in which case $S_0[l] = +$.
+If $\delta'[l] = (+,\_)$, then from the way combine works, $\delta_L[l] = (+, \_)$ and $\delta_R[l] = (+, \_)$.
+From the IH $S^L_m[l] = +$ and $S^R_k[l] = +$.
+From lemma \ref{sign_projection_lemma} $S_n[l] = + = sign(\delta')[l]$.
+
+On the other hand, if $\delta[l] = (+, \textunderscore)$ (in which case $S_0[l] = +$) but $\delta'[l] = (-,\_)$, then from the way combine works, either $\delta_L[l] = (-,\_)$ or $\delta_R[l] = (-, \_)$.
+WLOG suppose that $\delta_L[l] = (-,\_)$.
+From the IH, $S^L_m[l] = -$.
+From lemma \ref{sign_projection_lemma} $S_n[l] = - = sign(\delta')[l]$.
+
+\textbf{Case $l \in L(\delta_L), l \not\in L(\delta)$}: Then $l \not\in \delta_R$ so $\delta'[l] = \delta_L[l]$.
+From IH, $sign(\delta_L) = S^L_m$.
+Since $l \not\in \delta_R$, $l \not\in S^R_k$ so by applying lemma \ref{sign_projection_lemma}, $S^L_m[l] = S_n[l]$, as desired.
+
+The case where $l \in L(\delta_R), l \not\in L(\delta)$ is symmetric.
+
+Finally, we note that $(v_1||v_2) = (e^L_n||e^R_n)$ from the way $v_1,v_2$ were relabeled.
+
+\textbf{Part 2}:
+
+We case on arbitrary $l \in \delta'$.
+
+\textbf{Case $l \in L(\delta)$}:
+
+If $\delta[l] = (-,c)$ then $EC_l(\delta, \delta') = 0 = SC_l(T)$.
+
+Else suppose $\delta_L[l] = (+,c_1)$ and $\delta_R[l] = (+,c_2)$. Then $SC_l(T) = SC_l(T_L) + SC_l(T_R) \geq EC_l(\delta, \delta_L) + EC_l(\delta, \delta_R) = EC_l(\delta, \delta')$.
+
+Else suppose $\delta_L[l] = (+,c_1)$ and $\delta_R[l] = (-,c_2)$.
+We can show that all the cheap \get{}s in $T_R$ are cheap in $T$.
+So $SC_l(T) \geq SC_l(T_R) \geq EC_l(\delta, \delta_R) = EC_l(\delta, \delta)$.
+
+The case where $\delta_L[l] = (-,c_1)$ and $\delta_R[l] = (+,c_2)$ follows by symmetry.
+
+Finally, if $\delta_L[l] = (-,c_1)$ and $\delta_R[l] = (-,c_2)$, then $SC_l(T) \geq 0 = EC_l(\delta, \delta')$.
+
+\textbf{Case $l \in L(\delta_L), l \not\in L(\delta)$} Then $EC_l(\delta, \delta') = EC_l(\delta, \delta_L)$ since $l \not\in \delta_R$.
+By IH, $EC_l(\delta, \delta_L) \leq SC_l(S_0, S^L_m)$.
+Since $l \not\in S^R_k$, $SC_l(S_0, S^L_m) = SC_l(S_0, S_n)$ and so the claim holds.
+
+The case where $l \in L(\delta_R), l \not\in L(\delta)$ is similar.
+
+\textbf{Part 3:} We omit the formal proof of this part because it contains tedious algebraic details but the key ideas are described below.
+
+In the conservation expression, each \get{} on each array $(l,V)$ in both the evaluational and structural dynamics is charged $f(size(l))$.
+In particular, note that we add an additional $f(size(l))-1$ cost for cheap \get{}s that were charged 1 unit of work.
+Since the number of \get{}s is the same, the total cost of \get{}s is the same in both dynamics.
+
+The number of \set{}s that are charged $g(size(l))$ is the same for both the evaluational and structural dynamics.
+This can be shown by casing on the signs of the array on both sides of the fork.
+If and only if \set{} was called on an array on both sides of the fork, one of the \set{}s will cost $g(size(l))$.
+Since the number of \set{}s is the same, the total cost of \set{}s is the same in both dynamics.
+\end{proof}
+
+\begin{theorem}
+(Work Bound) Given the conditions in theorem \ref{main_work_theorem}, $w_E \geq W(T)$
+\end{theorem}
+
+\begin{proof}
+We assume that $\delta_E$ has been relabeled as described in part 1 of theorem \ref{main_work_theorem}.
+For all $l \in \delta_E$, $EC_l(\delta, \delta_E) \leq SC_l(T)$.
+This implies that $\displaystyle \sum_{l \in \delta_E} EC_l(\delta, \delta_E) (f(size(l))-1) \leq \sum_{l \in \delta_E} SC_l(T) (f(size(l))-1)$.
+But then from the conservation of sum of work and cheap gets, $w_E \geq W(T)$.
+\end{proof}
+
+\begin{theorem}
+(Existence) Suppose that $\delta, e_0 \Downarrow \delta_E, e_E, w_E, s_E$. Then, for some $n$, there exists a transition sequence $T$ from $s_0 = sign(\delta), e_0 \to^n s_n, e_n$.
+\end{theorem}
+
+\begin{proof}
+We can construct such a transition sequence $T$ inductively, by inducting on the rules of the evaluational dynamics.
+\end{proof}
+
+\begin{theorem}
+(Termination) Suppose that $\delta, e_0 \Downarrow \delta_E, e_E, w_E, s_E$. Then there exists $n$ s.t. all transition sequences starting at $sign(\delta), e_0$ have length $\leq n$.
+\end{theorem}
+
+\begin{proof}
+By induction on the rules of the evaluational dynamics.
+\end{proof}
+
+When evaluating an expression $e$ in our language, the stores are initially empty.
+In particular, the signs of the store in the evaluational dynamics and structural dynamics are the same.
+From the existence theorem, we know there exists a transition sequence $T$ starting at $e$.
+From the termination theorem, all transition sequences starting at $e$ have bounded length.
+Then, from the work bound theorem, we get that the work computed by the evaluational dynamics is more than the work of any transition sequence.
+Since the structural dynamics captures the (non-deterministic) execution time of the expression, this means that the evaluational dynamics overestimates the execution time of the expression.
+
+\subsection{Parallel Structural Dynamics}
+
+The parallel dynamics are given by the following judgement, where $e$ is the expression and $e'$ is the resulting expression:
+$$e \to_{par} e'$$
+The parallel dynamics represent the evaluation steps taken on a machine with infinite processors. We show the dynamics for get and fork-join.
+$$\frac{A \; \text{val} \;\;\; i \; \text{val}}{\tget{} \; A \; i \to_{par} A[i]} \text{ (get)}$$
+$$\frac{A \; \text{val} \;\;\; e_2 \to_{par} e_2'}{\tget{} \; e_1 \; e_2 \to_{par} \tget{} \; e_1 \; e_2'} \text{ (get-index)}$$
+$$\frac{e_1 \to_{par} e_1'}{\tget{} \; e_1 \; e_2 \to_{par} \tget{} \; e_1' \; e_2} \text{ (get-array)}$$
+$$\frac{e_1 \to_{par} e_1' \;\;\; e_2 \to_{par} e_2'}{e_1 || e_2 \to_{par} e_1' || e_2'} \text{ (fork-join)}$$
+$$\frac{v_1 \; \text{val} \;\;\; e_2 \to_{par} e_2'}{v_1 || e_2 \to_{par} v_1 || e_2'} \text{ (fork-join-right)}$$
+$$\frac{v_2 \; \text{val} \;\;\; e_1 \to_{par} e_1'}{e_1 || v_2 \to_{par} e_1' || v_2} \text{ (fork-join-left)}$$
+
+Notice that in fork-join, both sides of the fork-join take a step at the same time if possible.
+The other rules are similar to the rules for get.
+The parallel structural dynamics, unlike the sequential structural dynamics, are deterministic.
+
+A parallel transition sequence $T$ is a sequence $(e_0, ..., e_n)$ with $e_i \to_{par} e_{i+1}$ for all $0 \leq i < n$.
+We say that $e_0 \to^n_{par} e_n$.
+The span of $T$, denoted by $SP(T)$, is its length $n$.
+The span represents the number of steps taken to evaluate an expression on a machine with an unbounded number of processors.
+
+\subsection{Parallel Cost Analysis}
+
+\begin{theorem}
+Suppose that 
+$$\delta, e_0 \Downarrow \delta_E, e_E, w_E, s_E$$
+and consider arbitrary parallel transition sequence $T$ taking
+$$e_0 \to^n_{par} e_n$$
+Then $SP(T) \leq s_E$
+\end{theorem}
+
+\begin{proof}
+By induction on the rules of the evaluational dynamics.
+We prove the theorem for 2 of the rules in the evaluational dynamics: get and fork-join.
+The other cases follow from a similar line of reasoning.
+\\
+
+\noindent \textbf{Case get:} Suppose $e_0 = get(e_L, e_R)$ and
+$$\delta, e_L \Downarrow \delta_L, e_L', w_L, s_L$$
+$$\delta_L, e_R \Downarrow \delta_R, e_R', w_R, s_R$$
+$$\delta_R, get(e_L', e_R') \Downarrow \delta_E, e_E, 1, 1$$
+which gives us
+$$\delta, get(e_L, e_R) \Downarrow \delta_E, e_E, w_E, s_E$$
+with $s_E = s_L + s_R + 1$.
+
+The parallel structural dynamics first step the left argument to get, then the right argument, and finally evaluate the get.
+So we can split the parallel transition sequence $T$ into corresponding subsequences $T_L$, $T_R$, $T_g$.
+By the inductive hypothesis, $SP(T_L) \leq s_L$ and $SP(T_R) \leq s_R$.
+Also, $SP(T_g) = 1$.
+So $SP(T) = SP(T_L) + SP(T_R) + SP(T_g) \leq s_L + s_R + 1 = s_E$, and the claim holds.
+\\
+
+\noindent \textbf{Case fork-join}: Suppose $e_0 = (e_L || e_R)$ and
+$$\delta, e_L \Downarrow \delta_L, v_L, w_L, s_L$$
+$$\delta, e_R \Downarrow \delta_R, v_R, w_R, s_R$$
+which gives us
+$$\delta, (e_L || e_R) \Downarrow \delta', (v_1 || v_2), 1 + w_L + w_R + w', max(s_L, s_R)$$
+
+We can split the parallel transition sequence $T$ into two sub-sequences, $T_1$ where both sides of the fork-join take a step, and $T_2$ where only one side of the fork-join takes a step.
+We can project $T$ to a transition sequence $T_L$ for the left side of the fork-join  and a transition sequence $T_R$ for the right side of the fork-join.
+From the IH, $SP(T_L) \leq s_L$ and $SP(T_R) \leq s_R$.
+Then, $SP(T) \leq max(SP(T_L), SP(T_R)) \leq max(s_L, s_R) = s_E$, so the claim holds.
+\end{proof}
+
+The above theorem shows that the span in the evaluational dynamics is an upper bound for the span in the parallel structural dynamics.
+
+In reality, each step in our parallel structural dynamics could take up to $\log{N}$ steps when run on a machine with an unbounded number of processors (where N is the size of the array being operated on).
+Note that if an expression includes an array of size $N$, the array must have been created.
+Creating an array of size $N$ has work at least $N$, so $N \leq W$.
+This means that the time taken to evaluate $e$ on a machine with an unbounded number of processors is $\leq s_E \log{W}$.
+
+(Explain what Brent's theorem says) By applying Brent's theorem, we get that on a $p$ processor machine, the number of steps taken to evaluate an expression is $\leq \frac{W}{p} + s_E \log{W}$.
+
+\newpage
+
+\subsection{Correctness}
+
+(\textbf{Note: Everything from this point is going to be very significantly changed in the near future})
+
+A formal proof of correctness involves many steps, some of which are orthogonal to the goals of this paper.
+We give an outline of what is needed to prove that our array implementation is correct, and show the important parts of the proof.
+
+Step 1: First, we need to define an imperative target language and a target machine that supports concurrent memory accesses.
+We need to give an implementation of arrays that deals with concurrency issues.
+In the implementation, the execution of functions on the same array might overlap, so we prove that the implementation is linearizable.
+Linearizability allows us to assume that array functions happen atomically.
+We give a (not formal) sketch of this in a later section.
+
+Step 2: With atomicity, the target language is equivalent to an intermediate language where array operations happen in a single step, but have the same effect as in the target language.
+Like in the source language, the structural dynamics for fork-join in the intermediate language will be non-deterministic.
+
+Step 3: Finally, we need to show that an expression $e$ in the source language, and the compiled expression $e_I$ in the intermediate language will evaluate to the same value.
+This involves setting up a relation that describes when an expression $e$ in the source language and $e_I$ in the intermediate language are equivalent.
+Suppose $e_I \to e_I'$.
+We need to show that $\exists e'$ s.t.
+$e \to e'$ and $e'$ and $e_I'$ are equivalent.
+Since the value $e$ in the source language evaluates to is independent of the steps it takes, we can use induction to show that $e$ and $e_I$ evaluate to the same terminal value.
+
+The core part of steps 2 and 3 is showing sequential correctness of functional arrays.
+We define a data structure invariant, show that it is maintained by all array functions, and show that this implies that all calls to get return the value that they should.
+
+\begin{definition}
+Consider an array $A_I$ in the implementation and an array $A$ in the source language.
+We say that $A_I ~ A$ if the following holds.
+Consider arbitrary index $i$ and suppose that $A[i]$ has value $val$.
+Suppose that $A_I$ has version $V$ and is pointing to an ArrayData object $AD$.
+If there exists some version in $AD.logs[i]$ which is greater than $V$ then there exists a log entry $(v', val)$ where $v'$ is the smallest version in $AD.logs[i]$ that is $\geq V$.
+If no version in $AD.logs[i]$ is greater than $V$, then $AD.values[i]$ is val.
+\end{definition}
+
+\begin{theorem}
+Assuming that $f(1), ...f(n)$ evaluates to the same value in the source and implementation, $\text{tabulate}(f, n)$ evaluates to equivalent arrays in the source and implementation.
+\end{theorem}
+
+\begin{proof}
+\end{proof}
+
+\begin{theorem}
+Suppose that $A_I$ in the implementation and $A$ in the source are equivalent. Then $get(A,i)$ in the source and $get(A_I, i)$ in the implementation returns the same value.
+\end{theorem}
+
+\begin{proof}
+\end{proof}
+
+\begin{theorem}
+Suppose we have arrays $A_1, A_2, ..., A_m$ in the source and $B_1, B2, ..., B_m$ in the implementation with $A_i ~ B_i$ for all $i$.
+Suppose we execute $set(A_m, i, v)$ by $A$ and $set(B_m, i, v)$ by $B$.
+Then after executing the sets, $A_i ~ B_i$ for all $i$ and $A ~ B$.
+\end{theorem}
+
+\begin{proof}
+\end{proof}
 
@@
 
-An array $A$ is NEW if the set function has not been called with $A$ as an argument, otherwise $A$ is OLD.
-While the dynamics of our arrays are functional, the costs are different for OLD and NEW arrays.
+An array $A$ is a leaf array if the set function has not been called with $A$ as an argument, otherwise $A$ is an interior array.
+While the dynamics of our arrays are functional, the implementation is different for interior and leaf arrays.
 
@@
 
-Suppose that $A$ is a NEW functional array  (see figure \ref{fig:new_array_A}).
+Suppose that $A$ is a leaf array  (see figure \ref{fig:new_array_A}).
 $A$ has a version number $V$, and a pointer to an ArrayData object $AD$.
@@
 
-Get on NEW arrays: to get the $i^{\text{th}}$ element in $A$, simply access array[i] in $AD$.
+Get on leaf arrays: to get the $i^{\text{th}}$ element in $A$, simply access array[i] in $AD$.
 
@@
 
-Set on NEW arrays: Suppose that $A[i]$ is $v_{old}$ and we want to do $B = set(A, i, v_{new})$ (see figure \ref{fig:set_A_return_B}).
+Set on leaf arrays: Suppose that $A[i]$ is $v_{old}$ and we want to do $B = set(A, i, v_{new})$ (see figure \ref{fig:set_A_return_B}).
 Add an entry $(V, v_{old})$ to the log at index $i$, expressing that $A[i]$ was $v_{old}$ at version $V$.
@@
 Notice that both $A$ and $B$ point to $AD$.
-However, the version of $A$ is $V$ and the version of $AD$ is $V+1$, which indicates that $A$ is OLD.
+However, the version of $A$ is $V$ and the version of $AD$ is $V+1$, which indicates that $A$ is an interior array.
 The version of $B$ is $V+1$ which indicates that $B$ is NEW.
 
-Get on OLD array: Suppose we do $get(A, i)$ where $A$ is OLD and has version $V$. We binary search the log at index $i$ to find what value was stored in the array at version $V$.
+Get on interior array: Suppose we do $get(A, i)$ where $A$ is interior and has version $V$. We binary search the log at index $i$ to find what value was stored in the array at version $V$.
 
@@
 
-Suppose we do $D = set(A, i,v)$ where $A$ is old (see figure \ref{fig:set_old_array}).
+Suppose we do $D = set(A, i,v)$ where $A$ is interior (see figure \ref{fig:set_old_array}).
 Create a new ArrayData object $AD'$ (with a new array).
@@
 
-It follows that in the RAM model the work of set and get is $O(1)$ for a NEW array.
+It follows that in the RAM model the work of set and get is $O(1)$ for a leaf arrays.
 Get in an OLD array involves a binary search and has work $O(\log{n})$.
@@
 
-\begin{theorem}
-The non-deterministic structural dynamics and the evaluational dynamics produce the same result in the source language.
-\end{theorem}
-
-\begin{proof}
-This is a standard implicit parallelism type theorem, for example see Chapter 39 in~\cite{harper}, so we omit the details. 
-\end{proof}
-
 Now, to show that our implementation correctly implements the source language, it suffices to show that any valid execution of our implementation produces the same result as a little step transition sequence in the source language (since all transition sequences produce the same value as the evaluational dynamics).

### 2016-09-25 Paper/rebuttal.txt
Reorganize repository

Thanks to all the reviewers for detailed and useful feedback.

Review A:

- Thanks for your detailed comments, we will fix the issues you identified.

Review B:

- Thanks for the feedback on our presentation. We will make the workings of the data structure more clear.

- We will work on the clarity of our correctness sketch. Our argument is rely-guarantee, but we cite Birkedal et al for its extension to a higher order language using logical relations, which we depend on.

- Haskell's Diffarrays are an implementation of Bloss' idea of "trailers" (from "Update Analysis and the Efficient Implementation of Aggregates”). We cite the improved version by Chuang, but will add the earlier cite by Bloss and point out that Diffarrays use it. Diffarrays differ from ours in that: (1) reading in the past takes O(m) time, where m is the number of updates made (i.e. it is not even a function of the size), (2) it does not support concurrency, (3) there is no cost semantics.

Review C:

- Thanks for your detailed comments, we will fix all identified issues and ambiguities.

- O(n) vs O(n^2) time question: For every n operations on a leaf sequence, we only make a copy of the values in the leaf sequence. We don’t copy interior sequences, or the change logs stored at the leaf sequence. The aim is to keep the size of the change logs small - the newly copied sequence will have empty change logs. We will make this logic more clear in our paper.

- Exploring functional hashtables is a great idea.

Review D:

- Your observations are all valid. The sequential version of our data structure is conceptually simple, and the sequential analysis is easy, and not of significant interest from an algorithmic point of view. The main goal of our paper is to give a concurrent implementation, and analyze its complexity formally in a cost semantics. We present a way to do this without having to consider all possible interleavings of threads, and the analysis can generalize to other structures. From the data structures point of view, we are also interested in a wait-free concurrent implementation. We will identify our main contributions more clearly.

- Thanks for your detailed feedback on our experiments. Indeed, it is easy to construct cases where our data structure performs poorly (e.g. repeatedly setting interior nodes). Our aim was to give preliminary results that suggest our data structure works well if most operations are on leaf nodes, even if the data structure is used concurrently (the efficiency of leaf updates is not obvious due to potential overheads of the implementation). As you point out, more systematic testing should be done in future work.

Review E:

- Thanks for the detailed review, the full code (in SML and Java) is available at https://github.com/AnanyaKumar/PersistentArrays

- Indeed amortization would be a problem if applied to the span (depth), but this does not pose a problem for us since the copy can be done in parallel so the span of the copy is O(log(n)) in the worst case, not amortized (section 6.6). The amortization is only with respect to the work, which is valid since work adds. Given the work and depth, we rely on prior work to compute the costs for a machine with p processors (theorem 3.1). This analysis works even if the work is amortized.

- We will clarify the ambiguity about size vs capacity in Pusharray.new

- We will clarify that our algorithm only improves on the READ accesses to old versions (compared to Chuang’s array)

### 2016-09-25 Paper/conclusion.tex
Reorganize repository

\section{Future Work}
\label{sec:conclusion}
\begin{enumerate}
\item We do not have formal proofs for the cost bounds of our sequence implementation. Current techniques for analyzing concurrent data structures focus on proving correctness. Coming up with a framework to prove the costs of a concurrent implementation is an exciting opportunity. Note that our main results still hold without these proofs, because we present a general way to analyze data types where operating on leaf values and interior values have different costs.

\item Extending the approach of allowing operations on interior data values to be more expensive than for leaf data values to other data types (for example unordered sets or functional disjoint sets) is a possible future direction.

\item We conjecture that it is not possible for a sequence implementation to have constant (worst case) costs for all operations, which would ground the need for an implementation where costs for leaf and interior versions are different.

\item Our cost dynamics could be extended to situations that generalize beyond separate costs for interior and leaf versions. In particular, our interleaved structural dynamics does not give a tight bound for the cost of our sequence implementation. In our implementation, accessing a value in an interior version could involve constant work if there are no (or a constant number of) log entries at that index.

\end{enumerate}

### 2016-09-25 Paper/costproof.tex
Reorganize repository

--- 2016-03-11 Paper/paper.tex
@@
-%-----------------------------------------------------------------------------
-%
-%               Template for sigplanconf LaTeX Class
-%
-% Name:         sigplanconf-template.tex
-%
-% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
-%               file for SIGPLAN conference proceedings.
-%
-% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
-%               sigplanconf-guide.pdf
-%
-% Author:       Paul C. Anagnostopoulos
-%               Windfall Software
-%               978 371-2316
-%               paul@windfall.com
-%
-% Created:      15 February 2005
-%
-%-----------------------------------------------------------------------------
+\section{Cost Proofs}
+\label{sec:costproof}
 
-\documentclass[preprint]{sigplanconf}
-
-% The following \documentclass options may be useful:
-
-% preprint      Remove this option only once the paper is in final form.
-% 10pt          To set in 10-point type instead of 9-point.
-% 11pt          To set in 11-point type instead of 9-point.
-% numbers       To obtain numeric citation style instead of author/year.
-
-\usepackage{amsmath}
-\usepackage{graphicx}
-\usepackage{amsthm}
-\usepackage{listings}
-
-\newcommand{\cL}{{\cal L}}
-
-\begin{document}
-
-\special{papersize=8.5in,11in}
-\setlength{\pdfpageheight}{\paperheight}
-\setlength{\pdfpagewidth}{\paperwidth}
-
-\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country}
-\copyrightyear{20yy}
-\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
-\copyrightdoi{nnnnnnn.nnnnnnn}
-
-% Uncomment the publication rights you want to use.
-%\publicationrights{transferred}
-%\publicationrights{licensed}     % this is the default
-%\publicationrights{author-pays}
-
-% \titlebanner{banner above paper title}        % These are ignored unless
-% \preprintfooter{short description of paper}   % 'preprint' option specified.
-
-\title{A New Approach for Parallel Functional Arrays}
-
-\authorinfo{Ananya Kumar}
-           {Carnegie Mellon University}
-           {ananyak@andrew.cmu.edu}
-\authorinfo{Guy Blelloch}
-           {Carnegie Mellon University}
-           {blelloch@cs.cmu.edu}
-
-\maketitle
-
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{corollary}{Corollary}[theorem]
-\newtheorem{lemma}[theorem]{Lemma}
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}[section]
-
-\begin{abstract}
-In this paper we introduce a $O(1)$ wait-free, parallel, functional array, that allows $O(1)$ reads and writes to the most recent version of the array.
-We describe the cost dynamics and sketch out a provable implementation.
-We show favorable benchmarks comparing our functional arrays with regular arrays in Java.
-\end{abstract}
-
-\category{CR-number}{subcategory}{third-level}
-
-\keywords
-array, parallel, cost semantics
-
-\section{Introduction}
-
-Arrays are very important in functional programming languages because they allow work-efficient implementations of algorithms like depth-first search.
-Accessing old versions of arrays can be useful for efficient checkpointing, logging, and event handling.
-
-\subsection{Previous Approaches}
-
-There has been a lot of great work on functional arrays, but previous approaches have limitations.
-
-In many functional programming languages, like Haskell, monads ~\cite{wadler-monads} can be used to enforce single threaded use of arrays.
-However, enforcing single threaded use obstructs persistence - it prevents us from accessing or modifying old versions of an array.
-
-Another approach is to use reference counting ~\cite{hudak}.
-If the number of references to an array is one, the array can be updated in place, otherwise the array is copied and then updated.
-This approach makes it difficult for programmers to statically analyze the time complexity of array operations.
-Furthermore, even if there are multiple reference to an array the array might not need to be copied, so the unnecessary copying could make the implementation inefficient.
-
-A third approach is to use linear types ~\cite{wadler-lintypes} to ensure that programmers can only access and mutate arrays in valid ways.
-Linear types have not been used practically for a few decades since its inception, and would require a complete re-think of a language's type system to be useful.
-
-O' Neill's ~\cite{oneill} functional arrays guarantee $O(1)$ access and insertions when used like an imperative array.
-However, the arrays explicitly implement version trees which are inefficient in practice.
-Additionally, the reads and writes to the most recent version of the array might take $O(\log{n})$ work if the array is not used in a single threaded way.
-
-Chuang's approach ~\cite{chuang} supports $O(1)$ accesses and insertions to the most recent version of arrays, however accessing old versions of the array takes $O(n)$ work which is often impractical.
-
-None of the existing approaches support concurrent operations on arrays.
-O' Neill suggests having a lock for each element of the array.
-However, when many threads contend for the same element, this would serialize accesses and make accesses not $O(1)$.
-Additionally, per-element locks add significant memory overhead.
-Separation logic ~\cite{reynolds} could be used to parallelize divide and conquer algorithms on arrays however they would severely restrict the way functional arrays can be used.
-
-\subsection{Our Approach}
-
-Our arrays support 3 functions: tabulate, get, and set. The dynamics are functional, in particular set returns a new array and leaves the original array intact.
-
-\begin{figure}[!ht]
-\centering
-\includegraphics[scale=0.45]{leaf_interior_intro}
-\nocaptionrule \caption{Example usage of functional arrays}
-\label{fig:leaf_interior_intro}
-\end{figure}
-
-After all the operations performed in figure \ref{fig:leaf_interior_intro}, A and C are \emph{interior} arrays and B and D are \emph{leaf} arrays.
-Calling get and set on leaf arrays costs $O(1)$ but calling get and set on interior arrays is more expensive.
-In the cost dynamics, we thread a store that keeps track of whether an array is an interior or leaf array.
-
-We also give and prove fork join cost dynamics for our arrays.
-The dynamics for arrays used concurrently are non trivial because one thread might call get on a leaf array while another thread calls set.
-To deal with this, we compute the work of each fork separately, and then charge additional work at the join point if an array is used in both forks.
-
-In our implementation of a functional array, we keep an array of values (multiple functional arrays might point to the same array).
-For each index of the array, we also keep a history of values that were previously at that index.
-To access an element in an interior array, we need to search through the history of values.
-
-We give a wait-free concurrent implementation that uses careful synchronization.
-We prove linearizability of the implementation and use that to prove that our arrays work correctly when used concurrently.
-Finally, we give a provable implementation for our cost dynamics.
-
-\section{Dynamics}
-
-We use a standard applicative-order language defined as follows:
-$$e = x \; | \; c  \; | \; \lambda x.e \; | \; e_1 e_2 \; | \; e_1 || e_2 \; | \; \text{if } e_1 \text{ then } e_2 \text{ else } e_3$$
-
-$L$ contains the usual arithmetic types, such as the natural numbers, and numerical operations such as sums and products.
-Our dynamics are given by the judgement $e \Downarrow v$ which means that expression $e$ evaluates to $v$.
-Dynamics for function application and fork-join are given as examples.
-
-$$\frac{e_1 \Downarrow v_1 \;\;\; e_2 \Downarrow v_2}{e_1 || e_2 \Downarrow (v_1, v_2)} \text{ (fork-join)}$$
-$$\frac{f \Downarrow \lambda x . e \;\;\; y \Downarrow v}{f \; y \Downarrow [v/x]e} \text{ (function-app)}$$
-
-The language also includes 3 functions for working with arrays: tabulate, get, and set. The dynamics for tabulate is given below:
-$$\frac{f(1) \Downarrow v_1, ..., f(n) \Downarrow v_n}{\text{tabulate} \; f \; n \Downarrow [1 \mapsto v_1, ..., n \mapsto v_n]}$$
-
-$get(A,i)$ returns the value at the $i$th index of $A$.
-$set(A,i,v)$ returns a new array where the value at index $i$ is $v$ and the value at all other indices is the same as in $A$.
-The evaluation of a simple program in our language is shown below:\\
-
-$(\lambda A. \; get(A, 2) + get(A,3)) set(tabulate(\lambda i . \; i, 5), 2, 10)$
-
-$(\lambda A. \; get(A, 2) + get(A,3)) set([1,2,3,4,5], 2, 10)$
-
-$(\lambda A. \; get(A, 2) + get(A,3)) [1,10,3,4,5]$
-
-$get([1,10,3,4,5], 2) + get([1,10,3,4,5],3) $
-
-$10+3$
-
-$13$
-
-\section{Cost Dynamics}
-
-The dynamics given in the previous section are functional.
-However, the costs of get and set depend on whether the array is a leaf array or an internal array.
-To deal with this, we thread a store through the cost dynamics to keep track of whether each array is a leaf or internal array.
-
-The cost dynamics of fork-join pose an additional challenge.
-An array can be used concurrently by multiple threads and the costs could depend on the order in which the instructions are interleaved.
-Figure \ref{fig:fork_join_intuition} shows 3 ways a leaf array $A$ might be used.
-In particular, one side of the fork might call get on $A$ while the other side calls set (see the center panel).
-Calls to get on the left branch would be $O(1)$ if and only if they execute before the call to set on the right branch.
-
-\begin{figure}[!ht]
-\centering
-\includegraphics[scale=0.3]{fork_join_intuition}
-\nocaptionrule \caption{Different ways an array might be used in a fork join}
-\label{fig:fork_join_intuition}
-\end{figure}
-
-We cannot make any assumptions about how the instructions in different branches of a fork are interleaved, so we give the worst case time complexity over all possible interleavings.
-We first compute the work of each fork separately.
-Then for arrays that are used in both forks, we add additional work at the join point.
-To compute the additional work at the join point, we keep track of the number of gets that cost $O(1)$ on each side of the fork.
-If the array was modified on the other side of the fork, then the modification might have happened before the get, so we need to charge additional work for the get.
-
-Our cost dynamics are defined by the following judgement, where $\delta$ is the store, $e$ is the expression, $\delta'$ is the new store, $v$ is the value returned by evaluating $e$, $w$ is the work, and $s$ is the span.
-$$\delta; e \boldsymbol\Downarrow \delta'; v; w; s$$
-
-We give the cost dynamics for expressions in the language: (insert cost dynamics for standard functions later)
-
-Arrays are represented by $(l, V)$ where $l$ is a label that uniquely identifies the array, and $V$ contains the values in the array.
-$V[i]$ is the $i^{\text{th}}$ value in the array.
-If the set function has been called on an array, the array is an interior array (represented by $-$), otherwise the array is a leaf array (represented by $+$).
-
-Array functions have different costs depending on whether the array is a leaf or interior array.
-Let $C(l)$ be the number of get operations applied on array $A = (l, V)$ that had $O(1)$ work.
-We thread a store $\delta$ through each operation.
-$\delta$ maps array labels $l$ to $(+/-, c)$, where $c = C(l)$.
-Let $f(n)$ be the work of calling get on an interior array and $g(n)$ be the work of calling set on an interior array.
-
-The cost dynamics for the functions on a functional array of size $n$ are:
-$$\frac{l = \text{new label} \;\;\; f(1) \Downarrow v_1; w_1; s_1; ...
-f(n) \Downarrow v_{n}; w_{n}; s_{n}}{\delta; \text{tabulate } f \; n \Downarrow \delta[l \mapsto (+, 0)]; (l, \bigcup_{i=1}^n [i \mapsto v_i]); 1+\sum w_i; 1+\max s_i}$$
-$$\frac{A = (l,V) \;\;\; \delta[l \mapsto (+, c)] \;\;\; l' = \text{new label}}{\delta; \text{set} \; A \; i \; v \Downarrow \delta[l \mapsto (-, c), l' \mapsto (+, 0)]; (l', V[i \mapsto v]); 1; 1} \text{  (set-leaf)}$$
-$$\frac{A = (l,V) \;\;\; \delta[l \mapsto (-, c)] \;\;\;  l' = \text{new label}}{\delta; \text{set} \; A \; i \; v \Downarrow \delta[l' \mapsto (+, 0)]; (l', V[i \mapsto v]); g(n); 1} \text{  (set-interior)}$$
-$$\frac{A = (l,V) \;\;\; \delta[l \mapsto (+, c)]}{\delta; \text{get} \; A \; i \Downarrow \delta[l \mapsto (+, c+1)]; V[i]; 1; 1} \text{  (get-leaf)}$$
-$$\frac{A = (l,V) \;\;\; \delta[l \mapsto (-, c)]}{\delta; \text{get} \; A \; i \Downarrow \delta; V[i]; f(n); 1} \text{  (get-interior)}$$
-
-Set creates a new label and array and modifies the store to indicate that the newly created array is a leaf array, and the array set was called on has become an interior array.
-The work for set is higher for interior arrays.
-Get returns the value at the desired index, if the array is a leaf then the operation costs $O(1)$ but we increment the counter of $O(1)$ gets in the store.
-
-The fork-join cost semantics are the most interesting. Suppose we have an expression $e_L || e_R$ and
-\[ \delta, e_L \Downarrow \delta_L; v_L; w_L, s_L \]
-\[ \delta, e_R \Downarrow \delta_R; v_R; w_R, s_R \]
-
-We assume that newly created labels in $\delta_L$ and $\delta_R$ do not conflict.
-Consider an array $A = (l, V)$ with $\delta[l] = (s,c)$, $\delta_L[l] = (s_L, c + c_L)$, $\delta_R[l] = (s_R, c + c_R)$.
-When multiplying two signs or multiplying a sign with an integer, consider + to be 1 and - to be 0.
-
-Combine describes how to combine the values in the store on both sides of the fork-join.
-The array is a leaf array iff it is a leaf array on both sides of the fork-join.
-$O(1)$ gets on one side of the fork remain $O(1)$ iff there were no calls to set on the other side of the fork.
-\begin{equation*}
-  \begin{aligned}
-   combine&((s, c), (s_L, c + c_L), (s_R, c + c_R))= \\
-   &(s_L s_R, c + s_R c_L + s_L c_R)
-  \end{aligned}
-\end{equation*}
-Extrawork gives the additional cost incurred if an array was modified on either side of a fork-join.
-If the array was an interior array before the fork-join, then all functions incurred their maximal cost and there is no additional cost.
-Otherwise, $O(1)$ gets on one side of the fork become $O(f(n))$ iff the other side of the fork called set.
-If both sides of the fork called set, then one of the sets came first and is $O(1)$, the subsequent set is $O(g(n))$.
-\begin{equation*}
-  \begin{aligned}
-  extrawork&((s, c), (s_L, c + c_L), (s_R, c + c_R)) = \\
-  &s \cdot ( (\neg s_R) c_L f(n)  + (\neg s_L) c_R f(n) + (\neg s_R) (\neg s_L) g(n) )
- \end{aligned}
-\end{equation*}
-
-\begin{equation*}
-  \begin{aligned}
-  \delta' = \; &(\delta_L / \delta) \; \cup \; (\delta_R / \delta) \; \cup \\
-               &\bigcup_{l \in L(\delta_L) \cap L(\delta_R)} [l \mapsto combine(\delta[l], \delta_L[l], \delta_R[l])]
-  \end{aligned}
-\end{equation*}
-$$w' = \sum_{l \in L(\delta_1) \cap L(\delta_2)} extrawork(\delta[l], \delta_L[l], \delta_R[l])$$
-Then, the fork-join cost semantics are:
-$$\delta; e_1 || e_2 \Downarrow \delta'; (v_1, v_2); 1 + w_1 + w_2 + w'; 1 + \max(s_1, s_2)$$
-
-\section{Approach}
-
-An array $A$ is NEW if the set function has not been called with $A$ as an argument, otherwise $A$ is OLD.
-While the dynamics of our arrays are functional, the costs are different for OLD and NEW arrays.
-
-\begin{figure}[!ht]
-\centering
-\includegraphics[scale=0.3]{new_array_A}
-\nocaptionrule \caption{New functional array filled with 0s}
-\label{fig:new_array_A}
-\end{figure}
-
-Suppose that $A$ is a NEW functional array  (see figure \ref{fig:new_array_A}).
-$A$ has a version number $V$, and a pointer to an ArrayData object $AD$.
-$AD$ keeps a regular (mutable) array of values, which corresponds to the values in $A$.
-$AD$ has a version number, which is the same as $A'$s ($V$) to indicate that $A$ is NEW.
-For each element of the array in $AD$, keep a log of values that used to be at that index.
-The logs are initially empty.
-
-Get on NEW arrays: to get the $i^{\text{th}}$ element in $A$, simply access array[i] in $AD$.
-
-\begin{figure}[!ht]
-\centering
-\includegraphics[scale=0.3]{set_A_return_B}
-\nocaptionrule \caption{$B = set(A, 3, 5)$ changes the array in $AD$ and adds a log entry.}
-\label{fig:set_A_return_B}
-\end{figure}
-
-Set on NEW arrays: Suppose that $A[i]$ is $v_{old}$ and we want to do $B = set(A, i, v_{new})$ (see figure \ref{fig:set_A_return_B}).
-Add an entry $(V, v_{old})$ to the log at index $i$, expressing that $A[i]$ was $v_{old}$ at version $V$.
-Increment the version of $AD$ to $V+1$.
-Set $array[i]$ in $AD$ to $v_{new}$.
-Create a new functional array $B$, with version $V+1$, which points to $AD$.
-Notice that both $A$ and $B$ point to $AD$.
-However, the version of $A$ is $V$ and the version of $AD$ is $V+1$, which indicates that $A$ is OLD.
-The version of $B$ is $V+1$ which indicates that $B$ is NEW.
-
-Get on OLD array: Suppose we do $get(A, i)$ where $A$ is OLD and has version $V$. We binary search the log at index $i$ to find what value was stored in the array at version $V$.
-
-\begin{figure}[!ht]
-\centering
-\includegraphics[scale=0.3]{set_old_array}
-\nocaptionrule \caption{$D = set(A, 1, 9)$ copies $AD$ because $A$ is OLD}
-\label{fig:set_old_array}
-\end{figure}
-
-Suppose we do $D = set(A, i,v)$ where $A$ is old (see figure \ref{fig:set_old_array}).
-Create a new ArrayData object $AD'$ (with a new array).
-Copy the values from the array in $AD$ to the array in $AD'$, and then set $array[i]$ to $v$ in $AD'$.
-
-The logs are stored in a doubling array which doubles in size when full.
-This guarantees amortized $O(1)$ insertion of a log entry, and allows us to binary search with $O(\log{m})$ work, where $m$ is the number of log entries.
-
-Suppose that the size of the array is $n$.
-Every $n$ times an ArrayData object is modified, we create a new ArrayData object, and copy the values over to the new ArrayData object.
-This ensures that $m \leq n$ i.e.
-we don't have too many log entries in an ArrayData object.
-Copying is amortized $O(1)$.
-
-It follows that in the RAM model the work of set and get is $O(1)$ for a NEW array.
-Get in an OLD array involves a binary search and has work $O(\log{n})$.
-Set in an OLD array involves copying the array, and has work $O(n)$.
-
-\section{Concurrent Implementation}
-
-We implement the functional array functions using an imperative target language.
-
-\subsection{Concurrency Model}
-
-We assume a sequential consistency model of concurrency.
-That is, when analyzing the effects of a program, we assume that at each time step exactly one pending instruction is executed.
-Consider the execution of a program $P$.
-We label the time steps in the execution from 1 to $t$.
+We show that the work computed by the evaluational dynamics is a tight upper bound for the work in the interleaved structural dynamics.
+Proofs for the span bounds are omitted because they are standard (the parallel structural dynamics is deterministic).
 
 \begin{definition}
-Consider an object $A$.
-Suppose that the program invokes functions $M_1$, $M_2$, ..., $M_n$ on $A$.
-Consider arbitrary $i$, and suppose that the instructions in $M_i$ were executed at times $s_1$, $s_2$, ..., $s_m$.
-We say that $A$ is \emph{linearizable} if the effect of the program is as if $M_i$ executed atomically at some time $s_j$, and this holds for all $M_i$.
+Consider a transition sequence $T$.
+We say that $S_i, e_i \to S_{i+1}, e_{i+1}, w_{i+1}$ is a \get{} on $l$ if the step involves evaluating the \get{} function on structure $(l, V)$.
+We say it is a \emph{cheap \get{}} on $l$ if $w_{i+1} = g_l(l)$.
+The number of cheap \get{}s on $l$ in $T$ is denoted by $SC_l(T)$.
 \end{definition}
@@
 \begin{definition}
-Suppose that first instruction in function call $f$ was executed at time $s_1$ and the last instruction at time $e_1$.
-Suppose that first instruction in function call $g$ was executed at time $s_2$ and the last instruction at time $e_2$.
-We say the function calls do not \emph{overlap} if the intervals $[s_1, e_1]$ and $[s_2, e_2]$ do not overlap.
+Unlike the store in the evaluational dynamics, the store in the structural dynamics only stores whether each structure is $+$ (leaf) or $-$ (interior).
+$sign(\delta)$ takes a store that maps $l \mapsto (s,c)$ and returns a store which maps $l \mapsto s$.
 \end{definition}
 
-Our programs have access to the link load store conditional (LLSC) function.
-LLSC takes an (address, old value, new value) tuple as argument.
-LLSC first checks that the value at address is the same as old value, if it is different then LLSC returns false.
-Otherwise LLSC atomically stores new value at address and returns true, however if the value at address was modified between the load and store, LLSC returns false.
+\begin{definition}
+Suppose that $\delta, e \Downarrow \delta', e', w', s'$.
+We define the number of cheap \get{}s in going from $\delta$ to $\delta'$ in the evaluational dynamics as follows.
+Suppose $l \mapsto (s', c') \in \delta'$.
+If $l \mapsto (s,c) \in \delta$ then $EC_l(\delta, \delta') = c' - c$ and if $l \not\in L(\delta)$ then $EC_l(\delta, \delta') = c'$.
+If $l \not\in L(\delta')$ then $EC_l(\delta, \delta') = 0$.
+\end{definition}
 
-\subsection{Push Arrays}
+\begin{definition}
+A \emph{relabeling} of $L_1$ is a bijective function $R$ from label sets $L_1 \to L_2$.
+$R$ can be used to relabel stores and expressions.
+$R(\delta) = \{ R(l) \mapsto e \; | \; l \mapsto e \in \delta \wedge l \in L_1  \} \cup \{ l \mapsto e \; | \; l \mapsto e \in \delta \wedge l \not\in L_1  \}$.
+In other words, $R$ relabels some of the labels in $\delta$.
+Similarly, $R(e)$ returns an expression $e'$ where each occurrence of a label $l \in L_1$ in $e$ is substituted by $R(l)$.
+\end{definition}
 
-We use an auxiliary data structure, \emph{PushArrays}, to store log entries.
-PushArrays are initially empty and support 3 functions.
-Suppose we have a PushArray $A$.
-$A.push(e)$ inserts entry $e$, $A.size()$ returns the number of entries inserted, $A.get(i)$ returns the $i^{\text{th}}$ entry inserted.
-Note that $get$ is only defined between indices 0 and $A.size()-1$.
-All operations are amortized $O(1)$.
+\begin{lemma}
+\label{relabeling_lemma}
+Suppose that there exists a derivation of length $m$ that $\delta, e \Downarrow \delta', e', w, s$.
+Let $R$ be an arbitrary labeling.
+Then there exists a derivation of length $m$ that $R(\delta), R(e) \Downarrow R(\delta'), R(e'), w, s$.
+\end{lemma}
 
-PushArrays can be used semi-concurrently.
-At most one thread can execute $push$ at any time, but multiple threads can call $size$ and $get$.
-In a PushArray $A$, $A.size$ stores the number of entries and $A.data$ references an array that stores the actual entries.
-Pseudocode for $A.push(e)$ is shown below.
+\begin{proof}
+By applying the relabeling to each step of the derivation, and noting that all rules in the evaluational dynamics hold under relabelings.
+\end{proof}
 
-\begin{lstlisting}[language=Java]
-if (isFull(A.data)) {
-  newData = new Array(A.data.capacity * 2);
-  copyValues(from = A.data, to = newData);
-  A.data = newData;
-}
-A.data[A.size] = e;
-A.size += 1;
-\end{lstlisting}
+\begin{lemma}
+\label{stateless_lemma}
+If $S, e \to S', e', w$ and $U$ is a store s.t. for all $l \in e$, $l \in U$, then there exists $U', w'$ s.t. $U, e \to U', e', w'$
+\end{lemma}
+
+\begin{proof}
+By casing on each rule in the structural dynamics, and noting that transitions of expressions never depend on the values in the store (the dynamics are purely functional).
+\end{proof}
 
 \begin{theorem}
-Given that different calls to push (on the same PushArray) do not overlap, PushArrays are linearizable.
+\label{main_work_theorem}
+Suppose that 
+$$\delta, e_0 \Downarrow \delta_E, e_E, w_E, s_E$$
+
+and consider arbitrary maximal transition sequence T taking 
+$$S_0 = sign(\delta), e_0 \to^n S_n, e_n$$
+å
+The following hold:
+\begin{enumerate}
+\item For some relabeling $R$ of $L(\delta_E) \setminus L(\delta)$, $S_n = sign(R(\delta_E))$ and $e_n = R(e_E)$.
+\item For all $l \in \delta_E$, $EC_l(\delta, \delta_E) \leq SC_{R(l)}(T)$.
+\item The sum of work and cheap \get{}s is conserved: \begin{align*}
+w_E + &\sum_{l \in L(\delta_E)} EC_l(\delta, \delta_E) (g_i(l)-g_l(l)) =\\
+ &W(T) + \sum_{l \in L(\delta_E)} SC_{R(l)}(T) (g_i(R(l)) - g_l(R(l)))
+ \end{align*}
+\end{enumerate}
 \end{theorem}
@@
 \begin{proof}
-The linearization point of $push$ is $A.size$ $+=$ $1$, and the linearization point of the other functions is their single instruction.
-In push, expanding the capacity of $A.data$ when it is full does not interfere with accessing $A.data$ because we copy values to the new array before pointing $A.data$ to the new array.
-The effects of adding $e$ are only observable after we increase $size$, since, by the specification of PushArrays, the result of calling $A.get(i)$ where $i \geq A.size()$ is not defined.
+By induction on the length of the shortest derivation in the evaluation dynamics.
+We prove the theorem for 2 of the rules in the evaluation dynamics: get-leaf and fork-join.
+The other cases follow from a similar line of reasoning.
+\\
+
+\noindent \textbf{Case get-leaf}: Suppose $e_0 = \tget{}(e_L, e_R)$ and
+$$\delta, e_L \Downarrow \delta_L, e_L', w_L$$
+$$\delta_L, e_R \Downarrow \delta_R, e_R', w_R$$
+$$\delta_R, \tget{}(e_L', e_R') \Downarrow \delta_E, e_E, g_l(l')$$
+where $e_L' = (l', V)$ for some $V$, which gives us
+$$\delta, \tget{}(e_L, e_R) \Downarrow \delta_E, e_E, w_E$$
+with $w_E = w_L + w_R + g_l(l')$.
+
+Note that the structural dynamics first step the left argument to \get{}, then the right argument, and finally evaluate the \get{}.
+So there exists $m$ such that $T_{0,m}$ involves stepping the left argument, $T_{m,n-1}$ involves stepping the right argument, and $T_{n-1,n}$ involves evaluating the \get{}.
+
+\textbf{Part 1}: Suppose that in $T_{0,m}$, for $0 \leq i \leq m$, $e_i = get(e_{L,i}, e_R)$.
+Consider a new projected transition sequence $T_{0,m}'$ with states $[(S_0, e_{L,0}), ..., (S_m, e_{L,m})]$ and costs the same as $T_{0,m}$: $[w_1, ..., w_m]$.
+It is easy to verify that $T_{0,m}'$ is a valid transition sequence starting at $(S_0, e_{L,0})$ and is maximal.
+
+$S_0 = sign(\delta)$ and $e_{L,0} = e_L$ so we can apply the IH to $T_{0,m}'$.
+For some relabeling $R$ of labels in $L(\delta_L) / \delta$, $s_m = sign(R(\delta_L))$ and $e_{L,m} = R(e_L')$ $\Rightarrow$ $e_m' = (e_{L,m}, e_R) = (R(e_L'), e_R)$.
+WLOG suppose that $\delta_L$ was labeled such that the above hold (we can do this because of lemma \ref{relabeling_lemma}).
+Note that in subsequent parts of the proof, we omit creating the projected transition sequence and apply the IH directly to $T_{0,m}$ when needed.
+
+Using similar logic on $T_{m,n-1}$ we get that for some relabeling of $\delta_R / \delta_L$, applying that relabeling to $\delta_R$ and $(e_L', e_R')$ gives us $e_{n-1} = (e_L', e_R')$ and $S_{n-1} = sign(\delta_R)$.
+WLOG suppose that $\delta_R$ was labeled so that the above holds.
+
+By combining the relabelings of $\delta_L / \delta$ and $\delta_R / \delta_L$ and applying that to $(e_L', e_R')$ and $\delta_R$, we get that $e_{n-1} = get(e_L', e_R')$ and $S_{n-1} = sign(\delta_R)$.
+Note that get does not change the sign of any label in the store in either the structural or the evaluation dynamics, so $sign(\delta_E) = sign(\delta_R) = S_{n-1} = S_n$.
+Further, the structural and evaluational dynamics apply get in the same way, so $e_n = e_E$.
+Therefore the first part of this theorem holds.
+WLOG suppose that the stores and expressions were relabeled s.t.
+the first part holds.
+
+\textbf{Part 2}: From lemma \ref{relabeling_lemma}, the length of the shortest derivation for the relabeled evaluational dynamics does not change, so we can apply the inductive hypothesis even after the relabeling.
+Applying the IH to $T_{0,m}$, for all $l \in \delta_L$, $EC_l(\delta, \delta_L) \leq SC_l(T_{0,m}') = SC_l(T_{0,m})$.
+If $l \in \delta_R$ but $l \not\in \delta_L$ then $EC_l(\delta, \delta_L) = SC_l(T_{0,m}) = 0$ (intuitively the label did not exist and so did not have any cheap gets), so in particular, for all $l \in \delta_R$, $EC_l(\delta, \delta_L) \leq SC_l(T_{0,m})$.
+Applying the IH to $T_{m,n-1}$ we get that for all $l \in \delta_R$, $EC_l(\delta_L, \delta_R) \leq SC(T_{m,n-1})$.
+
+We note that $EC$ and $SC$ are additive, that is,
+$$EC_l(\delta, \delta_R) = EC_l(\delta, \delta_L) + EC_l(\delta_L, \delta_R)$$
+$$SC_l(T_{0,n-1}) = SC_l(T_{0,m}) + SC_l(T_{m,n-1})$$
+This implies that for all $l \in \delta_R$, $EC_l(\delta, \delta_R) \leq SC_l(T_{0,n-1})$.
+
+Now, let $e_L' = (l',V)$ with $l' \in \delta_R$.
+Then, $EC_{l'}(\delta, \delta_E) = EC_{l'}(\delta, \delta_R) + 1$ and $SC_{l'}(T) = SC_{l'}(T_{0,n-1}) + 1$, so $EC_{l'}(\delta, \delta_E) \leq SC_{l'}(T)$.
+If $l \neq l'$ then $EC_{l}(\delta, \delta_E) = EC_{l}(\delta, \delta_R)$ and $SC_{l}(T) = SC_{l}(T_{0,n-1})$, so $EC_{l}(\delta, \delta_E) \leq SC_{l}(T)$.
+So the second part of the theorem holds.
+
+\textbf{Part 3}: The third part of the theorem follows by similarly applying the IH on $T_{0,m}$ and $T_{m,n-1}$, and noting that $W(T) = W(T_{0,m}) + W(T_{m,n-1}) + g_l(l')$.
+\\
+
+\noindent \textbf{Case fork-join}: Suppose $e_0 = (e_L || e_R)$ and
+$$\delta, e_L \Downarrow \delta_L, v_L, w_L$$
+$$\delta, e_R \Downarrow \delta_R, v_R, w_R$$
+which gives us
+$$\delta, (e_L || e_R) \Downarrow \delta', (v_1 || v_2), w_L + w_R + w'$$
+with $w'$ and $\delta'$ defined in terms of \func{extrawork} and \func{combine} as given in the evaluational dynamics.
+
+In T, for all $0 \leq i \leq n$ let $e_i = (e^L_i || e^R_i)$.
+Let $I_L$ be the sequence of all $0 \leq i \leq n-1$ where transitioning from $e_i$ to $e_{i+1}$ involved stepping the left argument.
+Let $m$ be the length of $I_L$.
+Similarly define $I_R$ for the right argument, and let $k$ be the length of $I_R$.
+
+Let $A$ be the sequence of $e^L$s corresponding the indices in $I_L$, and add $e^L_n$ at the end of $A$.
+Similarly, let $B$ be the sequence of $e^R$s corresponding the indices in $I_R$, adding $e^R_n$ at the end.
+
+Let $S^L_0 = S_0 = sign(\delta)$.
+By inductively applying lemma \ref{stateless_lemma}, there exists $S^L_1, ...
+S^L_m$ s.t.
+$T_L = [(S^L_0, A_0), ..., (S^L_m, A_m)]$ is a transition sequence.
+$T_L$ is maximal, because if $A_m$ in $T_L$ can take a step then $e^L_n$ in $T$ can take the corresponding step.
+Similarly, let  $S^R_0 = S_0 = sign(\delta)$.
+Then, there exists $S^R_1, ...
+S^R_k$ s.t.
+$T_R = [(S^R_0, B_0), ..., (S^R_k, B_k)]$ is a maximal transition sequence.
+
+Applying the IH on $T_L$, we get a relabeling $R_L$ on $L(\delta_L) \setminus L(\delta)$ s.t.
+$R_L(v_L) = A_m = e^L_n$ and $sign(R_L(\delta_L)) = S^L_m$.
+Similarly, we get a relabeling $R_R$ on $L(\delta_R) \setminus L(\delta)$ s.t.
+$R_R(v_R) = B_k = e^R_n$ and $sign(R_R(\delta_R)) = S^R_k$.
+
+By inducting on the product $mk$, we can show that the new labels produced in $T_L$ and $T_R$ do not overlap, that is: $(L(S^L_m) \setminus L(S_0)) \cap (L(S^R_k) \setminus L(S_0)) = \emptyset$.
+Compose the relabelings to get relabeling R.
+
+Applying lemma $\ref{relabeling_lemma}$ and the fork join rule, we get,
+$$\delta, R(e_L) \Downarrow R(\delta_L), R(v_L), w_L$$
+$$\delta, R(e_R) \Downarrow R(\delta_R), R(v_R), w_R$$
+which gives us
+$$\delta, R((e_L || e_R)) \Downarrow R(\delta'), R((v_1 || v_2)), w_L + w_R + w'$$
+
+with $R((v_1||v_2)) = (R(v_1) || R(v_2)) = (e_n^L || e_n^R)$.
+To reduce clutter, WLOG suppose that the stores and expressions in the evaluational dynamics were labeled as such.
+By lemma \ref{relabeling_lemma} we can apply the IH on the relabeled evaluational dynamics since the length of the shortest derivation is invariant under relabelings.
+
+\begin{lemma}
+\label{label_projection_lemma}
+$l \in S_n$ if and only if either $l \in S^L_m$ or $l \in S^R_k$
+\end{lemma}
+
+\begin{proof}
+If $l \in S_0$ the claim holds trivially. Otherwise,
+
+$(\Rightarrow)$ By considering the step in $T$ that first introduced $l$, and examining the corresponding step in either $T_L$ or $T_R$.
+
+$(\Leftarrow)$ By considering the step in $T_L$ (or $T_R$) that first introduced $l$ and examining the corresponding step in $T$.
 \end{proof}
 
-\subsection{Functional Array Implementation}
+\begin{lemma}
+\label{sign_projection_lemma}
+$S_n[l] = -$ if and only if either $l \in S^L_m$ and $S^L_m[l] = -$ or $l \in S^R_k$ and $S^R_k[l] = -$
+\end{lemma}
 
-Suppose we have a FunctionalArray $A$.
-$A$ points to an ArrayData object which is referenced by $A.data$.
-The (imperative) array in $A.data$ is referenced by $A.data.values$, and the log entries corresponding to the $i^{\text{th}}$ index are stored in a PushArray $A.data.undo\_lists[i]$.
-We first explain the implementation of $A.set(pos, val)$.
+\begin{proof}
+Similar to lemma \ref{label_projection_lemma}.
+\end{proof}
 
-We use a link load store conditional to ensure that only 1 thread can modify an ArrayData object at any time.
-If $m$ threads try to modify $A.data$, then $m-1$ threads will fail the LLSC (returning false).
-Instead of modifying $A.data$, these threads will create a new ArrayData object, and will copy over values from $A.data$, as shown below.
+\textbf{Part 1}: We claim that the set of labels in $\delta'$ and $S_n$ are the same.
+From the evaluational dynamics, $L(\delta') = L(\delta_L) \cup L(\delta_R)$.
+Since $sign(\delta_L) = S^L_m$ and $sign(\delta_R) = S^R_k$, $L(\delta_L) = L(S^L_m)$ and $L(\delta_R) = L(S^R_k)$.
+From lemma $\ref{label_projection_lemma}$, $L(S_n) = L(S^L_m) \cup L(S^R_k) = L(\delta')$ which proves the claim.
 
-\begin{lstlisting}[language=Java]
-FunctionalArray newA;
-newA.version = A.version + 1;
-if (!LLSC(&A.data.version, A.version, 
-	newA.version) OR
-    A.version % A.size == 0) {
-  // Create a new ArrayData object AD
-  // Forall i = 1:n, AD.values[i] = get(A, i)
-  AD.values[pos] = val;
-  newA.data = AD;
-  return newA;
-}
-\end{lstlisting}
+To show that $sign(\delta') = S_n$ we case on arbitrary $l \in L(\delta')$ and show that $sign(\delta')[l] = S_n[l]$:
 
-At most one of the threads trying to modify the ArrayData object will get past the LLSC, will insert a log entry and then modify $A.data.values$.
-The ordering of inserting a log entry and modifying $A.data.values$ matters in the proof of get.
-The pseudocode for this step is shown below.
+\textbf{Case $l \in L(\delta)$}:
 
-\begin{lstlisting}[language=Java]
-newA.data = A.data;
-old_value = newA.data.values[pos];
-undo_list = newA.data.undo_lists[pos];
-undo_list.push((A.version, old_value));
-newA.data.values[pos] = val;
-return newA;
-\end{lstlisting}
+If $\delta[l] = (-,\_)$ then $S_0[l] = -$. The evaluational and structural dynamics do not change $-$ to $+$ so $\delta'[l] = (-,\_)$ and $S_n[l] = -$ so $sign(\delta')[l] = S_n[l]$.
 
-Next we explain the implementation of $A.get(pos)$.
+Suppose $\delta[l] = (+, \textunderscore)$, in which case $S_0[l] = +$.
+If $\delta'[l] = (+,\_)$, then from the way combine works, $\delta_L[l] = (+, \_)$ and $\delta_R[l] = (+, \_)$.
+From the IH $S^L_m[l] = +$ and $S^R_k[l] = +$.
+From lemma \ref{sign_projection_lemma} $S_n[l] = + = sign(\delta')[l]$.
 
-\begin{lstlisting}[language=Java]
-guess_val = A.data.values[pos];
-if (A.version == A.data.version) {
-  return guess_val;
-}
+On the other hand, if $\delta[l] = (+, \textunderscore)$ (in which case $S_0[l] = +$) but $\delta'[l] = (-,\_)$, then from the way combine works, either $\delta_L[l] = (-,\_)$ or $\delta_R[l] = (-, \_)$.
+WLOG suppose that $\delta_L[l] = (-,\_)$.
+From the IH, $S^L_m[l] = -$.
+From lemma \ref{sign_projection_lemma} $S_n[l] = - = sign(\delta')[l]$.
 
-undo_list = this.data.undo_lists[pos];
-if (undo_list.size() == 0) {
-  return guess_val;
-}
+\textbf{Case $l \in L(\delta_L), l \not\in L(\delta)$}: Then $l \not\in \delta_R$ so $\delta'[l] = \delta_L[l]$.
+From IH, $sign(\delta_L) = S^L_m$.
+Since $l \not\in \delta_R$, $l \not\in S^R_k$ so by applying lemma \ref{sign_projection_lemma}, $S^L_m[l] = S_n[l]$, as desired.
 
-upper = undo_list.size() - 1;
-if (undo_list[upper].version < A.version) {
-  return guess_val;
-}
-\end{lstlisting}
+The case where $l \in L(\delta_R), l \not\in L(\delta)$ is symmetric.
 
-The last step is to binary search $undo\_list$, between the indices 0 and $upper$ (inclusive of 0 and upper).
-We find the log entry $L$ with the smallest version $X$ such that $A.version \leq X$.
-We return $L.value$.
+Finally, we note that $(v_1||v_2) = (e^L_n||e^R_n)$ from the way $v_1,v_2$ were relabeled.
 
-The implementation of tabulate is straightforward - just allocate a new array and fill the array with the required values.
+\textbf{Part 2}:
 
-\subsection{Proof of Correctness}
+We case on arbitrary $l \in \delta'$.
+
+\textbf{Case $l \in L(\delta)$}:
+
+If $\delta[l] = (-,c)$ then $EC_l(\delta, \delta') = 0 = SC_l(T)$.
+
+Else suppose $\delta_L[l] = (+,c_1)$ and $\delta_R[l] = (+,c_2)$. Then $SC_l(T) = SC_l(T_L) + SC_l(T_R) \geq EC_l(\delta, \delta_L) + EC_l(\delta, \delta_R) = EC_l(\delta, \delta')$.
+
+Else suppose $\delta_L[l] = (+,c_1)$ and $\delta_R[l] = (-,c_2)$.
+We can show that all the cheap \get{}s in $T_R$ are cheap in $T$.
+So $SC_l(T) \geq SC_l(T_R) \geq EC_l(\delta, \delta_R) = EC_l(\delta, \delta)$.
+
+The case where $\delta_L[l] = (-,c_1)$ and $\delta_R[l] = (+,c_2)$ follows by symmetry.
+
+Finally, if $\delta_L[l] = (-,c_1)$ and $\delta_R[l] = (-,c_2)$, then $SC_l(T) \geq 0 = EC_l(\delta, \delta')$.
+
+\textbf{Case $l \in L(\delta_L), l \not\in L(\delta)$} Then $EC_l(\delta, \delta') = EC_l(\delta, \delta_L)$ since $l \not\in \delta_R$.
+By IH, $EC_l(\delta, \delta_L) \leq SC_l(S_0, S^L_m)$.
+Since $l \not\in S^R_k$, $SC_l(S_0, S^L_m) = SC_l(S_0, S_n)$ and so the claim holds.
+
+The case where $l \in L(\delta_R), l \not\in L(\delta)$ is similar.
+
+\textbf{Part 3:} We omit the formal proof of this part because it contains tedious algebraic details but the key ideas are described below.
+
+In the conservation expression, each \get{} on each \farray{} $(l,V)$ in both the evaluational and structural dynamics is charged $g_i(V)$ work (considering the sum from both terms).
+In particular, note that the formula adds an additional $g_i(V)-g_l(V)$ cost for cheap \get{}s.
+Since the number of \get{}s is the same, the total cost of \get{}s is the same in both dynamics.
+
+The number of \set{}s that are charged $s_i(V)$ is the same for both the evaluational and structural dynamics.
+This can be shown by casing on the signs of the \farray{} on both sides of the fork.
+If and only if \set{} was called on a \farray{} on both sides of the fork, one of the \set{}s will cost $s_i(V)$.
+Since the number of \set{}s is the same, the total cost of \set{}s is the same in both dynamics.
+\end{proof}
 
 \begin{theorem}
-Functional Arrays are linearizable.
+(Work Bound) Given the conditions in theorem \ref{main_work_theorem}, $w_E \geq W(T)$
 \end{theorem}
@@
 \begin{proof}
+We assume that $\delta_E$ has been relabeled as described in part 1 of theorem \ref{main_work_theorem}.
+For all $l \in \delta_E$, $EC_l(\delta, \delta_E) \leq SC_l(T)$.
+This implies that $\displaystyle \sum_{l \in \delta_E} EC_l(\delta, \delta_E) (g_i(l)-g_l(l)) \leq \sum_{l \in \delta_E} SC_l(T) (g_i(l)-g_l(l))$.
+But then from the conservation of sum of work and cheap gets, $w_E \geq W(T)$.
+\end{proof}
 
-The linearization point of get is when we compare if A.version and AD.version are the same.
-The linearization point of set is the LLSC.
-Any instruction in tabulate can be its linearization point - we omit discussion of tabulate because the details are uninteresting.
+\begin{theorem}
+(Tightness) Suppose that $\delta, e_0 \Downarrow \delta_E, e_E, w_E, s_E$.
+Then, for some $n$, there exists a transition sequence $T$ from $s_0 = sign(\delta), e_0 \to^n s_n, e_n$ with $2W(T) \geq w_E$.
+\end{theorem}
 
-First, we prove the linearization point of get. (fill this out)
+\begin{proof}
+By induction on the rules of the evaluational dynamics. We present the construction for the most interesting case, fork-join.
+\\
 
-Next, we prove the linearization point of set. (fill this out)
+\noindent\textbf{Case fork-join}: Suppose $e_0 = (e_L || e_R)$ and
+$$\delta, e_L \Downarrow \delta_L, v_L, w_L, s_L$$
+$$\delta, e_R \Downarrow \delta_R, v_R, w_R, s_R$$
+which gives us
+$$\delta, (e_L || e_R) \Downarrow \delta', (v_1 || v_2), w_L + w_R + w'$$
+
+From the IH, there exists transition sequence $T_L$ taking $s^L_0, e^L_0 \to^m s^L_m, e^L_m$ with $s^L_0 = sign(\delta)$, $e^L_0 = e_L$, $e^L_m = v_L$ and $2W(T_L) \geq w_L$.
+Similarly, there exists transition sequence $T_R$ taking $s^R_0, e^R_0 \to^n s^R_n, e^R_n$ with $s^R_0 = sign(\delta)$, $e^R_0 = e_R$, $e^R_n = v_R$ and $2W(T_R) \geq w_R$.
+
+The function \func{cheap-gets} computes the number of gets to a \farray{} on a particular side of the fork.
+$$\text{\func{cheap-gets}}((s, c), (s_{\text{my}}, c_{\text{my}}), (s_{\text{oth}}, c_{\text{oth}})) = s_{\text{oth}} (c_{\text{my}} - c)$$
+$c_L$ represents the extra costs of gets on the left fork that become expensive (because of sets on the right fork).
+Similarly, $c_R$ represents the extra costs of gets on the right fork that become expensive (because of sets on the left fork).
+$$c_L = \sum_{l \in L(\delta), \delta[l \mapsto (+, c)]} \text{\func{cheap-gets}}(\delta[l], \delta_L[l], \delta_R[l]) (g_i(l) - g_l(l))$$
+$$c_R = \sum_{l \in L(\delta), \delta[l \mapsto (+, c)]} \text{\func{cheap-gets}}(\delta[l], \delta_R[l], \delta_L[l]) (g_i(l) - g_l(l))$$
+The function \func{double-sets} computes whether a \farray{} was set on both sides of the fork-join.
+$$\text{\func{double-sets}}((s_L, c_L), (s_R, c_R)) = s_L s_R$$
+$$ss = \sum_{l \in L(\delta), \delta[l \mapsto (+, c)]} \text{\func{double-sets}}(\delta_L[l], \delta_R[l])s_i(l)$$
+
+If $c_R \geq c_L$ then we step the left side of the fork to completion before the right side of the fork.
+If $c_L > c_R$ then we step the right side of the fork to completion before the left side.
+We can then show that the resulting transition sequence $T$ satisfies $2W(T) \geq w_E$.
+
+WLOG suppose $c_R \geq c_L$.
+In the evaluational dynamics, $w' = c_L + c_R + ss$.
+Because we step the left side of the fork before the right side, and $c_R \geq c_L$, $2W(T) \geq 2(W(T_L) + W(T_R) + c_R + ss) \geq 2W(T_L) + 2W(T_R) + c_R + c_L + ss \geq w_L + w_R + c_L + c_R + ss \geq w_L + w_R + w'$.
 
@@
 
-Linearizability allows us to assume that array operations happen atomically.
-Note that all non array operations in our source language are executed in a single time step, so any well formed program is linearizable.
-In fact, since linearizability composes, we can show that more complicated languages that use our arrays are also linearizable.
-
-In a previous section, we gave the evaluational (big step) dynamics of functional arrays.
-Alternatively, we could give the non-deterministic structural (little step) dynamics which describes the steps in the execution of a program in the source language.
-At each fork-join, if both sides of the fork can take a step, we could take a step on either side.
-
 \begin{theorem}
-The non-deterministic structural dynamics and the evaluational dynamics produce the same result in the source language.
+(Conditional termination) Suppose that $\delta, e_0 \Downarrow \delta_E, e_E, w_E, s_E$. Then there exists $n$ s.t. all transition sequences starting at $sign(\delta), e_0$ have length $\leq n$.
 \end{theorem}
@@
 \begin{proof}
-This is a standard implicit parallelism type theorem, for example see Chapter 39 in~\cite{harper}, so we omit the details. 
+By induction on the rules of the evaluational dynamics.
 \end{proof}
 
-Now, to show that our implementation correctly implements the source language, it suffices to show that any valid execution of our implementation produces the same result as a little step transition sequence in the source language (since all transition sequences produce the same value as the evaluational dynamics).
-
-\begin{theorem}
-For any program $P$ in the source language, executing our implementation $I$ of $P$ produces the same result as some transition sequence of $P$. Note that $I$ is a program in the target language.
-\end{theorem}
-
-\begin{proof}
-We prove the claim by induction. We focus on the part of the proof dealing with arrays and give a sketch of the rest.
-
-Our inductive hypothesis is that after $t$ time steps in an arbitrary execution of $I$, there exists a transition sequence $S$ s.t.:
-\begin{enumerate}
-\item $I$ becomes $I'$ after this particular execution, $S$ takes $P$ to $P'$
-\item $I'$ and $P'$ are equivalent in the following sense.
-Any non array value is identical in $I'$ and $P'$.
-For any abstract array $A$ in $P'$ there exists a corresponding representation $A_I$ s.t.
-the following holds.
-Consider arbitrary index $i$ and suppose that $A[i]$ has value $val$.
-Suppose that $A_I$ has version $V$ and is pointing to an ArrayData object $AD$.
-If there exists some version in $AD.logs[i]$ which is greater than $V$ then there exists a log entry $(v', val)$ where $v'$ is the smallest version in $AD.logs[i]$ that is $\geq V$.
-If no version in $AD.logs[i]$ is greater than $V$, then $AD.values[i]$ is val.
-\end{enumerate}
-
-The base case holds since $P$ and $I$ are initially identical.
-For the inductive step, when $I'$ takes a step, we take the corresponding step in $P$ to get a new transition sequence $S'$.
-We know that such a step exists because $I'$ and $P'$ are equivalent.
-If the step is not an array function then the IH holds since the step will be executed in the same way in the source and target languages.
-
-Suppose that the next step is tabulate.
-By the IH, tabulate is called with the same arguments $n$ and $f$ in the source language and the implementation.
-In the source language, tabulate returns an array $A$ with $A[i] = f(i)$ for all $1 \leq i \leq n$.
-In the implementation, tabulate returns an array that points to a new ArrayData object $AD$ with empty logs and $AD.values[i] = f(i)$ for all $1 \leq i \leq n$.
-So the inductive step holds.
-
-Suppose that the next step is get.
-By the IH, get is called on equivalent arrays and the same index $i$ in the source language and implementation.
-It is easy to trace through the implementation of get and see that get returns the same value in the source language and implementation.
-
-Suppose that the next step is set.
-Suppose that in the target language we execute $set(A_I, i, v)$.
-Then by IH, in the source language the corresponding step is $set(A, i, v)$ where $A$ and $A_I$ are equivalent.
-Denote $set(A, i, v)$ by $B$ and $set(A_I, i, v)$ by $B_I$.
-We need to show that $B$ and $B_I$ are equivalent, and all arrays that were previously equivalent are still equivalent.
-
-Case 1: we create a new ArrayData object $AD$ in set.
-In this case, for all $j \neq i$ we set $AD.values[j]$ to $get(A_I, j)$ which by the IH is the same as $A[j]$.
-Since we only modified index $i$, $j \neq i \Rightarrow A[j] = B[j]$.
-Also, we set $AD.values[i]$ to $v$, so $AD.values[i] = B[i]$.
-Since all logs are empty, this means that $B$ and $B_I$ are equivalent.
-Since we create a new ArrayData object, we do not modify any other arrays, so arrays that were previously equivalent are still equivalent.
-
-Case 2: we modify the ArrayData object $AD$ that $A_I$ points to.
-This means that $A.version = AD.version$.
-We can show that all log entries in $AD$ have version $< AD.version$, and are in increasing order.
-Then the IH implies that before modifying $AD$, $AD.values[j] = A[j]$ for all indices $j$.
-So after setting $AD.values[i]$ to $v$, $AD.values[j] = B[j]$ for all $j$, which means $B_I$ and $B$ are equivalent.
-
-Consider arbitrary abstract array $C$ in the source language and corresponding representation $C_I$ in the implementation with $C_I \neq B_I$.
-If $C_I$ did not point to $AD$ then $C_I$ remains equivalent to $C$.
-Regardless, we note that for all indices $j \neq i$, $C_I$ remains the same and is equivalent to $C$.
-At index $i$, if the value at $C[i]$ was previously stored in a log entry, then $C_I$ and $C$ remain equivalent since log entries are added in increasing version numbers.
-If the value at $C[i]$ was stored in $AD[i]$ then we add a log entry at index $i$ so $C_I$ and $C$ remain equivalent.
-
-\end{proof}
-
-\section{Machine Model}
-
-Our target language is a Java-like language augmented with tabulate, fork-join, lambda functions, and link load store conditional (LLSC).
-Fork-join spawns two threads, one thread executing each sub-expression.
-Link load store conditional (LLSC) takes an (address, old value, new value) tuple as argument.
-LLSC first checks that the value at address is the same as old value, if it is different then LLSC returns false.
-Otherwise LLSC atomically stores new value at address and returns true, however if the value at address was modified between the load and store, LLSC returns false.
-
-Our target language is run on a P processor machine.
-In a single time step the machine takes at most P runnable instructions and executes them.
-The effect of executing the ($\leq P$) instructions in a time step is the same as some (arbitrary) sequential ordering of the instructions.
-We assume that the machine does not reorder instructions in the target language.
-In reality, languages like Java have very relaxed consistency models, and we deal with this in our real implementation using memory fences.
-
-The link load store conditional architecture deserves special mention.
-Call an LLSC operation pending if the load operation has completed but the store operation has not.
-Multiple threads can load the value at a particular address and compare it with old value in a single time step.
-Only a single store operation to a particular address can be executed at a time step.
-However, after the store operation, in each time step the bus arbiter notifies all pending LLSC operations to the same memory address that they have failed.
-This assumption is reasonable because all LLSC instructions are typically processed by the bus arbiter, so the bus arbiter can keep track of pending LLSC instructions to each memory address.
-
-% We recommend abbrvnat bibliography style.
-
-\bibliographystyle{abbrvnat}
-
-% The bibliography should be embedded for final submission.
-
-\bibliography{references}
-
-\end{document}
+When evaluating an expression $e$ in our language, the stores are initially empty.
+In particular, the signs of the store in the evaluational dynamics and structural dynamics are the same.
+From the tightness theorem, we know there exists a transition sequence $T$ starting at $e$.
+From the conditional termination theorem, all transition sequences starting at $e$ have bounded length.
+Then, from the work bound theorem, the work computed by the evaluational dynamics is at least the work of any transition sequence.
+Since the structural dynamics captures the (non-deterministic) execution time of the expression, this means that the evaluational dynamics gives an upper bound for the execution time of the expression.

### 2016-09-25 Paper/dynamics.tex
Reorganize repository

\section{Language}
\label{sec:dynamics}

The notation we use for our language and dynamics is from ~\cite{pfpl2}.

We use a standard applicative-order, call by value source language defined as follows:
$$e = x \; | \; c  \; | \; \lambda (x,y).e \; 
| \; e_1 e_2 \; 
| \; (e_1, e_2) \; 
| \; (e_1~||~e_2) \; 
| \; \text{if } e_1 \; e_2 \; e_3$$

The constants $c$ contains the usual arithmetic types, such as the
natural numbers and numerical functions. The expression $(e_1||e_2)$
indicates that the two expressions can run in parallel, returning a
pair $(v_1, v_2)$ when both expressions are fully evaluated, while
$(e_1,e_2)$ generates a pair sequentially.  Sequential and parallel
pairs only differ in the cost dynamics---the value dynamics are
identical. Our language can be augmented to add support for recursion using \func{letrec} or a fixed point combinator.

\begin{definition}
The terminal (fully evaluated) values are lambda expressions, constants,
and pairs of terminal values, and are denoted by the judgement \func{val}.
\end{definition}

Our language also contains functions to work with sequences. \new{}$(n,v)$ evaluates to a sequence of size $n$ with the value $v$ at each index. \get{}$(A, i)$ evaluates to the $i^{\text{th}}$ element of sequence $A$. \set{}$(A, i, v)$ evaluates to a new sequence where the $i^{\text{th}}$ element of $A$ is substituted with $v$.

### 2016-09-25 Paper/evaluational.tex
Reorganize repository

\section{Evaluational Cost Dynamics}
\label{sec:evaluational}

The evaluational (big-step) cost dynamics gives an easy, deterministic way to compute the cost of a program without having to reason about multiple interleavings. In this section, we give the rules for the evaluational dynamics. In section~\ref{sec:costproof}, we prove that the costs in the evaluational dynamics are a tight upper bound for the costs in the structural dynamics. Our results can be easily extended for data types besides sequences (like unordered sets) even if there are multiple varieties of \get{} and \set{} functions.

To handle the non-determinism in fork-join, we first compute the worst-case cost of each fork separately. Then for sequences that are used in both forks, we add additional work at the join point. To compute the additional work at the join point, we keep track of the number of \get{}s to leaf values on each side of the fork. If the value was \set{} on the other side of the fork, then the \set{} might have happened before the \get{}s, so we need to charge additional work for the \get{}s.

Our evaluational dynamics is defined by the following judgement, where $\delta$ is the store, $e$ is the expression, $\delta'$ is the new store, $v$ is the value that $e$ evaluates to, $w$ is the work, and $s$ is the span.
$$\delta, e \boldsymbol\Downarrow \delta', v, w, s$$

As in the previous section, sequences are represented by $(l, V)$ where $l$ is a label into the store $\delta$ and $V$ is a list of the elements in the sequence. The store is a mapping from labels to $(+/-, c)$, where $c$ represents the number of leaf \get{}s on the value indexed by the label. $L(\delta)$ denotes the set of labels in the store $\delta$.

The rules in the evaluational dynamics are given in figure \ref{fig:evaluational-dynamics}. Note that \get{} costs $g_l(V)$ instead of $g_i(V)$ on a leaf value $V$ but we increment the counter of leaf \get{}s in the store.

\begin{figure}[!ht]
$$\frac{}{\delta, c \Downarrow \delta, c, 1, 1} \text{ (constants)}$$

$$\frac{\begin{gathered}\delta, e_1 \Downarrow \delta', \lambda (x,y) . e, w_1 , s_1 \;\;\; \delta', e_2 \Downarrow \delta'',(v_1, v_2), w_2 , s_2 \;\;\; \\ \delta'', [v_1/x][v_2/y]e \Downarrow \delta''', v', w_3 , s_3\end{gathered}}{\delta, e_1 \; e_2 \Downarrow \delta''', v', 1+w_1+w_2+w_3, 1+s_1+s_2+s_3} \text{ (func-app)}$$

$$\frac{\delta, e_1 \Downarrow \delta', true, w_1, s_1 \;\;\, \delta', e_2 \Downarrow \delta'', v, w_2, s_2}{\delta, \text{if } e_1 \; e_2 \; e_3 \Downarrow \delta'', v, 1+w_1+w_2, 1+s_1+s_2} \text{ (if-true)}$$

$$\frac{\delta, e_1 \Downarrow \delta', false, w_1, s_1 \;\;\; \delta', e_3 \Downarrow \delta'', v, w_2, s_2}{\delta, \text{if } e_1 \; e_2 \; e_3 \Downarrow \delta'', v, 1+w_1+w_2, 1+s_1+s_2} \text{ (if-false)}$$

$$\frac{\delta, e \Downarrow \delta', a, w, s \;\;\; l \not\in L(\delta')}{\begin{gathered}\delta, \tnew{}(e) \Downarrow \delta'[l \mapsto (+,0)], \overline{\mbox{\new{}}}(a), w + n(a), s+1\end{gathered}} \text{ (new)}$$

$$\frac{A = (l,V) \;\;\; a \text{ val} \;\;\; \delta[l \mapsto (+, c)] \;\;\; l' \not\in L(\delta)}{\begin{gathered}\delta, \tset(A, a) \Downarrow \delta[l \mapsto (-, c), l' \mapsto (+, 0)], \\ (l', \overline{\mbox{\set{}}}(V, a)), s_l(V), 1\end{gathered}} \text{  (set-leaf)}$$

$$\frac{A = (l,V) \;\;\; a \text{ val} \;\;\; \delta[l \mapsto (-, c)] \;\;\;  l' \not\in L(\delta)}{\begin{gathered}\delta, \tset(A, a) \Downarrow \delta[l' \mapsto (+, 0)], \\ (l', \overline{\mbox{\set{}}}(V, a)), s_i(V), 1\end{gathered}} \text{  (set-int)}$$

$$\frac{\begin{gathered}\delta; e_1 \Downarrow \delta_1, A, w_1, s_1 \;\;\; \delta_1, e_2 \Downarrow \delta_2, a, w_2, s_2 \;\;\; \\ \delta_2, \tset(A, a) \Downarrow \delta', A', w', s' \end{gathered}}{\delta, \tset(e_1, e_2) \Downarrow \delta', A', w_1+w_2+w', s_1+s_2+s'} \text{ (set-eval)}$$

$$\frac{A = (l,V) \;\;\; a \text{ val} \;\;\; \delta[l \mapsto (+, c)]}{\begin{gathered}\delta, \tget{}(A, a) \Downarrow \delta[l \mapsto (+, c+1)], \\ \overline{\mbox{\get{}}}(V, a), g_l(V), 1\end{gathered}} \text{  (get-leaf)}$$

$$\frac{A = (l,V) \;\;\; a \text{ val} \;\;\; \delta[l \mapsto (-, c)]}{\delta, \tget{}(A, a) \Downarrow \delta, \overline{\mbox{\get{}}}(V, a), g_i(V), 1} \text{  (get-interior)}$$

$$\frac{\begin{gathered}\delta, e_1 \Downarrow \delta_1, A, w_1, s_1 \;\;\; \delta_1, e_2 \Downarrow \delta_2, a, w_2, s_2 \;\;\; \\ \delta_2, \tget{}(A, a) \Downarrow \delta', v', w', s' \end{gathered}}{\delta, \tget(e_1, e_2) \Downarrow \delta', v', w_1 + w_2 + w', s_1 + s_2 + s'}\text{ (get-eval)}$$

$$\frac{\begin{gathered}\delta, e_L \Downarrow \delta_L, v_L, w_L, s_L \;\;\; \delta, e_R \Downarrow \delta_R, v_R, w_R, s_R \;\;\; \\ (L(\delta_L) \setminus L(\delta)) \cap (L(\delta_R) \setminus L(\delta)) = \emptyset\end{gathered}}{\delta, (e_L || e_R) \Downarrow \delta', (v_L, v_R), 1 + w_L + w_R + w', 1 + \max(s_L, s_R)} \text{  (fj)}$$

\caption{Rules for the evaluational dynamics. $w'$ and $\delta'$ in the fork-join rule are defined in the paper.}
\label{fig:evaluational-dynamics}
\end{figure}

The fork-join rule is the most interesting and requires a few definitions. Suppose we have expression $(e_L || e_R)$ with
\[ \delta, e_L \Downarrow \delta_L, v_L, w_L, s_L \]
\[ \delta, e_R \Downarrow \delta_R, v_R, w_R, s_R \]

Further, suppose that $(L(\delta_L) \setminus L(\delta)) \cap (L(\delta_R) \setminus L(\delta)) = \emptyset$ (the new labels produced on both sides of the fork-join do not conflict). Consider a sequence $A = (l, V)$ with $\delta[l] = (s,c)$, $\delta_L[l] = (s_L, c + c_L)$, $\delta_R[l] = (s_R, c + c_R)$. When multiplying two signs or multiplying a sign with an integer, consider $+$ to be 1 and $-$ to be 0.

\func{Combine} describes how to combine store values on both sides of a fork-join. A sequence is a leaf iff it is a leaf on both sides of the fork-join. Leaf \get{}s on one side of the fork remain leaf \get{}s iff there were no calls to \set{} on the other side of the fork.
\begin{equation*}
  \begin{aligned}
   \text{\func{combine}}&((s, c), (s_L, c + c_L), (s_R, c + c_R))= \\
   &(s_L s_R, c + s_R c_L + s_L c_R)
  \end{aligned}
\end{equation*}
$$\delta' = \; (\delta_L / \delta) \; \cup \; (\delta_R / \delta) \; \cup \bigcup_{l \in L(\delta)} [l \mapsto \text{\func{combine}}(\delta[l], \delta_L[l], \delta_R[l])]$$
\func{Extrawork} gives the additional cost incurred if a sequence was modified on either side of a fork-join. If the value was interior before the fork-join, then all functions incurred their maximal cost and there is no additional cost. Otherwise, leaf \get{}s on one side of the fork are charged $g_i$ work iff the other side of the fork called \set{}. Additionally, if both sides of the fork called \set{}, then one of the \set{}s came first and has work $s_l$ and the subsequent \set{} has work $s_i$. We abuse notation so that $g_l, g_i, s_l, s_i$ directly take in a label $l$ instead of the corresponding sequence.
\begin{equation*}
  \begin{aligned}
  \text{\func{extra}}&\text{\func{work}}(l, (s, c), (s_L, c + c_L), (s_R, c + c_R)) = \\
  & ((\neg s_R) c_L  + (\neg s_L) c_R)(g_i(l)-g_l(l)) + \\
  & (\neg s_R) (\neg s_L) (s_i(l)-s_l(l))
 \end{aligned}
\end{equation*}
$$w' = \sum_{l \in L(\delta), \delta[l \mapsto (+, c)]} \text{\func{extrawork}}(l, \delta[l], \delta_L[l], \delta_R[l])$$
Then, the fork-join cost dynamics are:
$$\delta; (e_L || e_R) \Downarrow \delta'; (v_L || v_R); 1 + w_L + w_R + w'; 1 + \max(s_L, s_R)$$

### 2016-09-25 Paper/experiments.tex
Reorganize repository

\section{Test Results}
\label{sec:experiments}

Besides implementing our functional arrays in SML, we implemented our arrays in Java and compared the performance of functional arrays with regular Java arrays. Java has a relaxed memory consistency model (not a sequentially consistent memory model), so we added memory fences in suitable locations to prevent the compiler and machine from reordering instructions.

We compared the performance of leaf functional arrays and regular Java arrays of size 3,000,000 on a dual-core machine. The arrays occupied 12mb of space, and the machine had 3mb of shared L3 cache, so we ensured that the entire arrays cannot be cached. Since many of our tests involved looping and accessing elements in the array, without performing any useful computations, we disabled compiler optimizations (which optimize away the loops).

We ran each test 5 times, and took the average time. We do not show standard errors because we were interested in orders of magnitude, however all timings differed from the mean by at most $15\%$ of the mean time.

\begin{table}[h!]
\centering
\begin{tabular}{ |c|c|c|c| } 
 \hline
  & Regular & Functional & Slowdown \\ 
 \hline
 Seq. reading array & 1.35s & 1.50s & 10.8\% \\ 
 \hline
 Rand. reading array & 8.88s & 9.10s & 3.4\% \\ 
 \hline
 Seq. writing to array & 2.78s & 9.14s & 3.3$\times$ \\ 
 \hline
 Rand. writing to array & 5.57s & 13.2s & 2.4$\times$ \\ 
 \hline
\end{tabular}
\caption{Speed of leaf functional arrays vs. regular arrays in Java}
\label{mytable}
\end{table}

In the sequential read test, after creating an empty array, we sequentially get elements at indices $0,1,2,...$, starting over at $0$ when we reach the end of the array. In the random read test, we repeatedly generate a random number $r$ and get the $r^{\text{th}}$ element of the array. We performed 15,000,000 accesses in both of these tests.

In the sequential write test, starting from an empty array, we sequentially set the elements at indices $0,1,2,...$, starting over at $0$ when we reach the end of the array. In the random write test, we repeatedly generate a random number $r$ and set the $r^{\text{th}}$ element of the array. We performed 5,000,000 writes in both of these tests.

The results suggest that operations on leaf functional arrays are almost as efficient as regular arrays. The additional 2-3 times slowdown in \set{} is expected because we incur an additional cache miss when we insert a log entry. Note that similar benchmarks on alternative implementations of functional arrays, for example persistent binary search trees, are likely to be slower by a much larger factor.

Additionally, we compared the time taken to access elements in leaf arrays and interior arrays in a specific benchmark. We wrote 20,000,000 values into random indices of an array of size 2,100,000. We then read 5,000,000 values from random indices of the leaf array and the interior array. Reading from the interior array was $4.5 \times$ slower, which is not too much of a slowdown.

We also performed two simple tests to profile multi-threaded accesses in our functional array implementation. In the first test, 2 threads simultaneously accessed 500,000 random elements in an array. The total time taken was $1.77 \times$ less than a single thread accessing 1,000,000 random elements in the array. In the second test, 2 threads simultaneously accessed the same element 500,000 times. The total time taken was $1.76 \times$ less than a single thread accessing the element 1,000,000 times. The results of the second test are particularly good. We cannot expect a $2\times$ speedup because the element will keep moving between the L1 caches of the two cores. However, the speedup of $1.76 \times$ means that the accesses are not serialized (as they would be with a per-element lock).

### 2016-09-25 Paper/implementations.tex
Reorganize repository

\section{Implementation}
\label{sec:implementations}

In this section we give our implementation of functional arrays (sequences). We show the correctness of our implementation. We show that \get{} and \set{} on leaf sequences require constant work, and \get{} and \set{} on interior sequences require bounded work. As such, our implementation matches the structural dynamics of a versioned data structure, and we can use the evaluational dynamics we derived to analyze costs of parallel programs involving sequences.

We give SML-like code for our implementation. In our implementation we construct and manipulate mutable arrays (henceforth called arrays). We assume that an array of size $n$ can be created with $n$ work on a single processor machine and $\log{n}$ work on a machine with an unbounded number of processors. new$(n, v)$ creates a new array of size $n$ with $v$ at each index. tabulate$(n, f)$ creates a new array of size $n$ which for all $i$ contains $f(i)$ at the $i^{\text{th}}$ index. sub$(A,i)$ evaluates to the $i^{\text{th}}$ element of array $A$ with constant work. update$(A, i, v)$ mutates index $i$ of array $A$ to have value $v$ with constant work.

Our implementation assumes an atomic compare and swap function \func{cmpswap}. \func{cmpswap}$(a$ : int ref, $v$ : int, $v'$ : int$)$ : bool is a single atomic operation. It compares the value at $a$ to $v$, and if and only if they are the same sets the value at $a$ to $v'$ and returns true, otherwise it returns false.

We assume a $p$-processor parallel machine model. For our proofs of correctness, we assume a sequentially consistent memory model. When analyzing costs, we assume that all processors are synchronized with respect to a global clock. At each time step, as many processors as possible execute a single pending instruction.

\subsection{Pusharray Implementation}

We use an auxiliary data structure, pusharrays, to store log entries. Conceptually, pusharrays are lists that support 3 functions. Suppose we have a pusharray $A$. $\tpush{}(A,e)$ inserts entry $e$, $\tsize{}(A)$ returns the number of entries inserted, $\tget(A,i)$ returns the $i^{\text{th}}$ entry inserted. Note that \get{} is only defined between indices 0 and $\tsize{}(A)-1$. All operations have amortized constant work.

Pusharrays can be used semi-concurrently. At most one thread can execute instructions in \push{} at any time, but multiple threads can call \size{} and \get{}. The type definition for pusharrays is given below.

\begin{lstlisting}[language=ML]
type capacity = int
type size = int
type 'a pusharray = (capacity ** size 
                              ** 'a array) ref
\end{lstlisting}

The pusharray's array stores the entries inserted into the pusharray. Size represents the number of entries inserted and capacity represents the size of the array. If the pusharray does not have the capacity for more insertions, it copies all entries to an array that is twice as large. The implementation for \push{} is given below.

\begin{lstlisting}[language=ML]
val push : 'a pusharray ** 'a -> unit
fun push$(A \mbox{ as } \mbox{ref}(c,s,D), v)$ =
  if $c = s$ then 
    let val $c' = 2 \times c$
       val $D'$ = Array.new$(c', v)$
    in copyArray$(D,D')$;
       update$(D',s,v)$;
       A := $(c',s+1,D')$
    end
  else 
    update$(D, s, v)$;
    A := $(c, s+1, D)$
\end{lstlisting}

\size{} simply returns the size attribute of the pusharray, and \get{}$(A,i)$ returns the $i^{\text{th}}$ index in the pusharray's data array. Intuitively, the reason that pusharrays are needed for our sequence implementation is that they are linearizable ~\cite{linearizability} given that different calls to \push{} (on the same pusharray) do not overlap.

\subsection{Sequence Implementation}
Our sequence implementation keeps a value array of the most recent values, which represents the values at a leaf node of the version tree. For each index of the array we store a change-log that keeps track of values at interior nodes of the version tree. The type definition for sequences is given below.

\begin{lstlisting}[language=ML]
type version = int
type 'a logs = ('a pusharray) array
type 'a arraydata = version ref ** 'a array 
                                ** 'a logs
type 'a sequence = version ** 'a arraydata
\end{lstlisting}

Note that multiple sequences can reference the same arraydata. Figure \ref{fig:new_array_A} visualizes a newly created sequence $A$.

\begin{figure}[!ht]
\centering
\includegraphics[scale=0.3]{new_array_A}
\nocaptionrule \caption{A = \new{}(5, 0)}
\label{fig:new_array_A}
\end{figure}

The implementation of \new{} is given below.

\begin{lstlisting}[language=ML]
val new : int ** 'a -> 'a sequence
fun new(size, init) =
  (1, (ref 1, Array.new(size, init), 
    tabulate(size, 
      fn _ => Pusharray.new$(0, \mbox{init})$)))
\end{lstlisting}

\set{} uses a compare and swap to increment the arraydata's version so that only one thread can modify an arraydata at any point in time. If the compare and swap is successful (which means the sequence is a leaf sequence), a log entry is inserted and the value array is mutated. Otherwise, the values in the sequence are copied over to produce a new arraydata. The new arraydata will have empty logs at each index. Note that the values are not directly copied into the new arraydata from the value array, we need to call \get{} at each index of the sequence when getting the values.

The ordering of the instructions in \set{} is critical. If the vals array is modified before the log entry is inserted then a \get{} evaluated between the 2 instructions could evaluate to the wrong value.

Figure \ref{fig:set_old_array} visualizes \set{}s on interior and leaf sequences.
\begin{figure}[!ht]
\centering
\includegraphics[scale=0.3]{set_old_array}
\nocaptionrule \caption{$B = \tset(A, 2, 5)$ changes the value array and adds a log entry. Then $D = \tset(A, 0, 9)$ creates $AD'$ because $A$ is interior.}
\label{fig:set_old_array}
\end{figure}

The implementation of \set{} is given below.

\begin{lstlisting}[language=ML,escapechar=|]
val set : 'a sequence ** int ** 'a -> 'a sequence
fun set$((V,(V_r, A, L)),i,v)$ = 
  if not(cmpswap$(V_r, V, (V+1))$)|\label{line:set-cmpswap}| orelse
     (!$V_r \bmod \mbox{length}(A) = 0$) then
    let val $A'$ = getFarrayValues$((V,(V_r, A, L)))$
       val $L'$ = emptyLogs(size($A$))
    in update$(A', i, v)$; $(1, (\mbox{ref }1, A', L')$ 
    end
  else
    Pusharray.push$(\mbox{sub}(L,i),(V,\mbox{sub}(A,i)))$; |\label{line:start-cmpswap}|
    update$(A, i, v)$;
    $(V+1, (V_r, A, L))$ |\label{line:end-cmpswap}|
\end{lstlisting}

\get{}$(A,i)$ first loads the value $v$ at the $i^{\text{th}}$ index of the value array. If the sequences' version and corresponding arraydata's version match (meaning that the sequence is a leaf sequence), then \get{} evaluates to $v$. If the log at index $i$ is empty or all versions in the log at index $i$ are less than the sequence version, then \get{} evaluates to $v$. Otherwise, \get{} binary searches the log at index $i$ for the value corresponding to the least version $\geq$ the sequence's version.

As in \set{}, the ordering of the instructions in \get{} is important. In particular, index $i$ of the value array must be loaded before the versions are compared or the logs are examined. If the versions are compared first and a \set{} is evaluated between the 2 instructions, the \set{} might modify the value array and cause the \get{} to evaluate to the wrong value.

The implementation of \get{} is given below.

\begin{lstlisting}[language=ML,escapechar=|]
val get : 'a sequence ** i -> 'a
fun get$((V, (V_r, A, L)),i)$ =
let val guess = sub$(A,i)$ |\label{line:load-value}|
   val $l$ = sub$(L, i)$
   val $n$ = Pusharray.size$(l)$
in if $V = \;!V_r$ then guess |\label{line:get-version-check}|
   else if $n = 0$ then guess |\label{line:log-empty}|
   else
      let val $(V', \_)$ = Pusharray.get$(l, n-1)$
      in if $V' < V$ then guess |\label{line:log-small}|
        else valueAtLeastVersionGeq$(l,V)$ |\label{line:binary-search}|
      end
\end{lstlisting}

\subsection{Target Language}

For our proofs we give a more formal definition of the target language in which we implement sequences\footnote{Our target language is different from the SML-like code used in the code samples. This for clarity of our code samples. It is easy to translate our code samples to our target language.}. The target language extends the source language to support mutable references and compare and swap. The structural dynamics of the target language is given by the following judgement, where $(\pi, \mu)$ is the memory and $(\pi', \mu')$ is the memory after the step has been taken,
$$(\pi, \mu), e \to_T (\pi', \mu'), e'$$

For simplicity we assume all sequences and arraydatas are int sequences and int arraydatas. To avoid dealing with overflow issues, we assume that all integers in the target language are unbounded. Sequences are represented as $\&l$ where the label $l$ indexes into the store $\pi$.

The memory contains an immutable store $\pi$ that maps labels to $(\lambda, *k)$ where $\lambda$ is an integer version number and $k$ is an index into $\mu$. Intuitively, $\pi$ contains all the sequences created so far in the evaluation of the expression. New key-value pairs can be inserted into $\pi$ but existing mappings cannot be modified. The memory also contains a mutable store $\mu$ that maps labels to values $(V, A, L)$ of type int arraydata.

For concision we do not give the full dynamics of our target language. As examples, we give the rules for creating a new arraydata, pushing a log entry, and compare and swap. Let $a_n$ denote an array of $n$ $a$s and let $\{\}$ denote an empty $(\mbox{int} * \mbox{int})$ pusharray.
$$\frac{n \text{ int val} \;\;\; k \not\in L(\mu)}{(\pi, \mu), \mbox{\func{newad}}(n) \to_T (\pi, \mu[k \mapsto (0, 0_n, \{\}_n)]), *k}\text{ (new-ad)}$$

$$\frac{V, V' \text{ int val} \;\;\; \mu[k] = (V, A, L)}{(\pi, \mu), \mbox{\func{cmpswap}}(*k, V, V') \to_T (\pi, \mu[k \mapsto (V', A, L)]), \text{true}}\text{ (cas-true)}$$

$$\frac{V, V' \text{ int val} \;\;\; \mu[k] \neq (V, A, L)}{(\pi,\mu), \text{\func{cmpswap}}(*k, V, V') \to_T (\pi,\mu), \text{false}}\text{ (cas-false)}$$

$$\frac{\begin{gathered}\mu[k] = (V, A, L) \;\;\; i, V', v \text{ int val} \;\;\; \\ L[i] = (p_1, ..., p_n) \;\;\; L' = L[i \mapsto (p_1, ..., p_n, (V',v))]\end{gathered}}{(\pi,\mu), \text{\func{push}}(*k, i, (V', v)) \to (\pi,\mu[k \mapsto (V, A, L')]), *k} \text{ (push)}$$

In the source language, evaluating \new{}, \get{}, and \set{} involves a single step. In the target language, \new{}, \get{}, and \set{} are function calls to our implementation. Evaluating the function calls requires multiple steps. Let $C(\tnew{})$, $C(\tget{})$, $C(\tset{})$ be the implementation (that is, the lambda expression) for \new{}, \get{}, and \set{} respectively.

The following rules show the evaluation of \get{} and \set{} in the target. \get{} is simply substituted with its implementation. For \set{}, we use a \set{}' ``wrapper" so that the sequence \set{} evaluates to can be added to the store $\pi$ in the end-set rule. We keep track of the sequences created so we can show invariants on sequences in our proof of correctness.

$$\frac{i \text{ int val}}{(\pi, \mu), \tget{}(\&l, i) \to_T (\pi, \mu), C(\tget{})(\&l, i)} \text{ (get)}$$

$$\frac{i, v \text{ int val}}{(\pi,\mu), \text{\set{}}(\&l, i, v) \to_T (\pi, \mu), \text{\set{}}'(C(\text{\set{}})(\&l, i, v))} \text{ (start-set)}$$

$$\frac{(\pi, \mu), e \to_T (\pi',\mu'), e'}{(\pi,\mu), \text{\set{}}'(e) \to_T (\pi',\mu'), \text{\set{}}'(e')} \text{ (eval-set)}$$

$$\frac{l \not\in L(\pi)}{(\pi, \mu), \text{\set{}}'((V,*k)) \to_T (\pi[l \mapsto (V, *k)], \mu), \&l} \text{ (end-set)}$$

The rules for \new{} are similar to \set{}. As in the source structural dynamics, either side of a fork-join expression in the target can take a step. This makes the target structural dynamics non-deterministic. However, because \get{} and \set{} involve multiple steps, the target has a greater number of possible interleavings than the source.

\begin{definition}
A transition sequence in the target language is a sequence of memory stores $[(\pi_0, \mu_0), ..., (\pi_n, \mu_n)]$ and expressions $[e_0, ..., e_n]$ such that for all $0 \leq i < n$, $(\pi_i, \mu_i), e_i \to_T (\pi_{i+1}, \mu_{i+1}), e_{i+1}$. We say that $(\pi_0, \mu_0), e_0 \to^*_T (\pi_n, \mu_n), e_n$ or $(\pi_0, \mu_0), e_0, \to^n_T (\pi_n, \mu_n), e_n$.
\end{definition}

An expression $e$ in the source language is compiled to an expression $e'$ in the target and then executed according to the target's dynamics. Since the source language is a subset of the target language, $e = e'$. In other words an expression $e$ in the source language can be viewed as an expression in the target.

We show that our implementation of sequences in the target correctly implements the source. The non-atomicity of \set{} and \get{} means that we cannot appeal to standard parametricity theorems for our proofs.

\subsection{Proof of Correctness}

The full correctness proof follows along the lines of~\cite{concurrent-logical} and~\cite{concurrent-logical2}. We concentrate here on the critical parts, involving \new{}, \get{}, and \set{}. We give a rely-guarantee style argument~\cite{rely-guarantee} for the correctness of our implementation (for integer sequences).

We show invariants on the memory store and constraints on the possible ways that the memory store can evolve in a valid target program. Given these invariants and constraints, we show that \new{}, \get{}, and \set{} behave the same way in the source and target. The tricky part of the proof is setting up the invariants and theorems---the proofs follow from extensive but straightforward casework.

\begin{definition}
(Valid memories) Consider memory store $(\pi, \mu)$. Consider arbitrary label $l \in L(\pi)$ and suppose $\pi[l] = (\lambda, k)$. We say $(\pi, \mu)$ is valid at $l$ if the following hold.
\begin{enumerate}
\item $k \in L(\mu)$. Assuming this is true, let $\mu[k] = (V, A, L)$.
\item $0 \leq \lambda \leq V$
\item For all $i$ and $0 \leq j < \text{\func{len}}(L[i])$ if $L[i][j] = (V', v)$ then $V' < V$
\item For all $i$ and $0 \leq j < k < \text{\func{len}}(L[i])$, if $L[i][j] = (V_1, v_1)$ and $L[i][k] = (V_2, v_2)$ then $V_1 < V_2$
\end{enumerate}
We say $(\pi, \mu)$ \emph{valid} if for all $l \in L(\pi)$, $(\pi, \mu)$ is valid at $l$. 
\end{definition}

\begin{theorem}
Let $e$ be an expression in the source language. Suppose that $(\{\}, \{\}), e \to^*_T (\pi, \mu), e'$. Then $(\pi, \mu)$ \emph{valid}.
\end{theorem}

\begin{proof}
(Outline) By induction on the length of the transition sequence taking $(\{\}, \{\}), e$ to $(\pi, \mu), e'$. For the base case, verify that each condition holds for $(\{\}, \{\})$.

Our IH is that that the theorem holds for all transition sequences of length $n$. We then consider an arbitrary transtion sequence of length $n+1$. From the IH, the first $n$ steps leads to a valid memory state. We then case on all possible modifications that the $n+1^{\text{th}}$ step can make to memory. Showing properties (1) to (3) is straightforward. To prove property (4), we need to appeal to a lemma that different executions of lines~\ref{line:start-cmpswap} to~\ref{line:end-cmpswap} in the implementation of \set{} on the same arraydata cannot overlap. The lemma holds because of property (2) and the compare and swap.
\end{proof}

\begin{definition}
(Memory transformations) Consider $(\pi, \mu), (\pi', \mu')$ \emph{valid} \footnote{Note that the relation is only defined for valid memory states.}. Consider arbitrary label $l$ with $l \in L(\pi)$ and $l \in L(\pi')$. Suppose that $\pi[l] = (\lambda, k)$ and $\pi'[l] = (\lambda', k')$. We say that $(\pi, \mu) \leq (\pi, \mu)$ at $l$ if the following hold.
\begin{enumerate}
\item $\lambda = \lambda'$ and $k = k'$ (in other words, values in $\pi$ are immutable). Assuming this is true, let $\mu[k] = (V, A, L)$ and $\mu'[k] = (V', A', L')$.
\item $V \leq V'$. In other words, the versions are non-decreasing.
\item For all $i$, \func{len}$(L[i]) \leq$ \func{len}$(L'[i])$ and for all $i, j$ with $0 \leq j < $ \func{len}$(L[i])$, $L[i][j] = L'[i][j]$. Intuitively, this means that logs can only be extended, existing entries cannot be overwritten.
\item If \func{len}$(L[i]) <$ \func{len}$(L'[i])$ then $L'[$\func{len}$(L[i])] = (V'', A[i])$ where $V \leq V''$.
\item For all $i$, if $A[i] \neq A'[i]$ then $\lambda < V'$ and exists $0 \leq j < \text{\func{len}}(L'[i])$ with $L'[i][j] = (V'', A[i])$ and $\lambda \leq V''$.
\end{enumerate}
We say that $(\pi, \mu) \leq (\pi', \mu')$ if $L(\pi) \subseteq L(\pi')$ and for all $l$ s.t. $l \in L(\pi)$ and $l \in L(\pi')$, $(\pi, \mu) \leq (\pi', \mu')$ at $l$.
\end{definition}

\begin{lemma}
(Reflexivity) If $(\pi, \mu)$ \emph{valid} then $(\pi, \mu) \leq (\pi, \mu)$. 
\end{lemma}

\begin{lemma}
(Transitivity) If $(\pi_0, \mu_0) \leq (\pi_1, \mu_1)$ and $(\pi_1, \mu_1) \leq (\pi_2, \mu_2)$ then $(\pi_0, \mu_0) \leq (\pi_2, \mu_2)$.
\end{lemma}

\begin{theorem}
Let $e$ be an expression in the source language. Suppose that $(\{\}, \{\}), e \to^*_T (\pi, \mu), e'$ and $(\pi, \mu), e' \to^*_T (\pi', \mu'), e''$. Then $(\pi, \mu) \leq (\pi', \mu')$.
\end{theorem}

\begin{proof}
(Outline) By induction on the length of the transition sequence taking $(\pi, \mu), e'$ to $(\pi', \mu'), e''$. The base case is trivial. For the inductive step case on the possible modifications the implementation can make to memory and then use transitivity of $\leq$.
\end{proof}

\begin{definition}
Suppose $A_S, A_T$ val. We say that $\sigma, A_S \sim (\pi, \mu), A_T$ if $A_S, A_T$ have the same length and for all valid $i$, $\sigma, \tget{}(A_S, i) \to^* \sigma', v$ and $(\pi, \mu), \tget{}(A_T, i) \to^*_T (\pi', \mu'), v$ for some $v$ int val.
\end{definition}

\begin{lemma}
\label{preservation-lemma}
(Monotonicity) If $\sigma, A_S \sim (\pi, \mu), A_T$ and $(\pi, \mu) \leq (\pi', \mu')$ then $\sigma, A_S \sim (\pi', \mu'), A_T$.
\end{lemma}

Because the target language supports concurrency, the memory store might be modified (by other evaluations of \new{} and \set{}) while \new{}, \get{}, and \set{} are being evaluated. To account for this we define concurrent transition sequences.

\begin{definition}
A \emph{concurrent transition sequence} $\tau$ in the target is a left sequence of memories $[(\pi_0,\mu_0), ..., (\pi_{n-1},\mu_{n-1})]$, right sequence of memories $[(\pi_0',\mu_0'), ..., (\pi_{n-1}',\mu_{n-1}')]$ and expressions $[e_0, ..., e_n]$ with $(\pi_i,\mu_i), e_i \to_T (\pi_i',\mu_i'), e_{i+1}$ for all $i$, and $(\pi_i', \mu_i') \leq (\pi_{i+1}, \mu_{i+1})$ for all $0 \leq i < n-1$. We say that $\tau$ takes $(\pi_0, \mu_0), e_0$ to $(\pi_{n-1}',\mu_{n-1}'), e_{n-1}'$.
\end{definition}

\begin{theorem}
(\new{}) Suppose that for some $A_S, A_T$ \emph{val}, \\$\sigma, \tnew{}(n, v) \to^* \sigma', A_S$ and $(\pi, \mu), \tnew{}(n,v) \to^*_T (\pi', \mu'), A_T$. Then $\sigma', A_S \sim (\pi', \mu'), A_T$.
\end{theorem}

\begin{proof}
Because the logs are empty, \get{} on every index of $A_T$ evaluates to 0. Similarly, \get{} on every index of $A_S$ evaluates to 0.
\end{proof}

\begin{theorem}
\label{thm-get}
(\get{}) Suppose that $\sigma, A_S \sim (\pi, \mu), A_T$. Fix $i$ \emph{int val}. Suppose that $\sigma, \tget{}(A_S, i) \to^* \sigma’, v$ where $v$ \emph{int val}. Let $e_0 = \tget{}(A_T, i)$ and suppose that $(\pi, \mu) \leq (\pi_0, \mu_0)$. Consider a concurrent transition sequence from $(\pi_0, \mu_0), e_0$ to $(\pi_{n-1}', \mu_{n-1}'), e_n$ with $e_n$ \emph{int val}. Then $e_n = v$.
\end{theorem}

\begin{proof}
(Outline) We case on the line that the implementation of \get{} returns. There are 4 cases: lines \ref{line:get-version-check}, \ref{line:log-empty}, \ref{line:log-small}, and \ref{line:binary-search}. In each case, we show that $e_n$ is the same as the value $(\pi, \mu), \tget{}(A_T, i)$ evaluates to. Since $\sigma, A_S \sim (\pi, \mu), A_T$, this is the same value that $\sigma, \tget{}(A_S, i)$ evaluates to, which implies $e_n = v$.

The first case is on line \ref{line:get-version-check} in the implementation of \get{}, where \emph{guess} is returned if the version of the sequence and arraydata are the same. Since sequence versions are immutable, arraydata versions are non-decreasing, and sequence versions are $\leq$ array data versions, the sequence and arraydata versions are also the same on line \ref{line:load-value}. From the implementation of \get{}, if the entire \get{} was evaluated atomically on line \ref{line:load-value}, it would evaluate to \emph{guess}. By lemma~\ref{preservation-lemma} this means that \emph{guess} and $v$ are the same.

The other cases on lines \ref{line:log-empty} and \ref{line:log-small} are similar. In the last case, \get{} involves a binary search on the logs. The theorem follows for the binary-search case from property (3) in memory transformations (logs can only be extended) and property (4) in valid memories (versions in a log are strictly increasing).
\end{proof}

\begin{theorem}
(\set{}) Suppose that $\sigma, A_S \sim (\pi,\mu), A_T$. Fix $i, v$ \emph{int val}. Suppose that $\sigma, \tset{}(A_S, i) \to^* \sigma', A_S'$ where $A_S'$ \emph{val}. Let $e_0 = \tset{}(A_T, i, v)$ and suppose that $(\pi, \mu) \leq (\pi_0, \mu_0)$. Consider a concurrent transition sequence from $(\pi_0, \mu_0), e_0$ to $(\pi_{n-1}', \mu_{n-1}'), e_n$ with $e_n$ \emph{val}. Then $\sigma', A_S' \sim (\pi_{n-1}', \mu_{n-1}'), e_n$.
\end{theorem}

\begin{proof}
(Outline) Case on whether the compare and swap in \set{} evaluates to true or false.

If it evaluates to false, then the implementation calls \get{} at each index of $A_T$ and creates a new arraydata in $\mu$. In this case $\sigma', A_S' \sim (\pi_{n-1}', \mu_{n-1}'), e_n$ follows from theorem~\ref{thm-get}.

If it evaluates to true, then from lemma~\ref{preservation-lemma} $A_S$ and $A_T$ are related when the compare and swap evaluates. In the source, \set{}$(A_S, i, v)$ evaluates to $A_S[i \mapsto v]$. In the target, suppose that $A_T = \&l$, $\pi[l] = (\lambda, k)$ and $\mu[k] = (V, A, L)$. As argued before, different executions of lines~\ref{line:start-cmpswap} to~\ref{line:end-cmpswap} in the implementation of \set{} on the same arraydata cannot overlap. So in the target, the implementation sets $\mu[k] = (V+1,A[i \mapsto v], L')$ for some $L'$ and returns $\&l'$ s.t. $\pi[l'] = (V+1, k)$. Going through the definition of $\sim$ this gives us $\sigma', A_S' \sim (\pi_{n-1}', \mu_{n-1}'), \&l'$.

\end{proof}

\begin{theorem}
(Bounded cost) There exists $k$ s.t. if $A_T$ has size $n$ then there does not exist $(\pi, \mu)$ \emph{valid} and a concurrent transition sequence of length $> kn$ starting at either $(\pi, \mu), \tget{}(A_T, i)$ or $(\pi, \mu), \tset{}(A_T, i, v)$ for all $i, v$ \emph{int val}.
\end{theorem}

\begin{proof}
(Outline) Because we copy the contents of a sequence after $n$ updates, where $n$ is the size of the sequence, for some $k$ independent of $n$, \set{} involves at most $kn$ steps, and \get{} involves at most $k\log{n}$ steps.
\end{proof}

\subsection{Interleaved Cost Bounds}

We want to show that work done in the interleaved structural dynamics is an upper bound for the work done in the implementation. We first sketch a linearizability-type result~\cite{linearizability}.

Consider the execution $E$ of a program. We assume a sequentially consistent model of computation. Consider a sequential execution $S$ of instructions in $E$ that produces the same result as $E$.

Consider any \get{} in the sequential execution $S$. From the previous subsection, the result that the program evaluates to does not depend on how the instructions in the \get{} are interleaved. Consider line \ref{line:get-version-check} in \get{} where the version of the array is compared with its arraydata. If the versions match then \get{} involves a constant amount of work. Otherwise, in the worst case \get{} involves a binary search on $n$ elements, where $n$ is the size of the array. This takes work proportional to $\log{n}$ \footnote{\set{} also copies the sequence data once every $n$ times the version number is increased, but the copying is amortized constant time.}.

So the execution $S$ is upper bounded (in cost) by an execution where the entire \get{} is evaluated atomically at the instruction where the versions of the array and arraydata are compared, where the \get{} has constant work if the version check succeeds and $\log{n}$ work if the check fails.

Similarly, consider any \set{} in the sequential execution $S$. From the previous subsection, the result that the program evaluates to does not depend on how the instructions in the \set{} are interleaved. Consider line \ref{line:set-cmpswap} in \set{} which involves a compare and swap on the arraydata's version. If the compare and swap is successful then \set{} involves a constant amount of work. Otherwise, in the worst case \set{} involves copying over $n$ elements, where $n$ is the size of the array. This takes work proportional to $n$.

So the execution $S$ is upper bounded (in cost) by an execution where the entire \set{} is evaluated atomically at the compare and swap instruction, where the \set{} has constant work if the compare and swap succeeds and $n$ work if the compare and swap fails.

We can trivially assume that \new{} is evaluated atomically since none of the effects of \new{} are observable until it finishes evaluating, and \new{} always takes time proportional to the size of the array being created.

We then define a relation between sequences in the source and target. $\sigma, A_S$ in the source and $(\pi, \mu), A_T$ in the target are related if \get{} evaluates to the same value at all indices, and $A_S$ is a leaf sequence if and only if $\lambda = V$ where $A_T = \&l$, $\pi[l] = (\lambda, k)$, and $\mu[k] = (V, A, L)$. We show that the relation is preserved by \new{} and \set{} and that costs of operations in the source are an upper bound for operations in the target for related sequences. Since the execution was transformed so that \new{}, \get{}, and \set{} evaluate atomically, we can then prove and apply a parametricity theorem to show that programs in the source and target evaluate to the same value.

\subsection{Parallel Cost Bounds}

Next, we sketch the case where we have an unbounded number of processors. In the worst case, \get{} involves a binary search on $n$ elements, where $n$ is the size of the array. This takes $\log{n}$ time on a machine with an unbounded number of processors. In the worst case, \set{} involves copying $n$ array elements and $n$ log entries where $n$ is the size of the array. In a machine with an unbounded number of processors, the copying can be done in $\log{n}$ time. Similarly, \new{} can execute in $\log{n}$ time on a machine with an unbounded number of processors. Note that the implementation of \new{}, \get{}, and \set{} are wait-free so these costs are independent of what other threads are doing.

It follows that we can use the evaluational dynamics to compute the work $W$ and span $S$ of evaluating an expression $e$ in our source language. We can then use theorem~\ref{thm:brent} to get that the cost of evaluating $e$ on a $p$-processor machine is $\leq c(\frac{W}{P} + S\log{P})$ for some constant $c$ independent of $e$.

### 2016-09-25 Paper/intro.tex
Reorganize repository

@@
 
-Supporting sequences (arrays) with efficient (constant time) random
-access updates has been a persistent problem in functional languages
-(pun intended).  It is easy to implement such sequences with constant
+Supporting sequences (arrays) with efficient (constant time)
+accesses and updates has been a persistent problem in functional languages
+(pun not intended).  It is easy to implement such sequences with constant
 time random access reads (using arrays stored as contiguous memory),
 or to implement with logarithmic time reads and updates using balanced
-trees, but getting both in constant time cannot be done without some
+trees, but it seems that getting both in constant time cannot be
+achieved without some
 form of extension.
@@
 threaded and can thread mutable arrays (or other mutable
-state) implicitly so the only reference to them is the current one.
+state) so the only reference to them is the current one.
 Haskell supplies the ST monad for this purpose.  King and Launchbery,
-for example, use STMonads with an array to implement depth first (DFS)
-search in linear work~\cite{KL95}.  Guzman and Hudak's single-threaded
+for example, use STMonads with an array to implement depth first 
+search (DFS) in linear work~\cite{KL95}.  Guzman and Hudak's single-threaded
 lambda calculus also uses the idea of single threading the
-computation~\cite{GH90}, and motivate the approach based on allowing
+computation~\cite{GH90}, and they motivate the approach based on allowing
 for fast array updates.  The problem with using these approaches is
-that it is basically implementing a second imperative language within
+that they are basically implementing a second imperative language within
 a functional language.  Using monads means using a new syntax not just
-for the operations themselves but for everything the operations are
+for the operations themselves but for much of the code the operations are
 embedded in.  Indeed King and Launchbery have to write completely
@@
 versions of structures, exactly because the state can be changed.
-More importantly, however, it forces the program itself to be single
+More importantly, however, monads force the program itself to be single
 threaded, inhibiting parallelism.
@@
 A third approach is to support fully functional arrays in a general
-functional language, and to check in the program if they happen to be
+functional language, and to check either statically or dynamically if they happen to be
 used in a single threaded manner.  This can be done, for example, by
 using reference counting~\cite{HB85,Hudak86}.  The idea is to keep
-track of how many references there are to an array and update inplace
+track of how many references there are to an array and update in place
 if there is only one.  Hudak describes techniques to statically
@@
 must be one, and therefore it is safe to replace an update with an
-inplace update.  The problem with this approach is that the efficiency
+in place update.  The problem with this approach is that the efficiency
 of the code depends heavily on the effectiveness of the compiler and
@@
 The last approach is to fully support functional arrays, even with
-sharing, but more efficiently than as a simple array or tree.  This is
+sharing, but using more sophisticated data structures.  This is
 sometimes referred to as version tree arrays~\cite{AHN88} or fully
@@
 
-None of the existing approaches support concurrent operations on
-arrays.  O' Neill suggests (in passing) having a lock for each element
+None of the existing approaches properly support concurrent operations on
+functional arrays.  O' Neill suggests (in passing) having a lock for each element
 of the array. However, when many threads contend for the same element,
-this would serialize accesses and make accesses not
-$O(1)$. Additionally, per-element locks add significant memory
+this would serialize accesses and therefore they would not take
+constant time.   Additionally, per-element locks add significant memory
 overhead.
+
+% Using arrays in a concurrent or persistent manner is not simply of theoretical importance.
+As an example application, consider a producer-consumer situation where a producer is rapidly writing values to a mutable array and a consumer has registered a callback that triggers when the array reaches a certain state.
+The callback expects to see what the array was like when the event was triggered, even though the producer may continue writing values to the array.
+The naive way to implement this is to copy the array and pass the copy to the callback.
+However, copying the array is inefficient and the producer is blocked from making progress during the copy.
+We can use concurrent functional arrays for this application.
+
 % Separation logic ~\cite{reynolds} could be used to
@@
 
-\subsection*{Our Approach}
+\subsection*{Our Approach: Sequences}
 
 In this paper we present an approach for efficiently supporting
-functional arrays.  It uses some ideas from the previous approaches
+functional arrays (sequences).  It uses some ideas from the previous approaches
 but unlike the previous approaches it supplies a well
-defined cost semantics, and supports parallelism, and without language
+defined cost dynamics, and supports parallelism, and without language
 extensions.  More specifically the approach has the
@@
 \begin{enumerate}
-\item It has fully functional value semantics (dynamic
-  semantics)---i.e., when not considering costs, arrays act no
-  differently than purely functional arrays.
+\item It has fully functional value dynamics---i.e., when not considering costs, sequences act no
+  differently than purely functional sequences.
 \item It requires no changes in existing languages and no special
@@
   functional programming style.
-\item The approach supports nested parallelism---arrays can be passed
+\item The approach supports nested parallelism---sequences can be passed
   to parallel calls and safely updated and read, again with a purely
-  functional semantics.
-\item We supply a well defined cost semantics, which can be used to
-  formally analyze the cost of any program.  The semantics captures
+  functional dynamics.
+\item We supply a well defined cost dynamics, which can be used to
+  formally analyze the cost of any program.  The dynamics captures
   both sequential and parallel costs.
@@
 \item
-Although accessing old versions of an array is more expensive than
-accessing new versions (as defined in the cost semantics), reading old
+Although accessing old versions of a sequence is more expensive than
+accessing new versions (as defined in the cost dynamics), reading old
 versions never costs more than $O(\log n)$ work.
@@
 
-Although our value semantics are purely functional (and compositional)
-our cost semantics are not---they require passing a store and modeling
-costs based on the order the store is threaded.  The reader might
-object to this non-functional nature of the cost semantics, but we
-note that non-functional costs are also requires in call-by-need
-(lazy) languages, which supply a functional value semantics but an
-imperative cost semantics.  In fact from the value point of view,
-call-by-need is no different from call-by-name, and it is only when
-considering costs that they differ.  The cost semantics of
-call-by-need~\cite{AMOFW95} is completely non-functional
-(non-compositional).  Furthermore it is likely significantly more
-complicated to reason about than the cost semantics we present in this
-paper.
+Although our value dynamics is purely functional (and compositional)
+our cost dynamics is not---they require passing a store and modeling
+costs based on the order the store is threaded.  To allow for
+parallelism we allow for arbitrary interleaving of the steps in the
+cost dynamics.  We note that lazy languages also have a pure value
+%dynamics but impure cost dynamics---call-by-name and call-by-value
+dynamics but impure cost dynamics---call-by-name and call-by-need
+only differ in their costs, and a store is required to model the
+difference~\cite{AMOFW95}.
 
@@
 \includegraphics[scale=0.45]{leaf_interior_intro}
-\nocaptionrule \caption{Example usage of functional arrays}
+\nocaptionrule \caption{Example usage of sequences}
 \label{fig:leaf_interior_intro}
@@
 
-In this paper we consider arrays with three functions: \tabulate{},
-\get{} (read), and \set{} (write).  As usual, \tabulate{}$(f,n)$
-creates a new array of length $n$ by applying the function $f$ to
-$[0, \ldots, n-1]$, \get$(A,i)$ returns the $i^{th}$ element of $A$, and
-\set$(A,i,v)$ sets the $i^{th}$ element of $A$ to $v$ returning a
-``new'' array (in the following discussion $n$ refers to the length of
-an array).  Figure~\ref{fig:leaf_interior_intro} gives an example.  As
-in previous work~\cite{AHN88} the history of an array after it is
-created with \tabulate{} can be viewed as a version tree.  A version
+In this paper we consider sequences with three functions: \new{},
+\get{} (read), and \set{} (write). \new{}$(n,v)$
+creates a new sequence of length $n$ with the value $v$ at each index, \get$(A,i)$ returns the $i$-th element of $A$, and
+\set$(A,i,v)$ returns a ``new'' sequence where the $i^{th}$ element of $A$ is replaced by $v$ (in the following discussion $n$ refers to the length of
+an sequence).  Figure~\ref{fig:leaf_interior_intro} gives an example.
+
+As in previous work~\cite{AHN88} the history of a sequence after it is
+created with \new{} can be viewed as a version tree.  A version
 tree has interior nodes and leaves.  In the figure, after all
-functions are applied, arrays $A$ and $C$ are \emph{interior nodes},
-and arrays $B$ and $D$ are \emph{leaves} of the version tree.  In our
-cost semantics applying \get{} and \set{} to arrays at the leafs takes
+functions are applied, sequences $A$ and $C$ are \emph{interior nodes},
+and sequences $B$ and $D$ are \emph{leaves} of the version tree.  In our
+cost dynamics applying \get{} and \set{} to sequences at the leafs takes
 constant work, but applying them to interior nodes is more expensive.
-We define a cost semantics that identifies whether an array is a leaf
-or interior node and charges for the cost of \get{} and \set{}
-appropriately.  This requires threading a store that maps labels
-(corresponding to each array) to indicate whether the array is a leaf
-or interior.
 
-Since we are also interested in parallelism, we include fork-join
-parallelism in our language and semantics.    Our cost semantics then
-models the cost of arrays when used in parallel.  This is non-trivial
-since parallel tasks can read and update an array in parallel.
-To deal with this, we compute the work of each forked task separately,
-and then charge additional work at the join point if an array is used
-in both forks.
+We start with a simple pure call-by-value lamba calculus extended with
+sequences.  To define abstract costs and show the relationship to
+concrete cost (time) on a parallel machine we use two levels of cost
+dynamics: a structural cost dynamics and a evaluation cost dynamics.
+Our structural cost dynamics is a single step dynamics and is closer
+to the implementation.   The dynamics defines the work by summing the
+cost per transition across an arbitrary interleaving of the potential transitions
+(due to concurrency), and defines the span by allowing all potentially
+parallel transitions to proceed together.    We prove a mapping of the
+cost in the structural dynamics to time on a $P$ processor machine,
+assuming given costs for the sequence functions.   The cost is
+non-deterministic due to different possible interleavings.
 
-To implement our functional arrays, we maintain at each leaf of the
-version tree an array of the most recent values.  We also keep with
-each location a change-log that keeps track of values at interior
-nodes of the version tree.  Applying \get{} to a leaf only requires a
-lookup in the array.  Applying \set{} to a leaf only requires updating
-the array, and adding an element to the change log for the old
-version.  Applying \get{} to an interior nodes requires a binary
-search on the change log, which can be bounded by $O(\log n)$ time.
-To ensure that change logs do not get too large, whenever the total
-size across all change logs reaches $n$, the array is copied.   This
-requires $O(n)$ time, but can be amortized against the updates.
-Applying \set{} to an interior node requires creating a new array,
-and copying values to it.   This requires $O(n)$ time.
+The evaluation dynamics is a higher-level big-step dynamics more
+appropriate for a programmer to reason about costs.  In particular the
+costs in the evaluation dynamics are deterministic independent of
+interleaving.  We prove that the costs in the evaluation dynamics
+gives a tight worst case upper bound on the cost of the underlying structural
+dynamics, allowing for any interleaving.
+
+We describe the implementation in an impure language---our pure
+language extended with references and an atomic compare-and-swap on
+those references.  To implement sequences, we maintain at
+each leaf of the version tree a mutable array of the most recent values.  We
+also keep with each location a change-log that keeps track of values
+at interior nodes of the version tree.  Applying \get{} to a leaf only
+requires a lookup in the array.  Applying \set{} to a leaf only
+requires updating the array, and adding an element to the change log
+for the old version.  Applying \get{} to an interior nodes requires a
+binary search on the change log, which can be bounded by $O(\log n)$
+time.  To ensure that change logs do not get too large, whenever the
+total size across all change logs reaches $n$, the array is copied.
+This requires $O(n)$ time, but can be amortized against the updates.
+Applying \set{} to an interior node requires creating a new array, and
+copying values to it.  This requires $O(n)$ time.
 
 We give a wait-free concurrent implementation that uses careful
-synchronization. We prove linearizability of the implementation and
-use that to prove that our arrays work correctly when used
-concurrently. Finally, we give a provable implementation for our cost
-dynamics.
+synchronization.   We assume that instructions in the target language
+can be interleaved in any way--i.e., at any given step many \get{}s and
+\set{}s can be in progress.   All we require is that memory reads,
+writes and compare-and-swap are atomic.  We show correctness with
+respect to any interleaving, and show that specific linearization
+points within the implementation define the relative order of \get{}s
+and \set{}s with respect to the structural dynamics.
+
+To evaluate our approach in practice we ran a variety of experiments
+on a preliminary concrete implementation.  The
+experiments show that accesses are $3\%$ to $12\%$ slower in leaf sequences
+than in regular Java arrays, and updates are $2.4$ to $3.3$ times slower.
+Considering the purity and additional functionality provided by our sequence
+implementation, this is a small slowdown. Furthermore, preliminary experiments 
+back up our theoretical claim that threads can access our sequences with a high 
+level of concurrency.
+
+Our results on cost dynamics can be easily extended to other data
+types where the cost of \get{} and \set{} are different for leaf and
+interior versions, even if there are multiple varieties of \get{} and
+\set{} functions.   In particular, our sequence implementation can likely
+be extended to functional unordered sets implemented with hash tables,
+and our approach might be useful in coming up with more efficient
+implementations of functional disjoint sets (although a functional
+disjoint set implementation where all operations cost $\log{n}$ is
+known~\cite{persistentufds}).

### 2016-09-25 Paper/paper.tex
Reorganize repository

--- 2016-03-11 Paper/paper.tex
@@
 
+\usepackage{pseudocode}
 \usepackage{amsmath}
@@
 \usepackage{amsthm}
-\usepackage{listings}
+\usepackage{amsfonts}
 
 \newcommand{\cL}{{\cal L}}
+
+\newcommand{\func}[1]{{\sc #1}}
+\newcommand{\farray}{\text{sequence}}
+\newcommand{\tfunc}[1]{{\text{\func{#1}}}}
+\newcommand{\val}{\func{val}}
+\newcommand{\tval}{\text{\func{val}}}
+\newcommand{\abs}{\func{abs}}
+\newcommand{\new}{\func{new}}
+\newcommand{\cost}{\func{cost}}
+\newcommand{\tabulate}{\func{tabulate}}
+\newcommand{\get}{\func{get}}
+\newcommand{\set}{\func{set}}
+\newcommand{\push}{\func{push}}
+\newcommand{\size}{\func{size}}
+\newcommand{\tget}{\text{\get}}
+\newcommand{\tset}{\text{\set}}
+\newcommand{\tpush}{\text{\push}}
+\newcommand{\tsize}{\text{\size}}
+\newcommand{\tnew}{\text{\new}}
+\newcommand{\ttab}{\text{\tabulate}}
+\newcommand{\MutableArray}{\text{MutableArray }}
+\newcommand{\ArrayData}{\text{ArrayData }}
+\newcommand{\Array}{\text{Array }}
+\newcommand{\intref}{\text{int ref }}
 
@@
 
-\title{A New Approach for Parallel Functional Arrays}
+\title{Parallel Functional Arrays}
 
+% \authorinfo{}{}{\\[-3.0ex]}
 \authorinfo{Ananya Kumar}
            {Carnegie Mellon University}
-           {ananyak@andrew.cmu.edu}
-\authorinfo{Guy Blelloch}
+           {ananyak@andrew.cmu.edu\\[-5.0ex]}
+\authorinfo{Guy E. Blelloch}
            {Carnegie Mellon University}
            {blelloch@cs.cmu.edu}
-
+\authorinfo{Robert Harper}
+           {Carnegie Mellon University}
+           {rwh@cs.cmu.edu}
 \maketitle
@@
 \begin{abstract}
-In this paper we introduce a $O(1)$ wait-free, parallel, functional array, that allows $O(1)$ reads and writes to the most recent version of the array.
-We describe the cost dynamics and sketch out a provable implementation.
-We show favorable benchmarks comparing our functional arrays with regular arrays in Java.
+The goal of our paper is to come up with functional arrays (sequences) that are as efficient as imperative arrays, can be used in parallel, and have well defined cost-semantics.
+Our approach is to consider arrays with functional value semantics but non functional cost semantics.
+Because the value semantics is functional, ``updating" an array returns a new array.
+We allow operations on ``older" arrays (called interior arrays) to be more expensive than operations on the ``most recent" arrays (called leaf arrays) .
+
+We embed sequences in a language that supports fork join parallelism.
+Because we support fork-join parallelism, the cost of programs is non-deterministic and depends on the order in which operations are interleaved.
+This is because the ordering of operations can affect whether a sequence is a leaf or interior.
+Consequently the concurrent costs of such programs are difficult to analyze.
+Our main result is the derivation of a deterministic evaluational dynamics which makes analyzing the costs much easier.
+Our theorems are not array-specific and can be applied to any data type with different costs for operating on interior and leaf versions.
+
+We give a wait-free concurrent sequence implementation which requires constant work for accessing and updating leaf arrays and bounded work for operations on interior arrays.
+We sketch out our proof of correctness for the array implementation.
+The key advantages of our sequence approach compared to current approaches is that our implementation requires no changes to existing programming languages, supports nested parallelism, and has well defined cost semantics.
+At the same time, it allows for functional implementations of algorithms like depth-first search with the same asymptotic complexity as imperative implementations.
 \end{abstract}
 
-\category{CR-number}{subcategory}{third-level}
+\category{D.3.1}{Programming Languages}{Formal Definitions and Theory}
+\category{F.3.2}{Logics and Meanings of Programs}{Semantics of Programming Languages}
 
 \keywords
-array, parallel, cost semantics
+cost semantics, concurrency, parallel, functional data structures, persistence, arrays
 
-\section{Introduction}
+\input{intro}
 
-Arrays are very important in functional programming languages because they allow work-efficient implementations of algorithms like depth-first search.
-Accessing old versions of arrays can be useful for efficient checkpointing, logging, and event handling.
+\input{dynamics}
 
-\subsection{Previous Approaches}
+\input{structural}
 
-There has been a lot of great work on functional arrays, but previous approaches have limitations.
+\input{evaluational}
 
-In many functional programming languages, like Haskell, monads ~\cite{wadler-monads} can be used to enforce single threaded use of arrays.
-However, enforcing single threaded use obstructs persistence - it prevents us from accessing or modifying old versions of an array.
+\input{costproof}
 
-Another approach is to use reference counting ~\cite{hudak}.
-If the number of references to an array is one, the array can be updated in place, otherwise the array is copied and then updated.
-This approach makes it difficult for programmers to statically analyze the time complexity of array operations.
-Furthermore, even if there are multiple reference to an array the array might not need to be copied, so the unnecessary copying could make the implementation inefficient.
+\input{implementations}
 
-A third approach is to use linear types ~\cite{wadler-lintypes} to ensure that programmers can only access and mutate arrays in valid ways.
-Linear types have not been used practically for a few decades since its inception, and would require a complete re-think of a language's type system to be useful.
+\input{experiments}
 
-O' Neill's ~\cite{oneill} functional arrays guarantee $O(1)$ access and insertions when used like an imperative array.
-However, the arrays explicitly implement version trees which are inefficient in practice.
-Additionally, the reads and writes to the most recent version of the array might take $O(\log{n})$ work if the array is not used in a single threaded way.
-
-Chuang's approach ~\cite{chuang} supports $O(1)$ accesses and insertions to the most recent version of arrays, however accessing old versions of the array takes $O(n)$ work which is often impractical.
-
-None of the existing approaches support concurrent operations on arrays.
-O' Neill suggests having a lock for each element of the array.
-However, when many threads contend for the same element, this would serialize accesses and make accesses not $O(1)$.
-Additionally, per-element locks add significant memory overhead.
-Separation logic ~\cite{reynolds} could be used to parallelize divide and conquer algorithms on arrays however they would severely restrict the way functional arrays can be used.
-
-\subsection{Our Approach}
-
-Our arrays support 3 functions: tabulate, get, and set. The dynamics are functional, in particular set returns a new array and leaves the original array intact.
-
-\begin{figure}[!ht]
-\centering
-\includegraphics[scale=0.45]{leaf_interior_intro}
-\nocaptionrule \caption{Example usage of functional arrays}
-\label{fig:leaf_interior_intro}
-\end{figure}
-
-After all the operations performed in figure \ref{fig:leaf_interior_intro}, A and C are \emph{interior} arrays and B and D are \emph{leaf} arrays.
-Calling get and set on leaf arrays costs $O(1)$ but calling get and set on interior arrays is more expensive.
-In the cost dynamics, we thread a store that keeps track of whether an array is an interior or leaf array.
-
-We also give and prove fork join cost dynamics for our arrays.
-The dynamics for arrays used concurrently are non trivial because one thread might call get on a leaf array while another thread calls set.
-To deal with this, we compute the work of each fork separately, and then charge additional work at the join point if an array is used in both forks.
-
-In our implementation of a functional array, we keep an array of values (multiple functional arrays might point to the same array).
-For each index of the array, we also keep a history of values that were previously at that index.
-To access an element in an interior array, we need to search through the history of values.
-
-We give a wait-free concurrent implementation that uses careful synchronization.
-We prove linearizability of the implementation and use that to prove that our arrays work correctly when used concurrently.
-Finally, we give a provable implementation for our cost dynamics.
-
-\section{Dynamics}
-
-We use a standard applicative-order language defined as follows:
-$$e = x \; | \; c  \; | \; \lambda x.e \; | \; e_1 e_2 \; | \; e_1 || e_2 \; | \; \text{if } e_1 \text{ then } e_2 \text{ else } e_3$$
-
-$L$ contains the usual arithmetic types, such as the natural numbers, and numerical operations such as sums and products.
-Our dynamics are given by the judgement $e \Downarrow v$ which means that expression $e$ evaluates to $v$.
-Dynamics for function application and fork-join are given as examples.
-
-$$\frac{e_1 \Downarrow v_1 \;\;\; e_2 \Downarrow v_2}{e_1 || e_2 \Downarrow (v_1, v_2)} \text{ (fork-join)}$$
-$$\frac{f \Downarrow \lambda x . e \;\;\; y \Downarrow v}{f \; y \Downarrow [v/x]e} \text{ (function-app)}$$
-
-The language also includes 3 functions for working with arrays: tabulate, get, and set. The dynamics for tabulate is given below:
-$$\frac{f(1) \Downarrow v_1, ..., f(n) \Downarrow v_n}{\text{tabulate} \; f \; n \Downarrow [1 \mapsto v_1, ..., n \mapsto v_n]}$$
-
-$get(A,i)$ returns the value at the $i$th index of $A$.
-$set(A,i,v)$ returns a new array where the value at index $i$ is $v$ and the value at all other indices is the same as in $A$.
-The evaluation of a simple program in our language is shown below:\\
-
-$(\lambda A. \; get(A, 2) + get(A,3)) set(tabulate(\lambda i . \; i, 5), 2, 10)$
-
-$(\lambda A. \; get(A, 2) + get(A,3)) set([1,2,3,4,5], 2, 10)$
-
-$(\lambda A. \; get(A, 2) + get(A,3)) [1,10,3,4,5]$
-
-$get([1,10,3,4,5], 2) + get([1,10,3,4,5],3) $
-
-$10+3$
-
-$13$
-
-\section{Cost Dynamics}
-
-The dynamics given in the previous section are functional.
-However, the costs of get and set depend on whether the array is a leaf array or an internal array.
-To deal with this, we thread a store through the cost dynamics to keep track of whether each array is a leaf or internal array.
-
-The cost dynamics of fork-join pose an additional challenge.
-An array can be used concurrently by multiple threads and the costs could depend on the order in which the instructions are interleaved.
-Figure \ref{fig:fork_join_intuition} shows 3 ways a leaf array $A$ might be used.
-In particular, one side of the fork might call get on $A$ while the other side calls set (see the center panel).
-Calls to get on the left branch would be $O(1)$ if and only if they execute before the call to set on the right branch.
-
-\begin{figure}[!ht]
-\centering
-\includegraphics[scale=0.3]{fork_join_intuition}
-\nocaptionrule \caption{Different ways an array might be used in a fork join}
-\label{fig:fork_join_intuition}
-\end{figure}
-
-We cannot make any assumptions about how the instructions in different branches of a fork are interleaved, so we give the worst case time complexity over all possible interleavings.
-We first compute the work of each fork separately.
-Then for arrays that are used in both forks, we add additional work at the join point.
-To compute the additional work at the join point, we keep track of the number of gets that cost $O(1)$ on each side of the fork.
-If the array was modified on the other side of the fork, then the modification might have happened before the get, so we need to charge additional work for the get.
-
-Our cost dynamics are defined by the following judgement, where $\delta$ is the store, $e$ is the expression, $\delta'$ is the new store, $v$ is the value returned by evaluating $e$, $w$ is the work, and $s$ is the span.
-$$\delta; e \boldsymbol\Downarrow \delta'; v; w; s$$
-
-We give the cost dynamics for expressions in the language: (insert cost dynamics for standard functions later)
-
-Arrays are represented by $(l, V)$ where $l$ is a label that uniquely identifies the array, and $V$ contains the values in the array.
-$V[i]$ is the $i^{\text{th}}$ value in the array.
-If the set function has been called on an array, the array is an interior array (represented by $-$), otherwise the array is a leaf array (represented by $+$).
-
-Array functions have different costs depending on whether the array is a leaf or interior array.
-Let $C(l)$ be the number of get operations applied on array $A = (l, V)$ that had $O(1)$ work.
-We thread a store $\delta$ through each operation.
-$\delta$ maps array labels $l$ to $(+/-, c)$, where $c = C(l)$.
-Let $f(n)$ be the work of calling get on an interior array and $g(n)$ be the work of calling set on an interior array.
-
-The cost dynamics for the functions on a functional array of size $n$ are:
-$$\frac{l = \text{new label} \;\;\; f(1) \Downarrow v_1; w_1; s_1; ...
-f(n) \Downarrow v_{n}; w_{n}; s_{n}}{\delta; \text{tabulate } f \; n \Downarrow \delta[l \mapsto (+, 0)]; (l, \bigcup_{i=1}^n [i \mapsto v_i]); 1+\sum w_i; 1+\max s_i}$$
-$$\frac{A = (l,V) \;\;\; \delta[l \mapsto (+, c)] \;\;\; l' = \text{new label}}{\delta; \text{set} \; A \; i \; v \Downarrow \delta[l \mapsto (-, c), l' \mapsto (+, 0)]; (l', V[i \mapsto v]); 1; 1} \text{  (set-leaf)}$$
-$$\frac{A = (l,V) \;\;\; \delta[l \mapsto (-, c)] \;\;\;  l' = \text{new label}}{\delta; \text{set} \; A \; i \; v \Downarrow \delta[l' \mapsto (+, 0)]; (l', V[i \mapsto v]); g(n); 1} \text{  (set-interior)}$$
-$$\frac{A = (l,V) \;\;\; \delta[l \mapsto (+, c)]}{\delta; \text{get} \; A \; i \Downarrow \delta[l \mapsto (+, c+1)]; V[i]; 1; 1} \text{  (get-leaf)}$$
-$$\frac{A = (l,V) \;\;\; \delta[l \mapsto (-, c)]}{\delta; \text{get} \; A \; i \Downarrow \delta; V[i]; f(n); 1} \text{  (get-interior)}$$
-
-Set creates a new label and array and modifies the store to indicate that the newly created array is a leaf array, and the array set was called on has become an interior array.
-The work for set is higher for interior arrays.
-Get returns the value at the desired index, if the array is a leaf then the operation costs $O(1)$ but we increment the counter of $O(1)$ gets in the store.
-
-The fork-join cost semantics are the most interesting. Suppose we have an expression $e_L || e_R$ and
-\[ \delta, e_L \Downarrow \delta_L; v_L; w_L, s_L \]
-\[ \delta, e_R \Downarrow \delta_R; v_R; w_R, s_R \]
-
-We assume that newly created labels in $\delta_L$ and $\delta_R$ do not conflict.
-Consider an array $A = (l, V)$ with $\delta[l] = (s,c)$, $\delta_L[l] = (s_L, c + c_L)$, $\delta_R[l] = (s_R, c + c_R)$.
-When multiplying two signs or multiplying a sign with an integer, consider + to be 1 and - to be 0.
-
-Combine describes how to combine the values in the store on both sides of the fork-join.
-The array is a leaf array iff it is a leaf array on both sides of the fork-join.
-$O(1)$ gets on one side of the fork remain $O(1)$ iff there were no calls to set on the other side of the fork.
-\begin{equation*}
-  \begin{aligned}
-   combine&((s, c), (s_L, c + c_L), (s_R, c + c_R))= \\
-   &(s_L s_R, c + s_R c_L + s_L c_R)
-  \end{aligned}
-\end{equation*}
-Extrawork gives the additional cost incurred if an array was modified on either side of a fork-join.
-If the array was an interior array before the fork-join, then all functions incurred their maximal cost and there is no additional cost.
-Otherwise, $O(1)$ gets on one side of the fork become $O(f(n))$ iff the other side of the fork called set.
-If both sides of the fork called set, then one of the sets came first and is $O(1)$, the subsequent set is $O(g(n))$.
-\begin{equation*}
-  \begin{aligned}
-  extrawork&((s, c), (s_L, c + c_L), (s_R, c + c_R)) = \\
-  &s \cdot ( (\neg s_R) c_L f(n)  + (\neg s_L) c_R f(n) + (\neg s_R) (\neg s_L) g(n) )
- \end{aligned}
-\end{equation*}
-
-\begin{equation*}
-  \begin{aligned}
-  \delta' = \; &(\delta_L / \delta) \; \cup \; (\delta_R / \delta) \; \cup \\
-               &\bigcup_{l \in L(\delta_L) \cap L(\delta_R)} [l \mapsto combine(\delta[l], \delta_L[l], \delta_R[l])]
-  \end{aligned}
-\end{equation*}
-$$w' = \sum_{l \in L(\delta_1) \cap L(\delta_2)} extrawork(\delta[l], \delta_L[l], \delta_R[l])$$
-Then, the fork-join cost semantics are:
-$$\delta; e_1 || e_2 \Downarrow \delta'; (v_1, v_2); 1 + w_1 + w_2 + w'; 1 + \max(s_1, s_2)$$
-
-\section{Approach}
-
-An array $A$ is NEW if the set function has not been called with $A$ as an argument, otherwise $A$ is OLD.
-While the dynamics of our arrays are functional, the costs are different for OLD and NEW arrays.
-
-\begin{figure}[!ht]
-\centering
-\includegraphics[scale=0.3]{new_array_A}
-\nocaptionrule \caption{New functional array filled with 0s}
-\label{fig:new_array_A}
-\end{figure}
-
-Suppose that $A$ is a NEW functional array  (see figure \ref{fig:new_array_A}).
-$A$ has a version number $V$, and a pointer to an ArrayData object $AD$.
-$AD$ keeps a regular (mutable) array of values, which corresponds to the values in $A$.
-$AD$ has a version number, which is the same as $A'$s ($V$) to indicate that $A$ is NEW.
-For each element of the array in $AD$, keep a log of values that used to be at that index.
-The logs are initially empty.
-
-Get on NEW arrays: to get the $i^{\text{th}}$ element in $A$, simply access array[i] in $AD$.
-
-\begin{figure}[!ht]
-\centering
-\includegraphics[scale=0.3]{set_A_return_B}
-\nocaptionrule \caption{$B = set(A, 3, 5)$ changes the array in $AD$ and adds a log entry.}
-\label{fig:set_A_return_B}
-\end{figure}
-
-Set on NEW arrays: Suppose that $A[i]$ is $v_{old}$ and we want to do $B = set(A, i, v_{new})$ (see figure \ref{fig:set_A_return_B}).
-Add an entry $(V, v_{old})$ to the log at index $i$, expressing that $A[i]$ was $v_{old}$ at version $V$.
-Increment the version of $AD$ to $V+1$.
-Set $array[i]$ in $AD$ to $v_{new}$.
-Create a new functional array $B$, with version $V+1$, which points to $AD$.
-Notice that both $A$ and $B$ point to $AD$.
-However, the version of $A$ is $V$ and the version of $AD$ is $V+1$, which indicates that $A$ is OLD.
-The version of $B$ is $V+1$ which indicates that $B$ is NEW.
-
-Get on OLD array: Suppose we do $get(A, i)$ where $A$ is OLD and has version $V$. We binary search the log at index $i$ to find what value was stored in the array at version $V$.
-
-\begin{figure}[!ht]
-\centering
-\includegraphics[scale=0.3]{set_old_array}
-\nocaptionrule \caption{$D = set(A, 1, 9)$ copies $AD$ because $A$ is OLD}
-\label{fig:set_old_array}
-\end{figure}
-
-Suppose we do $D = set(A, i,v)$ where $A$ is old (see figure \ref{fig:set_old_array}).
-Create a new ArrayData object $AD'$ (with a new array).
-Copy the values from the array in $AD$ to the array in $AD'$, and then set $array[i]$ to $v$ in $AD'$.
-
-The logs are stored in a doubling array which doubles in size when full.
-This guarantees amortized $O(1)$ insertion of a log entry, and allows us to binary search with $O(\log{m})$ work, where $m$ is the number of log entries.
-
-Suppose that the size of the array is $n$.
-Every $n$ times an ArrayData object is modified, we create a new ArrayData object, and copy the values over to the new ArrayData object.
-This ensures that $m \leq n$ i.e.
-we don't have too many log entries in an ArrayData object.
-Copying is amortized $O(1)$.
-
-It follows that in the RAM model the work of set and get is $O(1)$ for a NEW array.
-Get in an OLD array involves a binary search and has work $O(\log{n})$.
-Set in an OLD array involves copying the array, and has work $O(n)$.
-
-\section{Concurrent Implementation}
-
-We implement the functional array functions using an imperative target language.
-
-\subsection{Concurrency Model}
-
-We assume a sequential consistency model of concurrency.
-That is, when analyzing the effects of a program, we assume that at each time step exactly one pending instruction is executed.
-Consider the execution of a program $P$.
-We label the time steps in the execution from 1 to $t$.
-
-\begin{definition}
-Consider an object $A$.
-Suppose that the program invokes functions $M_1$, $M_2$, ..., $M_n$ on $A$.
-Consider arbitrary $i$, and suppose that the instructions in $M_i$ were executed at times $s_1$, $s_2$, ..., $s_m$.
-We say that $A$ is \emph{linearizable} if the effect of the program is as if $M_i$ executed atomically at some time $s_j$, and this holds for all $M_i$.
-\end{definition}
-
-\begin{definition}
-Suppose that first instruction in function call $f$ was executed at time $s_1$ and the last instruction at time $e_1$.
-Suppose that first instruction in function call $g$ was executed at time $s_2$ and the last instruction at time $e_2$.
-We say the function calls do not \emph{overlap} if the intervals $[s_1, e_1]$ and $[s_2, e_2]$ do not overlap.
-\end{definition}
-
-Our programs have access to the link load store conditional (LLSC) function.
-LLSC takes an (address, old value, new value) tuple as argument.
-LLSC first checks that the value at address is the same as old value, if it is different then LLSC returns false.
-Otherwise LLSC atomically stores new value at address and returns true, however if the value at address was modified between the load and store, LLSC returns false.
-
-\subsection{Push Arrays}
-
-We use an auxiliary data structure, \emph{PushArrays}, to store log entries.
-PushArrays are initially empty and support 3 functions.
-Suppose we have a PushArray $A$.
-$A.push(e)$ inserts entry $e$, $A.size()$ returns the number of entries inserted, $A.get(i)$ returns the $i^{\text{th}}$ entry inserted.
-Note that $get$ is only defined between indices 0 and $A.size()-1$.
-All operations are amortized $O(1)$.
-
-PushArrays can be used semi-concurrently.
-At most one thread can execute $push$ at any time, but multiple threads can call $size$ and $get$.
-In a PushArray $A$, $A.size$ stores the number of entries and $A.data$ references an array that stores the actual entries.
-Pseudocode for $A.push(e)$ is shown below.
-
-\begin{lstlisting}[language=Java]
-if (isFull(A.data)) {
-  newData = new Array(A.data.capacity * 2);
-  copyValues(from = A.data, to = newData);
-  A.data = newData;
-}
-A.data[A.size] = e;
-A.size += 1;
-\end{lstlisting}
-
-\begin{theorem}
-Given that different calls to push (on the same PushArray) do not overlap, PushArrays are linearizable.
-\end{theorem}
-
-\begin{proof}
-The linearization point of $push$ is $A.size$ $+=$ $1$, and the linearization point of the other functions is their single instruction.
-In push, expanding the capacity of $A.data$ when it is full does not interfere with accessing $A.data$ because we copy values to the new array before pointing $A.data$ to the new array.
-The effects of adding $e$ are only observable after we increase $size$, since, by the specification of PushArrays, the result of calling $A.get(i)$ where $i \geq A.size()$ is not defined.
-\end{proof}
-
-\subsection{Functional Array Implementation}
-
-Suppose we have a FunctionalArray $A$.
-$A$ points to an ArrayData object which is referenced by $A.data$.
-The (imperative) array in $A.data$ is referenced by $A.data.values$, and the log entries corresponding to the $i^{\text{th}}$ index are stored in a PushArray $A.data.undo\_lists[i]$.
-We first explain the implementation of $A.set(pos, val)$.
-
-We use a link load store conditional to ensure that only 1 thread can modify an ArrayData object at any time.
-If $m$ threads try to modify $A.data$, then $m-1$ threads will fail the LLSC (returning false).
-Instead of modifying $A.data$, these threads will create a new ArrayData object, and will copy over values from $A.data$, as shown below.
-
-\begin{lstlisting}[language=Java]
-FunctionalArray newA;
-newA.version = A.version + 1;
-if (!LLSC(&A.data.version, A.version, 
-	newA.version) OR
-    A.version % A.size == 0) {
-  // Create a new ArrayData object AD
-  // Forall i = 1:n, AD.values[i] = get(A, i)
-  AD.values[pos] = val;
-  newA.data = AD;
-  return newA;
-}
-\end{lstlisting}
-
-At most one of the threads trying to modify the ArrayData object will get past the LLSC, will insert a log entry and then modify $A.data.values$.
-The ordering of inserting a log entry and modifying $A.data.values$ matters in the proof of get.
-The pseudocode for this step is shown below.
-
-\begin{lstlisting}[language=Java]
-newA.data = A.data;
-old_value = newA.data.values[pos];
-undo_list = newA.data.undo_lists[pos];
-undo_list.push((A.version, old_value));
-newA.data.values[pos] = val;
-return newA;
-\end{lstlisting}
-
-Next we explain the implementation of $A.get(pos)$.
-
-\begin{lstlisting}[language=Java]
-guess_val = A.data.values[pos];
-if (A.version == A.data.version) {
-  return guess_val;
-}
-
-undo_list = this.data.undo_lists[pos];
-if (undo_list.size() == 0) {
-  return guess_val;
-}
-
-upper = undo_list.size() - 1;
-if (undo_list[upper].version < A.version) {
-  return guess_val;
-}
-\end{lstlisting}
-
-The last step is to binary search $undo\_list$, between the indices 0 and $upper$ (inclusive of 0 and upper).
-We find the log entry $L$ with the smallest version $X$ such that $A.version \leq X$.
-We return $L.value$.
-
-The implementation of tabulate is straightforward - just allocate a new array and fill the array with the required values.
-
-\subsection{Proof of Correctness}
-
-\begin{theorem}
-Functional Arrays are linearizable.
-\end{theorem}
-
-\begin{proof}
-
-The linearization point of get is when we compare if A.version and AD.version are the same.
-The linearization point of set is the LLSC.
-Any instruction in tabulate can be its linearization point - we omit discussion of tabulate because the details are uninteresting.
-
-First, we prove the linearization point of get. (fill this out)
-
-Next, we prove the linearization point of set. (fill this out)
-
-\end{proof}
-
-Linearizability allows us to assume that array operations happen atomically.
-Note that all non array operations in our source language are executed in a single time step, so any well formed program is linearizable.
-In fact, since linearizability composes, we can show that more complicated languages that use our arrays are also linearizable.
-
-In a previous section, we gave the evaluational (big step) dynamics of functional arrays.
-Alternatively, we could give the non-deterministic structural (little step) dynamics which describes the steps in the execution of a program in the source language.
-At each fork-join, if both sides of the fork can take a step, we could take a step on either side.
-
-\begin{theorem}
-The non-deterministic structural dynamics and the evaluational dynamics produce the same result in the source language.
-\end{theorem}
-
-\begin{proof}
-This is a standard implicit parallelism type theorem, for example see Chapter 39 in~\cite{harper}, so we omit the details. 
-\end{proof}
-
-Now, to show that our implementation correctly implements the source language, it suffices to show that any valid execution of our implementation produces the same result as a little step transition sequence in the source language (since all transition sequences produce the same value as the evaluational dynamics).
-
-\begin{theorem}
-For any program $P$ in the source language, executing our implementation $I$ of $P$ produces the same result as some transition sequence of $P$. Note that $I$ is a program in the target language.
-\end{theorem}
-
-\begin{proof}
-We prove the claim by induction. We focus on the part of the proof dealing with arrays and give a sketch of the rest.
-
-Our inductive hypothesis is that after $t$ time steps in an arbitrary execution of $I$, there exists a transition sequence $S$ s.t.:
-\begin{enumerate}
-\item $I$ becomes $I'$ after this particular execution, $S$ takes $P$ to $P'$
-\item $I'$ and $P'$ are equivalent in the following sense.
-Any non array value is identical in $I'$ and $P'$.
-For any abstract array $A$ in $P'$ there exists a corresponding representation $A_I$ s.t.
-the following holds.
-Consider arbitrary index $i$ and suppose that $A[i]$ has value $val$.
-Suppose that $A_I$ has version $V$ and is pointing to an ArrayData object $AD$.
-If there exists some version in $AD.logs[i]$ which is greater than $V$ then there exists a log entry $(v', val)$ where $v'$ is the smallest version in $AD.logs[i]$ that is $\geq V$.
-If no version in $AD.logs[i]$ is greater than $V$, then $AD.values[i]$ is val.
-\end{enumerate}
-
-The base case holds since $P$ and $I$ are initially identical.
-For the inductive step, when $I'$ takes a step, we take the corresponding step in $P$ to get a new transition sequence $S'$.
-We know that such a step exists because $I'$ and $P'$ are equivalent.
-If the step is not an array function then the IH holds since the step will be executed in the same way in the source and target languages.
-
-Suppose that the next step is tabulate.
-By the IH, tabulate is called with the same arguments $n$ and $f$ in the source language and the implementation.
-In the source language, tabulate returns an array $A$ with $A[i] = f(i)$ for all $1 \leq i \leq n$.
-In the implementation, tabulate returns an array that points to a new ArrayData object $AD$ with empty logs and $AD.values[i] = f(i)$ for all $1 \leq i \leq n$.
-So the inductive step holds.
-
-Suppose that the next step is get.
-By the IH, get is called on equivalent arrays and the same index $i$ in the source language and implementation.
-It is easy to trace through the implementation of get and see that get returns the same value in the source language and implementation.
-
-Suppose that the next step is set.
-Suppose that in the target language we execute $set(A_I, i, v)$.
-Then by IH, in the source language the corresponding step is $set(A, i, v)$ where $A$ and $A_I$ are equivalent.
-Denote $set(A, i, v)$ by $B$ and $set(A_I, i, v)$ by $B_I$.
-We need to show that $B$ and $B_I$ are equivalent, and all arrays that were previously equivalent are still equivalent.
-
-Case 1: we create a new ArrayData object $AD$ in set.
-In this case, for all $j \neq i$ we set $AD.values[j]$ to $get(A_I, j)$ which by the IH is the same as $A[j]$.
-Since we only modified index $i$, $j \neq i \Rightarrow A[j] = B[j]$.
-Also, we set $AD.values[i]$ to $v$, so $AD.values[i] = B[i]$.
-Since all logs are empty, this means that $B$ and $B_I$ are equivalent.
-Since we create a new ArrayData object, we do not modify any other arrays, so arrays that were previously equivalent are still equivalent.
-
-Case 2: we modify the ArrayData object $AD$ that $A_I$ points to.
-This means that $A.version = AD.version$.
-We can show that all log entries in $AD$ have version $< AD.version$, and are in increasing order.
-Then the IH implies that before modifying $AD$, $AD.values[j] = A[j]$ for all indices $j$.
-So after setting $AD.values[i]$ to $v$, $AD.values[j] = B[j]$ for all $j$, which means $B_I$ and $B$ are equivalent.
-
-Consider arbitrary abstract array $C$ in the source language and corresponding representation $C_I$ in the implementation with $C_I \neq B_I$.
-If $C_I$ did not point to $AD$ then $C_I$ remains equivalent to $C$.
-Regardless, we note that for all indices $j \neq i$, $C_I$ remains the same and is equivalent to $C$.
-At index $i$, if the value at $C[i]$ was previously stored in a log entry, then $C_I$ and $C$ remain equivalent since log entries are added in increasing version numbers.
-If the value at $C[i]$ was stored in $AD[i]$ then we add a log entry at index $i$ so $C_I$ and $C$ remain equivalent.
-
-\end{proof}
-
-\section{Machine Model}
-
-Our target language is a Java-like language augmented with tabulate, fork-join, lambda functions, and link load store conditional (LLSC).
-Fork-join spawns two threads, one thread executing each sub-expression.
-Link load store conditional (LLSC) takes an (address, old value, new value) tuple as argument.
-LLSC first checks that the value at address is the same as old value, if it is different then LLSC returns false.
-Otherwise LLSC atomically stores new value at address and returns true, however if the value at address was modified between the load and store, LLSC returns false.
-
-Our target language is run on a P processor machine.
-In a single time step the machine takes at most P runnable instructions and executes them.
-The effect of executing the ($\leq P$) instructions in a time step is the same as some (arbitrary) sequential ordering of the instructions.
-We assume that the machine does not reorder instructions in the target language.
-In reality, languages like Java have very relaxed consistency models, and we deal with this in our real implementation using memory fences.
-
-The link load store conditional architecture deserves special mention.
-Call an LLSC operation pending if the load operation has completed but the store operation has not.
-Multiple threads can load the value at a particular address and compare it with old value in a single time step.
-Only a single store operation to a particular address can be executed at a time step.
-However, after the store operation, in each time step the bus arbiter notifies all pending LLSC operations to the same memory address that they have failed.
-This assumption is reasonable because all LLSC instructions are typically processed by the bus arbiter, so the bus arbiter can keep track of pending LLSC instructions to each memory address.
+\input{conclusion}
 

### 2016-09-25 Paper/structural.tex
Reorganize repository

\section{Structural Dynamics}
\label{sec:structural}

Our goal is to bound the runtime cost of evaluating a program (in our language) on a $p$ processor machine. For this purpose, we define two separate structural (small step) cost dynamics for our language: the \emph{idealized parallel dynamics} and the \emph{interleaved dynamics}.

The parallel dynamics defines the parallel \emph{span} (or depth) of a
computation, which is intuitively the number of parallel steps
given an unbounded number of processors. It is a pure dynamics
and allows all parallel expressions to proceed on the same step.

The interleaved dynamics defines the total \emph{work} performed by a
computation, which can be thought of as the number of instructions
used by the computation (within constant factors). The dynamics is
not pure, requiring a store to keep track whether each sequence is a leaf or interior, so that different costs can be charged for \get{} and \set{} in the two cases.  To account for parallelism, the dynamics allows for non-deterministic interleaving of steps on parallel expressions.  In conjunction with the impure cost for \get{} and \set{}, this means the overall work is non-deterministic. In Section~\ref{sec:evaluational} we give a deterministic cost dynamics and
show that it provides an upper bound on the work over all possible
interleavings.

The two dynamics are useful together because we can bound the
runtime on any fixed number of processors based on the work and span.

\subsection{Idealized Parallel Dynamics}

The idealized parallel dynamics captures the evaluation steps taken on a
machine with an unbounded number of processors and
is given by the following judgement, where $e$ is the expression and
$e'$ is the resulting expression:
$$e \to_{par} e'$$

Let $\overline{\text{\get{}}}(A, i)$ denote $A[i]$, $\overline{\text{\set{}}}(A, (i,v))$ denote $A[i \mapsto v]$ (which denotes the sequence obtained if the value at index $i$ of $A$ is replaced with $v$), and $\overline{\text{\new{}}}((n,v))$ denote a sequence of size $n$ with the value $v$ at each index. Intuitively, $\overline{\text{\get{}}}$, $\overline{\text{\set{}}}$, and $\overline{\text{\new{}}}$ represent abstract versions of \get{}, \set{}, and \new{}.

We show the rules for get and fork-join in Figure \ref{fig:parallel-dynamics}. As usual, the judgement \func{val} describes terminal values. Notice that in fork-join, both sides of the fork-join take a step at the same time if possible. The rules for \set{} and \new{} are similar to the rules for \get{}, and the rules for function application and if-then are as in a standard applicative order functional programming language.

\begin{figure}
$$\frac{A, a \; \text{val}}{\tget{}(A, a) \to_{par} \overline{\mbox{\get{}}}(A, a)} \text{ (get-eval)}$$
$$\frac{e_1 \to_{par} e_1' \;\;\; e_2 \to_{par} e_2'}{e_1 || e_2 \to_{par} e_1' || e_2'} \text{ (fork)}$$
$$\frac{v_1 \; \text{val} \;\;\; e_2 \to_{par} e_2'}{v_1 || e_2 \to_{par} v_1 || e_2'} \text{ (step-right)}$$
$$\frac{v_2 \; \text{val} \;\;\; e_1 \to_{par} e_1'}{e_1 || v_2 \to_{par} e_1' || v_2} \text{ (step-left)}$$
$$\frac{v_1, v_2 \text{ val}}{(v_1 || v_2) \to_{par} (v_1, v_2)} \text{ (join)}$$

\caption{Example rules in the parallel dynamics for our language.}
\label{fig:parallel-dynamics}
\end{figure}

The parallel structural dynamics, unlike the interleaved
structural dynamics, is deterministic.

\begin{definition}
A \emph{parallel transition sequence} $T$ is a sequence $(e_0, ..., e_n)$ with $e_i \to_{par} e_{i+1}$ for all $0 \leq i < n$. We say that $e_0 \to^n_{par} e_n$ or $e_0 \to^*_{par} e_n$.
\end{definition}

\begin{definition}
The \emph{span} of a parallel transition sequence $T$, denoted by $SP(T)$, is its length $n$.
\end{definition}

\subsection{Interleaved Cost Dynamics}

The interleaved dynamics is a concurrent dynamics and is given by the following judgement, where $\sigma$ is the
store, $e$ is the expression, $\sigma'$ is the resulting store, $e'$ is the
resulting expression after a step has been taken, and $w$ is the work.
$$\sigma, e \to \sigma', e', w$$

Sequence values are represented by the pair $(l, V)$ where $l$ is a
label into the store $\sigma$ and $V$ is a list of the elements in the sequence.  The store is a mapping from labels to either $+$ (indicating a leaf sequence)
or $-$ (indicating an interior sequence). Let $L(\sigma)$ denote the set of labels in the store $\sigma$.

Let $g_l(V)$ and $g_i(V)$ be the work of \get{} applied to $V$ if it
is a leaf or interior sequence, respectively, and $s_l(V)$ and $s_i(V)$
be the work of \set{} applied to $V$ if it is a leaf or interior
sequence.  Let $n(a)$ be the work of evaluating \new{}$(a)$.  We assume
that $g_l(V) \leq g_i(V)$ and $s_l(V) \leq s_i(V)$ (operating on leaf
sequences is cheaper that operating on interior sequences).

\begin{figure}
$$\frac{v_1, v_2 \; \text{val}}{\sigma, (\lambda (x,y).e)(v_1,v_2) \to \sigma, [v_1/x][v_2/y]e, 1} \text{ (func-app)}$$

$$\frac{}{\sigma, \text{if } \text{true} \; e_2 \; e_3 \to \sigma, e_2, 1} \text{ (if-true)}$$

$$\frac{}{\sigma, \text{if } \text{false} \; e_2 \; e_3 \to \sigma, e_3, 1}  \text{ (if-false)}$$

$$\frac{a \; \text{val} \;\;\; l \not\in L(\sigma)}{\sigma, \tnew{}(a) \to \sigma[l \mapsto +], \overline{\mbox{\new{}}}(a), n(a)} \text{ (new)}$$

$$\frac{A = (l, V) \;\;\; \sigma[l] = + \;\;\; a \text{ val}}{\sigma, \tget{}(A, a) \to \sigma, \overline{\mbox{\get{}}}(V,a), g_l(V)} \text{ (get-leaf)}$$

$$\frac{A = (l, V) \;\;\; \sigma[l] = -  \;\;\; a \text{ val}}{\sigma, \tget(A, a) \to \sigma, \overline{\mbox{\get{}}}(V,a), g_i(V)} \text{ (get-interior)}$$

$$\frac{A = (l, V) \;\;\; \sigma[l] = + \;\;\;  l' \not\in L(\sigma) \;\;\; a \text{ val}}{\begin{gathered}\sigma, \tset(A, a) \to \sigma[l \mapsto -, l' \mapsto +], \\ (l', \overline{\mbox{\set{}}}(V, a)), s_l(V)\end{gathered}} \text{ (set-leaf)}$$

$$\frac{A = (l, V) \;\;\; \sigma[l] = - \;\;\;  l' \not\in L(\sigma) \;\;\; a \text{ val}}{\begin{gathered}\sigma, \tset(A, a) \to \sigma[l' \mapsto +], \\ (l', \overline{\mbox{\set{}}}(V,a)), s_i(V)\end{gathered}} \text{ (set-interior)}$$

$$\frac{\sigma, e_1 \to \sigma', e_1', w}{\sigma, e_1 || e_2 \to \sigma', e_1' || e_2, w} \text{ (step-left)}$$
$$\frac{\sigma, e_2 \to \sigma', e_2', w}{\sigma, e_1 || e_2 \to \sigma', e_1 || e_2', w} \text{ (step-right)}$$
$$\frac{v_1, v_2 \text{ val}}{\sigma, (v_1 || v_2) \to \sigma, (v_1, v_2), 1} \text{ (join)}$$

\caption{The interleaved dynamics for our language.  We omit
  rules that involve stepping the arguments to \get{}, \set{}, \new{},
  and other language constructs.}
\label{fig:interleaved}
\end{figure}

The dynamics is given in figure~\ref{fig:interleaved}.  Note that
\set$(A,a)$ where $A = (l,V)$ and $\sigma[l] = +$ creates a new label and
value, extends the store to indicate the new value is a leaf, and
updates the store at $l$ to indicate that $A$ is now interior.  This
is the impure aspect of the cost dynamics since there can be other
references to $l$.  Also note that the work for for \get{} and \set{}
depends on whether the sequence is a leaf or interior.

The rules for fork-join are non-deterministic---either side of the
fork can take a step.  This allows for arbitrary interleaving of
instructions on different sides of a fork-join.  After both sides of a
fork-join are fully evaluated the parallel pair is converted to a regular pair.

\begin{definition}
A \emph{transition sequence} $T$ is a sequence of states $[(\sigma_0, e_0),
..., (\sigma_n, e_n)]$ and per-step work $[w_1, ..., w_n]$ s.t. for all $0 \leq i < n$, $\sigma_i, e_i \to \sigma_{i+1}, e_{i+1}, w_{i+1}$ . We say that $T$ takes $\sigma_0, e_0$ to $\sigma_n, e_n$ and has length $n$ and denote this by $\sigma_0, e_0 \to^n \sigma_n, e_n$.
\end{definition}

\begin{definition}
We say that $T$ is \emph{maximal} if there does not exist $\sigma', e', w'$ s.t. $\sigma_n, e_n \to \sigma', e', w'$.
\end{definition}

\begin{definition}
A \emph{transition subsequence} $T_{i, j}$ of $T$ is given by the
sequence of states $[(\sigma_i, e_i), ..., (\sigma_j, e_j)]$ and per-step work
$[w_{i+1}, ..., w_j]$. Note that $T = T_{0,n}$.   
\end{definition}

\begin{definition}
The \emph{work of a transition sequence} $T$ is given by:
$$W(T) = \sum_{i=1}^n w_i$$
\end{definition}

Different transition sequences starting and ending at the same states
may have different work.  In particular, a sequence can be used
concurrently by multiple parallel expressions and the work could
depend on the order in which the instructions are interleaved.
Figure~\ref{fig:fork_join_intuition} shows 3 ways a leaf sequence $A$
might be used.  In particular, one side of the fork might call \get{}
on $A$ while the other side calls \set{} (see the center panel). Calls
to \get{} on the left branch would have work $g_l(A)$ if and only if
they execute before the call to \set{} on the right branch, otherwise
they would have work $g_i(A)$.

\begin{figure}[!ht]
\centering
\includegraphics[scale=0.3]{fork_join_intuition}
\nocaptionrule \caption{Different ways a sequence might be used in a fork join}
\label{fig:fork_join_intuition}
\end{figure}

We cannot make any assumptions about how the instructions in different
branches of a fork are interleaved, so we need to capture the worst
case work over all possible interleavings.

\begin{definition}
Let $w$ be the max of $W(T)$ over all transition sequences starting at $\sigma_0, e_0$. If $W(T)$ is unbounded then $w = \infty$. $w$ is the \emph{work of evaluating} $\sigma_0, e_0$. 
\end{definition}

\subsection{Machine Costs}

We now consider the cost of evaluating an expression on a shared
memory machine with $P$ processors.  We assume that
within a constant factor, the cost functions $g_l(V)$,
$g_i(V)$, $s_l(V)$, $s_i(V)$, and $n(a)$ are upper bounds on the
number of instructions needed by the implementation for each of the
corresponding functions.  In this case we say the cost functions have
valid work bounds. In Section~\ref{sec:implementations} we describe an implementation and prove work bounds for sequences.

\new{}, \get{} and \set{} can each have parallelism in their implementation, so in addition to the work we need to know the span for these functions (the number of time steps on an unbounded number of processors). Here we assume the span for \new{}, \get{}, and \set{} is bounded by some $f(P)$ where $P$ is the number of processors used.

Using the greedy scheduling theorem ~\cite{BlumofeLeiserson} or Brent's
theorem~\cite{brent} knowing the total work and total span (number of
steps on an unbounded number of processors) is sufficient to bound the
time on any finite number of processors. We assume that standard instructions such as reading and writing to memory, arithmetic operations, etc. take constant time (and work). This leads to the following theorem.

\begin{theorem}
\label{thm:brent}
Given cost functions with valid work bounds each with maximum span
$f(p)$, then for any expression $e$ with a constant number of
variable names if $w$ is the work of evaluating $\sigma, e$, and $S$ is the span for the (unique) maximal transition sequence starting at $e$, then $e$ can be evaluated using any greedy schedule on a $P$ processor machine in time
\[ T \leq c \left(\frac{w}{P} + S f(P)\right)\]
for some constant $c$ that is independent of the expression $e$.
\end{theorem}
\begin{proof} (outline).  This follows from the greedy scheduling theorem and previous work on bounded cost implementations. Bounding the number of variables is needed to bound the cost of evaluating each step of the program itself ~\cite{BG95}. It avoids issues of non-constant cost for looking up the values of variables in environments.
\end{proof}

### 2016-09-25 java_implementation/ArrayInterface.java

@@ -3,8 +3,6 @@
  * Set returns the array because we are simulating 
  * functional programming semantics.
  *
- * @author Ananya Kumar
- *
  */
 
 public interface ArrayInterface<T> {

### 2016-09-25 java_implementation/ArrayUnitTest.java

@@ -1,8 +1,6 @@
 /*
  * Simple unit test for PersistentArray.
  *
- * @author Ananya Kumar
- *
  */
 
 import java.util.*;

### 2016-09-25 java_implementation/PersistentArray.java

@@ -16,8 +16,6 @@
  * memory barriers (on top of performing their
  * intended functionality).
  *
- * @author Ananya Kumar
- *
  */
 
 import java.util.*;

### 2016-09-25 java_implementation/RWArray.java

@@ -10,8 +10,6 @@
  * RWArray functions are guaranteed to act as a memory
  * barrier/fence.
  *
- * @author Ananya Kumar
- *
  */
 
 public class RWArray<T> {

### 2016-09-25 java_implementation/STArray.java

@@ -2,8 +2,6 @@
  * Wrapper for regular Java Arrays to
  * facilitate comparisons with Persistent Arrays.
  *
- * @author Ananya Kumar
- *
  */
 
 public class STArray<T> implements ArrayInterface<T> {
